---
title: "02_Differential_Expresssion_Analysis - Zilucoplan"
#output: html_notebook
---
# Load libraries

```{r message=FALSE}
# Packages to load
packages <- c(
  "clusterProfiler","DESeq2", "dplyr", "edgeR", "ensembldb", "genefilter",  "limma", "RColorBrewer", "rhdf5", "rstatix", "rtracklayer", 
  "tidyverse", "tximport", "vioplot")

############################################################################################################
load_libraries <- function(packages) {
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

# Load all packages
load_libraries(packages)

```

# Prepare data

Set naming variables
```{r}
study_A <-"Standard_of_Care"
study_B <-"Zilucoplan"
root.name <-paste(study_B, "_Day5_vs_Day1", sep="")
```


Load data (counts and metadata) - stored in Rdata object
for this nested analysis, want both the treatment arm and the standard of care
```{r}
load("./data_files/zilucoplan_data.RData")
load("./data_files/soc_data.Rdata")

metadata<-rbind(zil_meta, soc_meta)
counts<-cbind(zil_counts, soc_counts)

raw_counts<-as.data.frame(counts)
```

Annotation file - Homo_sapiens.GRCh38.109
```{r}
#read in GTF file
gtf_file <- ('./data_files/Homo_sapiens.GRCh38.109.gtf.gz')
gtf_content <- import(gtf_file, feature.type = 'gene')
annotation <- data.frame(elementMetadata(gtf_content), stringsAsFactors = FALSE)
row.names(annotation) <- annotation$gene_id
# tx2gene
annotation_transcript <- elementMetadata(import(gtf_file, feature.type = 'transcript'))
tx2gene <- annotation_transcript[,c("gene_id","gene_name")]
tx2gene <- as.data.frame(tx2gene)
tx2gene <- tx2gene[!duplicated(tx2gene$gene_id), ]

```

# Create a Design
```{r}
stringsAsFactors = FALSE
group <- factor(metadata$TreatmentTime)
TimePoint <- factor(metadata$Day)
Donor <- factor(metadata$Unique_ID)
Treatment <- factor(metadata$Treatment_Arm)
col.cell <- brewer.pal(9,"Set1")[group]
head(col.cell)

TimePoint2_TreatmentA <- as.numeric(Treatment=="Standard of Care" & TimePoint=='D5')
TimePoint2_TreatmentB <- as.numeric(Treatment=="Zilucoplan" & TimePoint=='D5')
design <- model.matrix(~ Donor+TimePoint2_TreatmentA+TimePoint2_TreatmentB)

```


```{r fig.height=5, fig.width = 15}
dge <- DGEList(counts= raw_counts,genes=row.names(raw_counts))
dge_norm <- calcNormFactors(dge,lib.size=T,method="TMM")
cpm.count <- cpm(dge_norm)
keep <- filterByExpr(dge_norm,group=group)
dge_norm_filt<-dge_norm[keep,,keep.lib.sizes=FALSE]

v <- voomWithQualityWeights(dge_norm_filt, design=design, plot=TRUE)
voom_dataframe <- as.data.frame(v)

```


# Exploratory analysis and visualisation of data

## 1. Boxplots

```{r fig.width=25, fig.height=5}
logcounts1 <- cpm(dge_norm_filt,log=TRUE)
boxplot(logcounts1, xlab="", ylab="Log2 counts per million",las=2,col=col.cell)
abline(h=median(logcounts1),col="blue")
title("Boxplots of logCPMs (normalized)")
```


## 2. IQR v median plot
```{r fig.width=10, fig.height=10}
#  Make IQR_Median Plots
cpm<-cpm(dge_norm_filt)
IQR<-apply(cpm, 2, IQR)
Median<-apply(cpm, 2, median)
diff1<-mean(Median)-min(Median)
diff2<-max(Median)-mean(Median)
diff3<-mean(IQR)-min(IQR)
diff4<-max(IQR)-mean(IQR)

#  These plot settings work for this data and may need to be adjusted for your own data.
Xlim=c(mean(Median)-2*diff1,mean(Median)+2*diff2)
Ylim=c(mean(IQR)-2*diff3,mean(IQR)+2*diff4)

#plot as PDF - # if showing inline
#file.name = paste(root.name, "IQR_vs_Median_Plot.pdf", sep="")
#pdf(file.name)
plot(Median, IQR, main="IQR vs. Median TMM normaliztion", type="n", xlim=Xlim,ylim=Ylim)
text(Median, IQR, labels=names(IQR))

#  Make boxes for StDev.
Median_mean<-mean(Median)
c_sd1_mean<-sd(Median)
c_sd2_mean<-2*sd(Median)
c_sd3_mean<-3*sd(Median)
IQR_mean<-mean(IQR)
c_sd1_IQR<-sd(IQR)
c_sd2_IQR<-2*sd(IQR)
c_sd3_IQR<-3*sd(IQR)

x0_c<-Median_mean-c_sd1_mean
y0_c<-IQR_mean-c_sd1_IQR
x1_c<-Median_mean+c_sd1_mean
y1_c<-IQR_mean+c_sd1_IQR

x0_c.2<-Median_mean-c_sd2_mean
y0_c.2<-IQR_mean-c_sd2_IQR
x1_c.2<-Median_mean+c_sd2_mean
y1_c.2<-IQR_mean+c_sd2_IQR

x0_c.3<-Median_mean-c_sd3_mean
y0_c.3<-IQR_mean-c_sd3_IQR
x1_c.3<-Median_mean+c_sd3_mean
y1_c.3<-IQR_mean+c_sd3_IQR

segments(x0_c,y0_c, x1=x1_c, y1=y0_c, col="blue")
segments(x0_c,y0_c, x1=x0_c, y1=y1_c, col="blue")
segments(x1_c,y0_c, x1=x1_c, y1=y1_c, col="blue")
segments(x0_c,y1_c, x1=x1_c, y1=y1_c, col="blue")

segments(x0_c.2,y0_c.2, x1=x1_c.2, y1=y0_c.2, col="red")
segments(x0_c.2,y0_c.2, x1=x0_c.2, y1=y1_c.2, col="red")
segments(x1_c.2,y0_c.2, x1=x1_c.2, y1=y1_c.2, col="red")
segments(x0_c.2,y1_c.2, x1=x1_c.2, y1=y1_c.2, col="red")

#  This portion is out of range so I removed it.
segments(x0_c.3,y0_c.3, x1=x1_c.3, y1=y0_c.3, col="green")
segments(x0_c.3,y0_c.3, x1=x0_c.3, y1=y1_c.3, col="green")
segments(x1_c.3,y0_c.3, x1=x1_c.3, y1=y1_c.3, col="green")
segments(x0_c.3,y1_c.3, x1=x1_c.3, y1=y1_c.3, col="green")

outliers2 <- which(IQR < IQR_mean - 3*sd(IQR) | IQR > IQR_mean + 3*sd(IQR))
points(Median[outliers2], IQR[outliers2], col="red", pch=19)

```


## PCA plot

```{r fig.width=10, fig.height=10}
cpm<-cpm(v)
Y <- apply(cpm, 1, function(y) scale(y, center=TRUE, scale=FALSE))
s <- svd(Y)
#  Get Variance explained by the components
Var1<-s$d^2/sum(s$d^2)
Var2<-cumsum(s$d^2/sum(s$d^2))
Var<-cbind(Var1,Var2)
Var<-round(Var,3)
colnames(Var)<-c("Percent Variance","Cumulative % Variance")


plot(s$u[,1], s$u[,2], pch=19, cex=2, xlab=paste0("PC1, VarExp:", round(Var[1],3)), ylab=paste0("PC2, VarExp:", round(Var[2],3)),main="PCA plot TMM Normalization",col=col.cell, ylim=c((min(s$u[,2])*1.2),(max(s$u[,2])*1.2)),xlim=c((min(s$u[,1])*1.2),(max(s$u[,1])*1.2)))
text(s$u[,1], s$u[,2], labels=colnames(cpm),adj=c(0,2),cex=0.5)
legend("topleft",legend = unique(metadata$TreatmentTime),col = unique(col.cell), pch=19,cex=1)


```

# Differential Expression

```{r}
dge_norm<- estimateDisp(dge_norm, design,robust=TRUE)
dge.fit <- glmFit(dge_norm, design) # fitting the normalised DGE data, after calculating dispersion values
# Define contrasts for the time effect in each treatment
contrasts <- makeContrasts(
  TimeEffectInTreatmentA = TimePoint2_TreatmentB,
  levels = design
)
# Fit the contrasts
lrt <- glmLRT(dge.fit, contrast=contrasts)

# Print the top differentially expressed genes
topTags(lrt)
# show how many 
print(summary(decideTests(lrt)))
```



```{r}
# Merge the gene names to the lrt results
lrt$table <- merge(lrt$table, tx2gene, by.x= "row.names", by.y= "gene_id")
lrt$table <-as.data.frame(lrt$table)
# Rename the column names for clear understanding
colnames(lrt$table)[which(names(lrt$table) == "Row.names")] <- "ensembl_gene_id"
colnames(lrt$table)[which(names(lrt$table) == "external_gene_name")] <- "gene_symbol"
```

## Filtering

```{r}
reads.cpm <- cpm(dge_norm)
unfiltered.results <- data.frame(id=dge_norm$genes, reads.cpm, lrt$table)
```

```{r}
filter <- apply(X=reads.cpm, MARGIN=1,
                FUN=function(data) {data[order(rank(data), decreasing=TRUE)[2]]})
```

```{r}
# Calculate the mean of the variable 'filter' when its value is 0
lowerQuantile <- mean(filter == 0)

# Check if the lowerQuantile is less than .95. If it is, set upperQuantile to .95, otherwise set it to 1
if (lowerQuantile < .95) upperQuantile <- .95 else upperQuantile <- 1

# Create a sequence of 50 evenly spaced numbers between lowerQuantile and upperQuantile
theta <- seq(lowerQuantile, upperQuantile, length=50)

# Apply the BH method for p-value adjustment to the unfiltered results
# The function 'filtered_p' presumably adjusts the p-values based on some sort of filtering criterion
filtPadj <- filtered_p(filter=filter, test=unfiltered.results$PValue,
                       theta=theta, method="BH")

# Set a threshold for the false discovery rate
min.fdr <- 0.05

# Count the number of adjusted p-values that are less than the false discovery rate threshold
numRej <- colSums(filtPadj < min.fdr, na.rm = TRUE)

# Calculate the quantiles of the 'filter' variable based on the theta sequence
filter.quantiles <- quantile(filter, probs=theta)

# Fit a locally weighted scatterplot smoothing (LOWESS) to the number of rejected hypotheses as a function of theta
lo.fit.theta <- lowess(numRej ~ theta, f=1/5)

# Check if the maximum number of rejected hypotheses is less than or equal to 10
if (max(numRej) <= 10) {
  j <- 1
} else {
  # Calculate the residuals between the observed number of rejections and the LOWESS fit
  residual <- if (all(numRej==0)) {
    0
  } else {
    numRej[numRej > 0] - lo.fit.theta$y[numRej > 0]
  }

  # Calculate a threshold based on the maximum value of the LOWESS fit minus the standard deviation of the residuals
  thresh <- max(lo.fit.theta$y) - sqrt(mean(residual^2))

  # Set 'j' to the first index where the number of rejections exceeds the threshold, or to 1 if no such index exists
  j <- if (any(numRej > thresh)) {
    which(numRej > thresh)[1]
  } else {
    1
  }
}

plot(theta, numRej, type="b", xlab="", ylab="", lwd=3,
     frame.plot=FALSE, col="black")
```

Extract results
```{r}
results <- unfiltered.results
results$FDR <- filtPadj[, j, drop=TRUE]

results$de <- sign(results$logFC)*(results$FDR < 0.05)
results$sig <- abs(results$de)

results[is.na(results$sig),]$sig <- 0

#how many?
total_genes <- results[results$sig == 1, ]
upregulated_genes <- results[results$sig == 1 & results$de == 1, ]
downregulated_genes <- results[results$sig == 1 & results$de == -1, ]
print(paste("Total DEGs =", nrow(total_genes)))
print(paste("Upregulated DEGs =", nrow(upregulated_genes)))
print(paste("Downregulated DEGs =", nrow(downregulated_genes)))
#save
file.name = paste(root.name, "results.txt", sep="")
print(file.name)
write.table(results, file=file.name, sep="\t", row.names=FALSE, quote=FALSE)

file.name = paste(root.name, "sig_results.txt", sep="")
print(file.name)
write.table(results[results$sig==1,], file=file.name, sep="\t", row.names=FALSE, quote=FALSE)
```

# Enrichment analysis

```{r}
#this code loops through each of the GO categories at once and generates individual plots in one go
#checks if there is enrichment and if there isn't it skips and moves on to the next
#also prints out a warning message saying no enrichment found, skipping.... so you can see what hasn't got any enrichment
#output files will be used for combined visulisation in subsequent scripts for the figures in the paper
upregulated_genes <- results$gene_name[results$logFC > 0 & results$sig == 1]
downregulated_genes <- results$gene_name[results$logFC < 0 & results$sig == 1]

# Define the background gene set
background_genes <- results$gene_name

# List of GO categories
go_categories <- c("MF", "BP", "CC")

for (go in go_categories) {

  # Perform GO enrichment analysis for upregulated genes
  ego_up <- tryCatch({
    enrichGO(gene = upregulated_genes,
             universe = background_genes,
             OrgDb = org.Hs.eg.db,
             keyType = "SYMBOL",
             ont = go,
             pAdjustMethod = "BH",
             qvalueCutoff = 0.05,
             readable = TRUE)
  }, error = function(e) NULL)

   # Perform GO enrichment analysis for downregulated genes
  ego_down <- tryCatch({
    enrichGO(gene = downregulated_genes,
             universe = background_genes,
             OrgDb = org.Hs.eg.db,
             keyType = "SYMBOL",
             ont = go,
             pAdjustMethod = "BH",
             qvalueCutoff = 0.05,
             readable = TRUE)
  }, error = function(e) NULL)


  if (!is.null(ego_up$geneID) && nrow(ego_up) > 0) {
    # Barplot for top 20 enriched pathways for upregulated genes
    up_bar <- barplot(ego_up, showCategory = 20) +
      ggtitle(paste("Top 20 enriched", go, "pathways for upregulated genes")) +
      theme_minimal()

    # Save plot as PDF
    ggsave(filename = paste(root.name, "upregulated_", go, "_barplot.pdf", sep = ""), plot = up_bar)

    # Dotplot for top 20 enriched pathways for upregulated genes
    up_dot <- dotplot(ego_up, showCategory = 20) +
      ggtitle(paste("Top 20 enriched", go, "pathways for upregulated genes")) +
      theme_minimal()

    # Save plot as PDF
    ggsave(filename = paste(root.name, "upregulated_", go, "_dotplot.pdf", sep = ""), plot = up_dot)

    # Save enriched terms to CSV
    write.csv(ego_up, file = paste(root.name, "upregulated_", go, "_enriched_terms.csv", sep = ""))
  } else {
    cat(paste("No enriched terms found for", go, "in the upregulated gene set. Skipping..."))
  }

  if (!is.null(ego_down$geneID) && nrow(ego_down) > 0) {
    # Barplot for top 20 enriched pathways for downregulated genes
    down_bar <- barplot(ego_down, showCategory = 20) +
      ggtitle(paste("Top 20 enriched", go, "pathways for downregulated genes")) +
      theme_minimal()

    # Save plot as PDF
    ggsave(filename = paste(root.name, "downregulated_", go, "_barplot.pdf", sep = ""), plot = down_bar)

    # Dotplot for top 20 enriched pathways for downregulated genes
    down_dot <- dotplot(ego_down, showCategory = 20) +
      ggtitle(paste("Top 20 enriched", go, "pathways for downregulated genes")) +
      theme_minimal()

    # Save plot as PDF
    ggsave(filename = paste(root.name, "downregulated_", go, "_dotplot.pdf", sep = ""), plot = down_dot)

    # Save enriched terms to CSV
    write.csv(ego_down, file = paste(root.name, "downregulated_", go, "_enriched_terms.csv", sep = ""))
  } else {
    cat(paste("No enriched terms found for", go, "in the downregulated gene set. Skipping..."))
  }
}
```

