---
  title: "03_differential_expression_visualisation"
#output: html_notebook
---

Visualisation of the differential expression output for all treatment arms
# Contents

1.  Volcano Plots
2.  Top enriched process barplots
3.  GO bubble plots
4.  Immune pathway heatmap
5.  Enriched process overlap between treatments
6.  Unique and shared DEGs - venn and bubbleplot

## Load libraries

```{r message=FALSE}
# Packages to load
packages <- c(
 "ComplexUpset","cowplot", "EnhancedVolcano","forcats","ggnewscale","ggplot2", "ggrepel", "ggvenn", "grid", 
 "gridExtra", "patchwork", "pheatmap", "scales", "stringr", "tidyverse")


############################################################################################################
load_libraries <- function(packages) {
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

# Load all packages
load_libraries(packages)

```

# 1. Volcano Plots

For these plots, need to read in the differential expression files containing the logfc and pval for each treatment arm which were generated and saved within the differential expression notebooks

```{r fig.width=6, fig.height=15}

volcano_data<-read.table("Standard_of_Care_Day5_vs_Day1_results.txt", sep="\t", header = TRUE, row.names=1)

# create custom key-value pairs for fold-change expression
# Define and initialize keyvals as an empty list or vector
keyvals <- vector()
#modify keyvals to show colour the specific genes with your foldchange of interest
# modify keyvals for variables with fold change > 2
keyvals[which(volcano_data$logFC > 0)] <- 'firebrick2'
names(keyvals)[which(volcano_data$logFC > 0)] <- 'Upregulated'
# modify keyvals for variables with fold change < -2
keyvals[which(volcano_data$logFC < 0)] <- 'royalblue'
names(keyvals)[which(volcano_data$logFC < 0)] <- 'Downregulated'
#modify keyvals for non sig genes
keyvals[which(volcano_data$FDR >0.05)] <- 'grey'
names(keyvals)[which(volcano_data$FDR >0.05)] <- 'NS'
#for title of plot
soc_volcano_plot<- EnhancedVolcano(volcano_data,
                               lab=NA,  #rownames(volcano_data),
                               #selectLab = c('FABP4','IL1B', 'HIF1A'),
                               xlim=c(-5,5), #may need to change this to fit all points on
                              ylim=c(0, 10), #may need to change this to fit all points on
                               x='logFC',
                               y='FDR',
                               title = "Standard of Care",
                               titleLabSize =20,
                               subtitle = NULL, #turn off subtitle
                               pCutoff= 0.05, #SET LINES TO SHOW FDR PVAL OF CHOICE
                               FCcutoff=0, #SET LINES TO SHOW FC CUTOFF OF CHOICE
                               colCustom = keyvals,
                               pointSize=3,
                               #legendLabSize = 10,
                               gridlines.major = FALSE,
                               gridlines.minor=FALSE,
                               legendPosition = "none")

# Calculate the number of upregulated and downregulated genes
num_upregulated <- sum(volcano_data$logFC > 0 & !is.na(volcano_data$logFC) & !is.na(volcano_data$FDR) & volcano_data$FDR < 0.05)
num_downregulated <- sum(volcano_data$logFC < 0 & !is.na(volcano_data$logFC) & !is.na(volcano_data$FDR) & volcano_data$FDR < 0.05)
# Add labels above the plot

soc_plot<-soc_volcano_plot + geom_text(aes(x = Inf, y = Inf, label = paste("Upregulated: ", num_upregulated)), size=4, colour = "darkred", hjust = 1.25, vjust = 1, family = "sans") +
  geom_text(aes(x = -Inf, y = Inf, label = paste("Downregulated: ", num_downregulated)), size=4, colour = "navy",hjust = -0.15, vjust = 1, family = "sans")


###### Bemcenitnib #####

## volcano plots ###
#need the LFC  here for the comparison you want to plot
bem_volcano_data<-read.table("Bemcentinib_Day5_vs_Day1_results.txt", sep="\t", header = TRUE, row.names=1)
# create custom key-value pairs for fold-change expression
# Define and initialize keyvals as an empty list or vector
keyvals <- vector()
#modify keyvals to show colour the specific genes with your foldchange of interest
# modify keyvals for variables with fold change > 2
keyvals[which(bem_volcano_data$logFC > 0)] <- 'firebrick2'
names(keyvals)[which(bem_volcano_data$logFC > 0)] <- 'Upregulated'
# modify keyvals for variables with fold change < -2
keyvals[which(bem_volcano_data$logFC < 0)] <- 'royalblue'
names(keyvals)[which(bem_volcano_data$logFC < 0)] <- 'Downregulated'
#modify keyvals for non sig genes
keyvals[which(bem_volcano_data$FDR >0.05)] <- 'grey'
names(keyvals)[which(bem_volcano_data$FDR >0.05)] <- 'NS'
 #for title of plot
bem_volcano_plot<- EnhancedVolcano(bem_volcano_data,
                               lab=NA,  #rownames(volcano_data),
                               #selectLab = c('FABP4','IL1B', 'HIF1A'),
                               xlim=c(-10,10), #may need to change this to fit all points on
                               ylim=c(0, 20), #may need to change this to fit all points on
                               x='logFC',
                               y='FDR',
                               title = "Bemcentinib",
                               titleLabSize =20,
                               subtitle = NULL, #turn off subtitle
                               pCutoff= 0.05, #SET LINES TO SHOW FDR PVAL OF CHOICE
                               FCcutoff=0, #SET LINES TO SHOW FC CUTOFF OF CHOICE
                               colCustom = keyvals,
                               pointSize=3,
                               #legendLabSize = 10,
                               gridlines.major = FALSE,
                               gridlines.minor=FALSE,
                               legendPosition = "none")
# Calculate the number of upregulated and downregulated genes
bem_num_upregulated <- sum(bem_volcano_data$logFC > 0 & !is.na(bem_volcano_data$logFC) & 
                             !is.na(bem_volcano_data$FDR) & bem_volcano_data$FDR < 0.05)
bem_num_downregulated <- sum(bem_volcano_data$logFC < 0 & !is.na(bem_volcano_data$logFC) & 
                               !is.na(bem_volcano_data$FDR) & bem_volcano_data$FDR < 0.05)
final_bem<-bem_volcano_plot + geom_text(aes(x = Inf, y = Inf, label = paste("Upregulated: ", bem_num_upregulated)), size=4, colour = "darkred", hjust = 1.25, vjust = 1, family = "sans") +
  geom_text(aes(x = -Inf, y = Inf, label = paste("Downregulated: ", bem_num_downregulated)), size=4, colour = "navy",hjust = -0.15, vjust = 1, family = "sans")


###### Tozorakimab #####

## volcano plots ###
#need the LFC  here for the comparison you want to plot
toz_volcano_data<-read.table("Tozorakimab_Day5_vs_Day1_results.txt", sep="\t", header = TRUE, row.names=1)
# create custom key-value pairs for fold-change expression
# Define and initialize keyvals as an empty list or vector
keyvals <- vector()
#modify keyvals to show colour the specific genes with your foldchange of interest
# modify keyvals for variables with fold change > 2
keyvals[which(toz_volcano_data$logFC > 0)] <- 'firebrick2'
names(keyvals)[which(toz_volcano_data$logFC > 0)] <- 'Upregulated'
# modify keyvals for variables with fold change < -2
keyvals[which(toz_volcano_data$logFC < 0)] <- 'royalblue'
names(keyvals)[which(toz_volcano_data$logFC < 0)] <- 'Downregulated'
#modify keyvals for non sig genes
keyvals[which(toz_volcano_data$FDR >0.05)] <- 'grey'
names(keyvals)[which(toz_volcano_data$FDR >0.05)] <- 'NS'
#for title of plot
toz_volcano_plot<- EnhancedVolcano(toz_volcano_data,
                               lab=NA,  #rownames(volcano_data),
                               #selectLab = c('FABP4','IL1B', 'HIF1A'),
                               xlim=c(-5,5), #may need to change this to fit all points on
                               ylim=c(0, 30), #may need to change this to fit all points on
                               x='logFC',
                               y='FDR',
                               title = "Tozorakimab",
                               titleLabSize =20,
                               subtitle = NULL, #turn off subtitle
                               pCutoff= 0.05, #SET LINES TO SHOW FDR PVAL OF CHOICE
                               FCcutoff=0, #SET LINES TO SHOW FC CUTOFF OF CHOICE
                               colCustom = keyvals,
                               pointSize=3,
                               #legendLabSize = 10,
                               gridlines.major = FALSE,
                               gridlines.minor=FALSE,
                               legendPosition = "none")

# Calculate the number of upregulated and downregulated genes
toz_num_upregulated <- sum(toz_volcano_data$logFC > 0 & !is.na(toz_volcano_data$logFC) & 
                         !is.na(toz_volcano_data$FDR) & toz_volcano_data$FDR < 0.05)
toz_num_downregulated <- sum(toz_volcano_data$logFC < 0 & !is.na(toz_volcano_data$logFC) & 
                           !is.na(toz_volcano_data$FDR) & toz_volcano_data$FDR < 0.05)

final_toz<-toz_volcano_plot + 
  geom_text(aes(x = Inf, y = Inf, label = paste("Upregulated: ", toz_num_upregulated)), size=4, colour = "darkred", hjust = 1.25, vjust = 1, family = "sans") +
  geom_text(aes(x = -Inf, y = Inf, label = paste("Downregulated: ", toz_num_downregulated)), size=4, colour = "navy", hjust = -0.15, vjust = 1, family = "sans")


#### Zilucoplan ######

## volcano plots ###
#need the LFC  here for the comparison you want to plot
zil_volcano_data<-read.table("Zilucoplan_Day5_vs_Day1_results.txt", sep="\t", header = TRUE, row.names=1)
# create custom key-value pairs for fold-change expression
# Define and initialize keyvals as an empty list or vector
keyvals <- vector()
#modify keyvals to show colour the specific genes with your foldchange of interest
# modify keyvals for variables with fold change > 2
keyvals[which(zil_volcano_data$logFC > 0)] <- 'firebrick2'
names(keyvals)[which(zil_volcano_data$logFC > 0)] <- 'Upregulated'
# modify keyvals for variables with fold change < -2
keyvals[which(zil_volcano_data$logFC < 0)] <- 'royalblue'
names(keyvals)[which(zil_volcano_data$logFC < 0)] <- 'Downregulated'
#modify keyvals for non sig genes
keyvals[which(zil_volcano_data$FDR >0.05)] <- 'grey'
names(keyvals)[which(zil_volcano_data$FDR >0.05)] <- 'NS'
#for title of plot
zil_volcano_plot<- EnhancedVolcano(zil_volcano_data,
                               lab=NA,  #rownames(volcano_data),
                               #selectLab = c('FABP4','IL1B', 'HIF1A'),
                               xlim=c(-5,5), #may need to change this to fit all points on
                               ylim=c(0, 5), #may need to change this to fit all points on
                               x='logFC',
                               y='FDR',
                               title = "Zilucoplan",
                               titleLabSize =20,
                               subtitle = NULL, #turn off subtitle
                               pCutoff= 0.05, #SET LINES TO SHOW FDR PVAL OF CHOICE
                               FCcutoff=0, #SET LINES TO SHOW FC CUTOFF OF CHOICE
                               colCustom = keyvals,
                               pointSize=3,
                               #legendLabSize = 10,
                               gridlines.major = FALSE,
                               gridlines.minor=FALSE,
                               legendPosition = "none")


# Calculate the number of upregulated and downregulated genes
zil_num_upregulated <- sum(zil_volcano_data$logFC > 0 & !is.na(zil_volcano_data$logFC) & 
                         !is.na(zil_volcano_data$FDR) & zil_volcano_data$FDR < 0.05)
zil_num_downregulated <- sum(zil_volcano_data$logFC < 0 & !is.na(zil_volcano_data$logFC) & 
                           !is.na(zil_volcano_data$FDR) & zil_volcano_data$FDR < 0.05)

final_zil<- zil_volcano_plot + geom_text(aes(x = Inf, y = Inf, label = paste("Upregulated: ", zil_num_upregulated)), size=4, colour = "darkred", hjust = 1.5, vjust = 1, family = "sans") +
  geom_text(aes(x = -Inf, y = Inf, label = paste("Downregulated: ", zil_num_downregulated)), size=4, colour = "navy", hjust = -0.3, vjust = 1, family = "sans")

#combined plots
plot_grid(final_bem, final_toz, final_zil, nrow =3)


```

# 2. Top enriched process barplots

Plotting top enriched Biological Processes. This uses the enrichment_results.csv output from the clusterprofiler enrichment analysis.

```{r fig.width = 15, fig.height = 10}
SoC_upregulated<-read.csv("Standard_of_Care_Day5_vs_Day1_BP_upregulated_enrichment_results.csv", header=TRUE)
colnames(SoC_upregulated)
SoC_downregulated<-read.csv("Standard_of_Care_Day5_vs_Day1_BP_downregulated_enrichment_results.csv", header=TRUE)
#create combo column
SoC_upregulated$GO_ID <- paste(SoC_upregulated$ID, SoC_upregulated$Description, sep= " ")
SoC_downregulated$GO_ID <- paste(SoC_downregulated$ID, SoC_downregulated$Description, sep= " ")

#BARPLOT for UP
ggplot(SoC_upregulated, aes(x=Count, y=reorder(GO_ID, -qvalue), fill=qvalue))+
  ggtitle("Top 10 enriched upregulated BP pathways for SoC")+
  theme_bw(base_size = 20)+
  geom_bar(stat='identity', color="black")+
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5))+ 
  scale_fill_distiller(palette = "Blues", name="FDR")+ #set colour schemes using viridis OR
  scale_y_discrete(labels = wrap_format(60), expand = c(0, 0))+ #this changes if the label axis is shorter/longer by putting on more lines
  xlab("Gene Count")+
  theme(axis.title.y = element_blank())+  # Remove the Y-axis title
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())


#BARPLOT for down
ggplot(SoC_downregulated[1:10, ], aes(x=Count, y=reorder(GO_ID, -qvalue), fill=qvalue))+
  ggtitle("Top 10 enriched downregulated BP pathways for SoC")+
  theme_bw(base_size = 20)+
  geom_bar(stat='identity', color="black")+
  scale_x_continuous(expand = c(0, 0), limits = c(0, 15))+ 
  scale_fill_distiller(palette = "Blues", name="FDR")+ #set colour schemes using viridis OR
  scale_y_discrete(labels = wrap_format(60), expand = c(0, 0))+ #this changes if the label axis is shorter/longer by putting on more lines
  xlab("Gene Count")+
  theme(axis.title.y = element_blank())+  # Remove the Y-axis title
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())
  

```

# 3.  GO bubble plots

The output of the BP enrichment analysis for each treatment arm downregulated and upregulated genes were saved as csv in the 03_differential_expression_treatmentarm notebook. 

These files were input to Cytoscape for further visualisation using the EnrichmentMap plugin. Identified clusters were annotated using AutoAnnotate, followed by manual curation of the cluster names.

Original GO files were used to get mapping aesthetics (e.g. gene count) for the summarised categories

For this - need to create Rdata file containing the cluster analysis output from cytoscape code in "C:/Users/ja2y21/OneDrive - University of Southampton/PostDoc/Projects/EME COVID/Data/Analysis/4_Nested/Results/DE/GO_Comparison/ACCORD_GO_BubblePlot.R"

Load data
.Rdata file contains:
- the BP enrichment output for each of the treatment arms for up or downregulated genes
- the EnrichmentMap and autoannotate clusters saved from cytoscape (down_cluster_data or up_cluster_data)
- manually annotated cluster mapping (up_annot and down_annot). includes the higher level broad categories grouping similar cluster terms for plotting 


```{r}
# Read in  data
load("enrichment_analysis_ouptut_plotting_data.Rdata")
```

Set colours for the plot

```{r}

# Color vector for treatments, outcomes, and expressions
new_colors <- c("Bemcentinib" = "forestgreen",  # Forest Green for Bemcentinib
                "Tozorakimab" = "gold",        # Gold for Tozorakimab
                "Zilucoplan" = "coral",        # Coral for Zilucoplan
                
                "Downregulated" = "navy",
                "Upregulated" = "darkred",
                
                # Color palette for the terms (using ggplot2-friendly colors)
                "Immune cell activation" = "#D32F2F",            # Red
                "Immune response regulation" = "#66cc66",  # Blue
                "Inflammatory mediator signalling" = "#2196F3", #green 
                "Viral life cycle" = "#9C27B0",                   # Purple
                "Protein regulation" = "#FF5722",                 # Deep Orange
                "Gene expression regulation" = "#F57C00",        # Orange
                "Cellular signalling" = "#4CAF50",                 # Green
                "Antiviral immune response" = "#8E24AA",          # Magenta
                "Inflammatory mediator signalling" = "#9E9E9E",   # Grey
                "Coagulation and platelet activation" = "#0288D1", # Blue
                "Apoptosis" = "#FFCC00",                          # Amber
                "Complement" = "#673AB7",                         # Deep Purple
                "Oxidative stress" = "#E91E63",                   # Pink
                "Autophagy" = "#0288D1",                          # Light Blue
                "Molecular process" = "#3F51B5",                  # Indigo
                "Antibacterial response" = "#CDDC39",             # Lime
                "Cellular response" = "#FF5722",                  # Deep Orange
                "Antiviral response" = "#FF9800",                 # Yellow
                "Metabolic process" = "#8BC34A",  # Light Green
                "Cellular process" = "#00Cc00",
                "Ribosomal process" = "brown")

```

Merge new cluster data with the original enrichment files 
First downregulated enrichment:

```{r}
# Transform the data
down_transformed_data <- down_cluster_data %>%
  # Select relevant columns
  select(
    id = `EnrichmentMap..Name`,
    term = `EnrichmentMap..GS_DESCR`,
    gene_count = `EnrichmentMap..gs_size`,
    cluster = X__mclCluster,
    bem_NES = `EnrichmentMap..NES..bem.down.`,
    toz_NES = `EnrichmentMap..NES..toz.down.`,
    zil_NES = `EnrichmentMap..NES..zil.down.`,
    bem_fdr = `EnrichmentMap..fdr_qvalue..bem.down.`,
    toz_fdr = `EnrichmentMap..fdr_qvalue..toz.down.`,
    zil_fdr = `EnrichmentMap..fdr_qvalue..zil.down.`
  ) %>%
  # Pivot longer to reshape NES and fdr values
  pivot_longer(
    cols = starts_with("bem") | starts_with("toz") | starts_with("zil"), # Only reshape treatment-related columns
    names_to = c("treatment", ".value"),
    names_sep = "_"
  ) %>%
  # Filter rows with valid NES values (non-NA)
  filter(!is.na(NES)) %>%
  # Organize columns
  select(id, term, gene_count, cluster, treatment, NES, fdr)


#use the gene count column fro the original GO enrichment files for plotting - e.g. size of bubbles for gene counts
# Combine all treatment BP data frames into one
all_BP_down <- bind_rows(bem_BP_down, toz_BP_down, zil_BP_down) %>%
  select(id = ID, treatment, Count)  # Select necessary columns and rename for consistency
# Update gene_count in transformed_data based on all_BP
down_transformed_data <- down_transformed_data %>%
  left_join(all_BP_down, by = c("id", "treatment")) %>%  # Join on id and treatment
  mutate(gene_count = ifelse(!is.na(Count), Count, gene_count)) %>%  # Replace gene_count if Count is available
  select(-Count)  # Remove Count column after updating


#merge with manually annotated term file "down_annot" to use the higher level GO categories for plotting
down_merge_annot<-merge(down_transformed_data, down_annot, by.x="cluster", by.y = "cluster_n")
# Find unmatched clusters in transformed_data
down_unmatched_clusters <- setdiff(down_transformed_data$cluster, down_annot$cluster_n)
# these were not manually annotated as they were removed from the newtowrk plot on inspection due to forming singleton clusters

#Set plotting aesthetics
# Ensure `term` is ordered by `parent_term`, removing duplicates
down_merge_annot_ordered <- down_merge_annot %>%
  mutate(term = factor(term, levels = unique(term[order(parent_term)])))
#log fdr, better for plotting visualisation
down_merge_annot_ordered <- down_merge_annot_ordered %>%
  mutate(log10_fdr = log10(fdr), log2_fdr=log2(fdr))  # Create log10-transformed fdr column
#change treatment names
down_merge_annot_ordered <- down_merge_annot_ordered %>%
  mutate(treatment = case_when(
    treatment == "bem" ~ "Bemcentinib",  # Corrected syntax for condition and result
    treatment == "toz" ~ "Tozorakimab",  # Same here
    treatment == "zil" ~ "Zilucoplan",   # Same here
    TRUE ~ treatment  # This keeps the other treatments unchanged
  ))

```

Now for the upregulated enrichment:

```{r}
# Transform the upregulated data (similar to how you transformed the downregulated data)
up_transformed_data <- up_cluster_data %>%
  # Select relevant columns and make the necessary adjustments for upregulated data
  select(
    id = `EnrichmentMap..Name`,
    term = `EnrichmentMap..GS_DESCR`,
    gene_count = `EnrichmentMap..gs_size`,
    cluster = X__mclCluster,
    bem_NES = `EnrichmentMap..NES..Bemcentinib.`,
    toz_NES = `EnrichmentMap..NES..Tozorakimab.`,
    zil_NES = `EnrichmentMap..NES..Zilucoplan.`,
    bem_fdr = `EnrichmentMap..fdr_qvalue..Bemcentinib.`,
    toz_fdr = `EnrichmentMap..fdr_qvalue..Tozorakimab.`,
    zil_fdr = `EnrichmentMap..fdr_qvalue..Zilucoplan.`
  ) %>%
  # Pivot longer to reshape NES and fdr values
  pivot_longer(
    cols = starts_with("bem") | starts_with("toz") | starts_with("zil"), 
    names_to = c("treatment", ".value"),
    names_sep = "_"
  ) %>%
  # Filter rows with valid NES values (non-NA)
  filter(!is.na(NES)) %>%
  # Organize columns
  select(id, term, gene_count, cluster, treatment, NES, fdr)

# Combine all upregulated BP data frames into one
all_up_BP <- bind_rows(bem_BP_up, toz_BP_up, zil_BP_up) %>%
  select(id = ID, treatment, Count)  # Select necessary columns and rename for consistency

# Update gene_count in transformed_up_data based on all_up_BP
up_transformed_data <- up_transformed_data %>%
  left_join(all_up_BP, by = c("id", "treatment")) %>%  # Join on id and treatment
  mutate(gene_count = ifelse(!is.na(Count), Count, gene_count)) %>%  # Replace gene_count if Count is available
  select(-Count)  # Remove Count column after updating

# Merge the transformed upregulated data with the manual annotations "up_annot"
up_merge_annot <- merge(up_transformed_data, up_annot, by.x = "cluster", by.y = "cluster_n")

# Find unmatched clusters in transformed_up_data

# Ensure `term` is ordered by `parent_term`, removing duplicates
up_merge_annot_ordered <- up_merge_annot %>%
  mutate(term = factor(term, levels = unique(term[order(parent_term)])))

# Create log transformed fdr column for upregulated data for plotting 
up_merge_annot_ordered <- up_merge_annot_ordered %>%
  mutate(log10_fdr = log10(fdr), log2_fdr=log2(fdr)) 

# Change treatment names for upregulated data
up_merge_annot_ordered <- up_merge_annot_ordered %>%
  mutate(treatment = case_when(
    treatment == "bem" ~ "Bemcentinib",  # Corrected syntax for condition and result
    treatment == "toz" ~ "Tozorakimab",  # Same here
    treatment == "zil" ~ "Zilucoplan",   # Same here
    TRUE ~ treatment  # This keeps the other treatments unchanged
  ))

```


Now plot both as a combined plot

```{r}
## Plot on same axis -----------
# Create regulation indicator for up and downregulated terms
down_merge_annot_ordered <- down_merge_annot_ordered %>%
  mutate(regulation = "Downregulated")
up_merge_annot_ordered <- up_merge_annot_ordered %>%
  mutate(regulation = "Upregulated",
         log10_fdr = -log10(fdr),
         log2_fdr = -log2(fdr))  # No need to change the log10_fdr for upregulated terms

# Combine both data frames (upregulated and downregulated)
combined_data <- bind_rows(down_merge_annot_ordered, up_merge_annot_ordered)

# Custom function to format the y-axis labels as absolute values so they are shown as the same (log10 values)
#as oposed to the plotted values which have been signed to help plot above and below the axis
custom_y_labels <- function(x) {
  # Flip the sign of the y-axis labels and display them as absolute values
  abs_labels <- abs(x)  # Take the absolute value
  return(abs_labels)  # Return the modified labels
}

top_terms <- combined_data %>%
  group_by(treatment) %>%
  # For downregulated (lowest FDR values)
  arrange(treatment, log2_fdr) %>%
  slice_head(n = 5) %>%
  # For upregulated (highest FDR values)
  bind_rows(
    combined_data %>%
      group_by(treatment) %>%
      arrange(treatment, desc(log2_fdr)) %>%
      slice_head(n = 5)
  )
## Wrap the long labels in the 'term' column of the top_terms dataset
top_terms$term_wrapped <- str_wrap(top_terms$term, width = 25)  # Adjust 'width' to control line length

```




```{r fig.height= 10, fig.width=12}
ggplot(combined_data, aes(x = parent_term, y = log2_fdr, size = gene_count, color = parent_term)) +
  geom_jitter(alpha = 0.4, width = 0.5, height = 0) +  # Add jitter on the x-axis only
  scale_y_continuous(name = "log2 GO Enrichment FDR", limits = c(-200, 200), labels = custom_y_labels) +  # Adjust y-axis for FDR
  scale_size_continuous(name = "Gene Count", range = c(1, 10)) +  # Size legend
  scale_color_manual(name = "GO category", values = new_colors) + # Color legend
  theme_bw(base_size=14) +  # Use a clean theme
  theme(
    strip.text = element_text(size = 16),  # Adjust font size here
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",  # Place legend on the right
  #  legend.box = "horizontal",  # Place legend items horizontally
  #  plot.title = element_blank(),  # Turn off the plot title
    legend.key.size = unit(1, "lines"),  # Control the size of legend keys (symbol size)
    legend.text = element_text(size = 10),  # Adjust the size of legend text
    legend.spacing.x = unit(0.1, "cm"),  # Space between legend items
    legend.margin = margin(t = 0)  # Adjust top margin to fit it properly
  ) +
  labs(title = "Upregulated and Downregulated GO Terms") +
  facet_wrap(~ treatment, nrow = 3) +  # Separate up and down regulated terms
  geom_hline(yintercept = 0, color = "black") +  # Add a black horizontal line for the new y-axis
  # Add labels for the top 5 up and downregulated terms with ggrepel to avoid overlap
  geom_label_repel(
    data = top_terms,
    aes(label = term_wrapped), 
    size = 3, color = "black", 
    box.padding = 0.5, point.padding = 0.2,  # Add padding around the label and point
    min.segment.length = 0.5,  # Minimum length of the line connecting the label to the point
    max.overlaps = Inf, # Prevent overlapping of labels
    segment.color = "black", # Color of the segment lines
    force = 1000,  # Apply a force to help avoid overlap
    position = position_jitter(width = 0.5, height = 0),  # Ensure jittered positions for the labels
    show.legend = FALSE  # Hide labels from the legend
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 3), ncol=1)  # Increase dot size for color legend
  )

```

# 4.  Immune pathway heatmap

Now using the specific immune related pathways in the manually annotated cluster file to visualise the expression of these indiviual immune processes uisng the logfc of genes assigned to the immune processes. We are using the up/down_merge_annot_ordered df created in (3) to merge with the specific genes identified in each go term. The logfc of all the genes in the go terms within a single specific_immune_term will then be summarised for heatmap plotting. 



Downregulated
```{r}
#Bemcentinib
#select only the genes/name and their logfc 
down_bem_genes_col <- select(bem_degs, c("genes", "logFC", "gene_name")) #downregulated genes
# from the combined annotated downregulatd enrichment, select only treatment of interest
down_bem_cluster<- down_merge_annot_ordered %>% # go terms and their cluster assignment
  filter(treatment == "Bemcentinib")
# split out the genes from each go term into individual rows in order to merge with logfc data
down_bem_enrichment <- bem_BP_down %>%
  mutate(geneID = strsplit(as.character(geneID), "/")) %>%
  unnest(geneID)
#merge logfc data (down_bem_genes_col) with the enrichment df
down_bem_merged_data_genes <- down_bem_enrichment %>%
  left_join(down_bem_genes_col, by = c("geneID" = "gene_name"))
# now join with the annotated cluster data 
down_bem_merged_data_cluster <- down_bem_merged_data_genes %>%
  left_join(down_bem_cluster, by = c("ID" = "id"))
# summarise the logfc for each of the specific immune terms 
down_bem_average_fc <- down_bem_merged_data_cluster %>%
  group_by(specific_immune_term) %>%
  summarise(Bemcentinib = mean(logFC, na.rm = TRUE)) %>%
  ungroup()


#Tozorakimab
#select only the genes/name and their logfc 
down_toz_genes_col <- select(toz_degs, c("genes", "logFC", "gene_name")) #downregulated genes
# from the combined annotated downregulated enrichment, select only treatment of interest
down_toz_cluster<- down_merge_annot_ordered %>% # go terms and their cluster assignment
  filter(treatment == "Tozorakimab")
# split out the genes from each go term into individual rows in order to merge with logfc data
down_toz_enrichment <- toz_BP_down %>%
  mutate(geneID = strsplit(as.character(geneID), "/")) %>%
  unnest(geneID)
#merge logfc data (down_bem_genes_col) with the enrichment df
down_toz_merged_data_genes <- down_toz_enrichment %>%
  left_join(down_toz_genes_col, by = c("geneID" = "gene_name"))
# now join with the annotated cluster data 
down_toz_merged_data_cluster <- down_toz_merged_data_genes %>%
  left_join(down_toz_cluster, by = c("ID" = "id"))
# summarise the logfc for each of the specific immune terms 
down_toz_average_fc <- down_toz_merged_data_cluster %>%
  group_by(specific_immune_term) %>%
  summarise(Tozorakimab = mean(logFC, na.rm = TRUE)) %>%
  ungroup()



#Zilucoplan
down_zil_genes_col <- select(zil_degs, c("genes", "logFC", "gene_name")) #downregulated genes
# from the combined annotated downregulated enrichment, select only treatment of interest
down_zil_cluster<- down_merge_annot_ordered %>% # go terms and their cluster assignment
  filter(treatment == "Zilucoplan")
# split out the genes from each go term into individual rows in order to merge with logfc data
down_zil_enrichment <- zil_BP_down %>%
  mutate(geneID = strsplit(as.character(geneID), "/")) %>%
  unnest(geneID)
#merge logfc data (down_bem_genes_col) with the enrichment df
down_zil_merged_data_genes <- down_zil_enrichment %>%
  left_join(down_zil_genes_col, by = c("geneID" = "gene_name"))
# now join with the annotated cluster data 
down_zil_merged_data_cluster <- down_zil_merged_data_genes %>%
  left_join(down_zil_cluster, by = c("ID" = "id"))
# summarise the logfc for each of the specific immune terms 
down_zil_average_fc <- down_zil_merged_data_cluster %>%
  group_by(specific_immune_term) %>%
  summarise(Zilucoplan = mean(logFC, na.rm = TRUE)) %>%
  ungroup()


# Replace missing avg_logFC values with 0 for each treatment df
down_bem_average_fc <- down_bem_average_fc %>%
  mutate(Bemcentinib = replace_na(Bemcentinib, 0))

down_toz_average_fc <- down_toz_average_fc %>%
  mutate(Tozorakimab = replace_na(Tozorakimab, 0))

down_zil_average_fc <- down_zil_average_fc %>%
  mutate(Zilucoplan = replace_na(Zilucoplan, 0))

# Perform full join to combine all data
down_combined_df <- down_bem_average_fc %>%
  full_join(down_toz_average_fc, by = "specific_immune_term") %>%
  full_join(down_zil_average_fc, by = "specific_immune_term") %>%
  mutate(
    Bemcentinib = replace_na(Bemcentinib, 0),
    Tozorakimab = replace_na(Tozorakimab, 0),
    Zilucoplan = replace_na(Zilucoplan, 0)
  )%>% 
  filter(specific_immune_term!="")

# Create a matrix for the heatmap
down_heatmap_matrix <- down_combined_df %>%
  column_to_rownames("specific_immune_term") %>%  # Use terms as row names
  as.matrix()  # Convert to matrix format

```



upregulated
```{r}
#Bemcentinib
#select only the genes/name and their logfc 
up_bem_genes_col <- select(bem_degs, c("genes", "logFC", "gene_name")) #upregulated genes
# from the combined annotated upregulatd enrichment, select only treatment of interest
up_bem_cluster<- up_merge_annot_ordered %>% # go terms and their cluster assignment
  filter(treatment == "Bemcentinib")
# split out the genes from each go term into individual rows in order to merge with logfc data
up_bem_enrichment <- bem_BP_up %>%
  mutate(geneID = strsplit(as.character(geneID), "/")) %>%
  unnest(geneID)
#merge logfc data (up_bem_genes_col) with the enrichment df
up_bem_merged_data_genes <- up_bem_enrichment %>%
  left_join(up_bem_genes_col, by = c("geneID" = "gene_name"))
# now join with the annotated cluster data 
up_bem_merged_data_cluster <- up_bem_merged_data_genes %>%
  left_join(up_bem_cluster, by = c("ID" = "id"))
# summarise the logfc for each of the specific immune terms 
up_bem_average_fc <- up_bem_merged_data_cluster %>%
  group_by(specific_immune_term) %>%
  summarise(Bemcentinib = mean(logFC, na.rm = TRUE)) %>%
  ungroup()




#Tozorakimab
#select only the genes/name and their logfc 
down_toz_genes_col <- select(toz_degs, c("genes", "logFC", "gene_name")) #downregulated genes
# from the combined annotated downregulatd enrichment, select only treatment of interest
down_toz_cluster<- down_merge_annot_ordered %>% # go terms and their cluster assignment
  filter(treatment == "Tozorakimab")
# split out the genes from each go term into individual rows in order to merge with logfc data
down_toz_enrichment <- toz_BP_down %>%
  mutate(geneID = strsplit(as.character(geneID), "/")) %>%
  unnest(geneID)
#merge logfc data (down_toz_genes_col) with the enrichment df
down_toz_merged_data_genes <- down_toz_enrichment %>%
  left_join(down_toz_genes_col, by = c("geneID" = "gene_name"))
# now join with the annotated cluster data 
down_toz_merged_data_cluster <- down_toz_merged_data_genes %>%
  left_join(down_toz_cluster, by = c("ID" = "id"))
# summarise the logfc for each of the specific immune terms 
down_toz_average_fc <- down_toz_merged_data_cluster %>%
  group_by(specific_immune_term) %>%
  summarise(Tozorakimab = mean(logFC, na.rm = TRUE)) %>%
  ungroup()


#Zilucoplan
#select only the genes/name and their logfc 
down_zil_genes_col <- select(zil_degs, c("genes", "logFC", "gene_name")) #downregulated genes
# from the combined annotated downregulatd enrichment, select only treatment of interest
down_zil_cluster<- down_merge_annot_ordered %>% # go terms and their cluster assignment
  filter(treatment == "Zilucoplan")
# split out the genes from each go term into individual rows in order to merge with logfc data
down_zil_enrichment <- zil_BP_down %>%
  mutate(geneID = strsplit(as.character(geneID), "/")) %>%
  unnest(geneID)
#merge logfc data (down_zil_genes_col) with the enrichment df
down_zil_merged_data_genes <- down_zil_enrichment %>%
  left_join(down_zil_genes_col, by = c("geneID" = "gene_name"))
# now join with the annotated cluster data 
down_zil_merged_data_cluster <- down_zil_merged_data_genes %>%
  left_join(down_zil_cluster, by = c("ID" = "id"))
# summarise the logfc for each of the specific immune terms 
down_zil_average_fc <- down_zil_merged_data_cluster %>%
  group_by(specific_immune_term) %>%
  summarise(Zilucoplan = mean(logFC, na.rm = TRUE)) %>%
  ungroup()


# Replace missing avg_logFC values with 0
up_bem_average_fc <- up_bem_average_fc %>%
  mutate(Bemcentinib = replace_na(Bemcentinib, 0))

up_toz_average_fc <- up_toz_average_fc %>%
  mutate(Tozorakimab = replace_na(Tozorakimab, 0))

up_zil_average_fc <- up_zil_average_fc %>%
  mutate(Zilucoplan = replace_na(Zilucoplan, 0))

# Perform full join to combine all data
up_combined_df <- up_bem_average_fc %>%
  full_join(up_toz_average_fc, by = "specific_immune_term") %>%
  full_join(up_zil_average_fc, by = "specific_immune_term") %>%
  mutate(
    Bemcentinib = replace_na(Bemcentinib, 0),
    Tozorakimab = replace_na(Tozorakimab, 0),
    Zilucoplan = replace_na(Zilucoplan, 0)
  )%>% 
  filter(specific_immune_term!="")

# Create a matrix for the heatmap
up_heatmap_matrix <- up_combined_df %>%
  column_to_rownames("specific_immune_term") %>%  # Use terms as row names
  as.matrix()  # Convert to matrix format

```

Combined up and down for plotting

```{r}
# combine up and down for plotting
# Clean up column names by removing "up_" and "down_" prefixes ----- if present
colnames(up_combined_df) <- gsub("^up_", "", colnames(up_combined_df))
colnames(down_combined_df) <- gsub("^down_", "", colnames(down_combined_df))

#  Combine the data using 'specific_immune_term'
final_combined_df <- full_join(up_combined_df, down_combined_df, by = "specific_immune_term")

# Replace 0s with NA so we can combine based on actual values
final_combined_df[final_combined_df == 0] <- NA

# Combine up and down values into a single column for each treatment,
# using pmax to keep the non-NA values (and allow both negative and positive values)
final_combined_df <- final_combined_df %>%
  mutate(
    Bemcentinib = pmax(Bemcentinib.x, Bemcentinib.y, na.rm = TRUE),
    Tozorakimab = pmax(Tozorakimab.x, Tozorakimab.y, na.rm = TRUE),
    Zilucoplan = pmax(Zilucoplan.x, Zilucoplan.y, na.rm = TRUE)
  ) %>%
  select(specific_immune_term, Bemcentinib, Tozorakimab, Zilucoplan)  # Keep only the combined treatment columns
# Replace any remaining NA values with 0 
final_combined_df[is.na(final_combined_df)] <- 0
# Set 'specific_immune_term' as row names and convert to matrix for the heatmap
final_combined_matrix <- final_combined_df %>%
  column_to_rownames("specific_immune_term") %>%
  as.matrix()
```

Heatmap aesthetics
```{r}
# Combine down_categories and up_categories parent_term mapping
parent_term_mapping <- rbind(
  down_annot[, c("specific_immune_term", "parent_term")],
  up_annot[, c("specific_immune_term", "parent_term")]
)
# Keep only unique rows
parent_term_mapping <- unique(parent_term_mapping)%>%
  filter(specific_immune_term != "")

# set annotations
annotation_row = data.frame(
  Immune_category = parent_term_mapping$parent_term, 
  row.names = parent_term_mapping$specific_immune_term)
#set colours
ann_colours<-list("Immune_category" =c(
  "Immune cell activation" = "#D32F2F",            # Red
  "Immune response regulation" = "#66cc66",       # Green
  "Inflammatory mediator signalling" = "#2196F3"  # Blue
))
# row annotations
row_annotation <- rowAnnotation(
  Immune_category = annotation_row$Immune_category,
  col = list(Immune_category = ann_colours$Immune_category),
 # annotation_name = "Immune Category",  # Set the custom title
   show_annotation_name = FALSE,  # Show the annotation title
 annotation_legend_param = list(
   title = "Immune Category"))

#REORDER COLUIMNS
# order based on the mean of each treatment column across all rows
ordered_columns <- final_combined_df %>%
  select(Bemcentinib, Tozorakimab, Zilucoplan) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  gather(key = "Treatment", value = "Mean") %>%
  arrange(Mean) %>%
  pull(Treatment)
# Reorder the matrix columns based on the sorted treatment names
final_combined_matrix <- final_combined_matrix[, ordered_columns]
# Ensure row annotation order matches matrix rows
annotation_row <- annotation_row[rownames(final_combined_matrix), , drop = FALSE]
#plotting aesthetics
col = circlize::colorRamp2(c(-1, 0, 1), hcl_palette ="RdBu", reverse = TRUE)
```
Finally plot the heatmap
```{r fig.height=10, fig.width=7}
# Plot the heatmap with the columns ordered as desired
set.seed(1234)
heatmap<-ComplexHeatmap::pheatmap(
  final_combined_matrix,
  col = col,
  cluster_rows = TRUE,  # Hierarchical clustering for rows
  cluster_cols = FALSE,  # Do not cluster columns, as they are now ordered
  main = "Immune response pathways",
  border = TRUE,
  border_color = "darkgrey",
  cellheight = 15,
  cellwidth = 15,
  fontsize = 14,
  heatmap_legend_param = list(
    legend_direction = "horizontal", 
    legend_width = unit(5, "cm"),
    title= "logFC"), 
  angle_col = "45",
  #column_names_rot = 45, column_names_side = "top",
  #annotation_row = annotation_row,
  annotation_colors = ann_colours, 
 # annotation_legend_side = 'right',
 # left_annotation = NULL,  # No annotations on the left
  right_annotation = row_annotation) # Move row annotation to the right)

draw(heatmap, heatmap_legend_side = "bottom")

```
# 5.  Enriched process overlap between treatments - upsetplot
Visualise the overlap of enriched terms from the up/down enrichment results for each treatment arm (saved in enrichment_analysis_plotting_data.Rdata)

```{r}

# Extract the Description columns for both "up" and "down" data as these will be used to compare across treatment arms
bem_desc_down <- bem_BP_down$Description
bem_desc_up <- bem_BP_up$Description
toz_desc_down <- toz_BP_down$Description
toz_desc_up <- toz_BP_up$Description
zil_desc_down <- zil_BP_down$Description
zil_desc_up <- zil_BP_up$Description
# Create a list of sets of these desc
sets <- list(
  Bem_BP_down = bem_desc_down,
  Bem_BP_up = bem_desc_up,
  Toz_BP_down = toz_desc_down,
  Toz_BP_up = toz_desc_up,
  Zil_BP_down = zil_desc_down,
  Zil_BP_up = zil_desc_up
)

# Create a binary matrix indicating presence/absence of each description in each set
all_descriptions <- unique(c(bem_desc_down, bem_desc_up, toz_desc_down, toz_desc_up, zil_desc_down, zil_desc_up))
binary_matrix <- data.frame(Description = all_descriptions)

for (set_name in names(sets)) {
  binary_matrix[[set_name]] <- as.logical(all_descriptions %in% sets[[set_name]])
}

#change column names
current_colnames <- colnames(binary_matrix)
# Replace "BP" with an empty string in each column name
new_colnames <- gsub("_BP", "", current_colnames)
# Add additional descriptions
new_colnames[new_colnames == "Bem_down"] <- "Bemcentinib downregulated"
new_colnames[new_colnames == "Bem_up"] <- "Bemcentinib upregulated"
new_colnames[new_colnames == "Toz_down"] <- "Tozorakimab downregulated"
new_colnames[new_colnames == "Toz_up"] <- "Tozorakimab upregulated"
new_colnames[new_colnames == "Zil_down"] <- "Zilucoplan downregulated"
new_colnames[new_colnames == "Zil_up"] <- "Zilucoplan upregulated"

# Assign modified column names to the binary matrix
colnames(binary_matrix) <- new_colnames
#also for vector
names <- c("Bem_BP_down", "Bem_BP_up", "Toz_BP_down", "Toz_BP_up", "Zil_BP_down", "Zil_BP_up")
# Define categories as the column names of binary_matrix excluding "Description"
categories <- colnames(binary_matrix)[-1]

# plot the upset plot
upset(
  binary_matrix,
  categories,
  name = 'Treatment Arm',
  queries=list(
    upset_query(set='Bemcentinib downregulated', fill='forestgreen',color="black"),
    upset_query(set='Bemcentinib upregulated', fill=alpha('forestgreen', 0.5),color="black"),
    upset_query(set='Tozorakimab downregulated', fill=alpha('yellow', 0.75),color="black"),
    upset_query(set='Tozorakimab upregulated', fill='yellow',color="black"),
    upset_query(set='Zilucoplan downregulated', fill='red',color="black"),
    upset_query(set='Zilucoplan upregulated', fill=alpha('red', 0.5),color="black")), 
  
  matrix=(
    intersection_matrix(
      geom=geom_point(
        shape=21,
        size=3.5),
      #turn off borders(will get double border otherwise)
      outline_color=list(
        active=NA,
        inactive=NA))),
  
  base_annotations = list(
    'Intersection size' = intersection_size(bar_number_threshold=1,  # show all numbers on top of bars
                                            mapping = aes(fill = 'bars_color', color='bars_color'),
                                            text = list(color = "black")) +
      
      ggtitle("Intersection sizes: Biological Processes")+
      scale_fill_manual(values = alpha("grey", 0.5), guide = "none") + # Set alpha for fill color
      scale_color_manual(values = "black", guide = "none") + # Set border color to black
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
            axis.line = element_line(color = "black"), axis.title.y = element_blank()) +
      scale_y_continuous(expand = expansion(mult = c(0, .1)))),
  
  set_sizes=(
    upset_set_size(geom=geom_bar(width=0.8), mapping = aes(fill = 'bars_color', color='bars_color'))
    + theme(
      axis.line.x=element_line(colour='black'),
      axis.ticks.x=element_line()+ 
        scale_y_continuous(expand = expansion(mult = c(0, .1)))
    )))



```



# 6.  Unique and shared DEGs - venn and bubbleplot

Now looking at the overlap of DEGs between the treatment arms. 
Here we isolate the shared and unique genes for each intersection, perform enrichment analysis 

First plot venn diagram

```{r}

# Bemcentinib
up_bem_genes_ids <- as.character(bem_degs$ensembl_gene_id[bem_degs$logFC > 0])
down_bem_genes_ids <- as.character(bem_degs$ensembl_gene_id[bem_degs$logFC < 0])

# Tozorakimab
up_toz_genes_ids <- as.character(toz_degs$ensembl_gene_id[toz_degs$logFC > 0])
down_toz_genes_ids <- as.character(toz_degs$ensembl_gene_id[toz_degs$logFC < 0])
# Zilucoplan
up_Zil_genes_ids <- as.character(zil_degs$ensembl_gene_id[zil_degs$logFC > 0])
down_Zil_genes_ids <- as.character(zil_degs$ensembl_gene_id[zil_degs$logFC < 0])


# Create a list of up-regulated genes for each drug
up_genes_list <- list(
  Bemcentinib = up_bem_genes_ids,
  Tozorakimab = up_toz_genes_ids,
  Zilucoplan = up_Zil_genes_ids
)

# Create a list of down-regulated genes for each drug
down_genes_list <- list(
  Bemcentinib = down_bem_genes_ids,
  Tozorakimab = down_toz_genes_ids,
  Zilucoplan = down_Zil_genes_ids
)

ggvenn(
  down_genes_list, 
  fill_color = c("forestgreen", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 3, text_size =2)+
  ggtitle("Downregulated DEGs")+
  theme(plot.title = element_text(size = 10, face = "bold"))

ggvenn(
  up_genes_list, 
  fill_color = c("forestgreen", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 3, text_size =2)+
  ggtitle("Upregulated DEGs")+
  theme(plot.title = element_text(size = 10, face = "bold"))
```

```{r}
# Bemcentinib
up_bem_genes_ids <- as.character(bem_degs$ensembl_gene_id[bem_degs$logFC > 0])
down_bem_genes_ids <- as.character(bem_degs$ensembl_gene_id[bem_degs$logFC < 0])

# Tozorakimab
up_toz_genes_ids <- as.character(toz_degs$ensembl_gene_id[toz_degs$logFC > 0])
down_toz_genes_ids <- as.character(toz_degs$ensembl_gene_id[toz_degs$logFC < 0])
# Zilucoplan
up_Zil_genes_ids <- as.character(zil_degs$ensembl_gene_id[zil_degs$logFC > 0])
down_Zil_genes_ids <- as.character(zil_degs$ensembl_gene_id[zil_degs$logFC < 0])


# Create a list of up-regulated genes for each drug
up_genes_list <- list(
  Bemcentinib = up_bem_genes_ids,
  Tozorakimab = up_toz_genes_ids,
  Zilucoplan = up_Zil_genes_ids
)

# Create a list of down-regulated genes for each drug
down_genes_list <- list(
  Bemcentinib = down_bem_genes_ids,
  Tozorakimab = down_toz_genes_ids,
  Zilucoplan = down_Zil_genes_ids
)



ggvenn(
  down_genes_list, 
  fill_color = c("forestgreen", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 3, text_size =2)+
  ggtitle("Downregulated DEGs")+
  theme(plot.title = element_text(size = 10, face = "bold"))

ggvenn(
  up_genes_list, 
  fill_color = c("forestgreen", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 3, text_size =2)+
  ggtitle("Upregulated DEGs")+
  theme(plot.title = element_text(size = 10, face = "bold"))



## enrichment
# need to extract the genes from the intersection
#down
names(down_genes_list) <- c("Bemcentinib","Tozorakimab","Zilucoplan") 
data_down <- get.venn.partitions(down_genes_list, force.unique = TRUE, keep.elements = TRUE, hierarchical = FALSE)
# Extract relevant columns
# Convert the "..values.." column to a character vector
data_down$..values.. <- sapply(data_down$..values.., function(x) paste(x, collapse = ","))
# Convert the list to a data frame
down_partition_df <- as.data.frame(data_down)
# Extract relevant values for Zilucoplan, Bemcentinib, Tozorakimab for the intersections
down_zil_values <- data_down$`..values..`[data_down$Zilucoplan & !data_down$Bemcentinib & !data_down$Tozorakimab][[1]] #zil only
down_bem_values <- data_down$`..values..`[!data_down$Zilucoplan &  data_down$Bemcentinib & !data_down$Tozorakimab][[1]] #Bemcentinib only
down_toz_values <- data_down$`..values..`[!data_down$Zilucoplan &  !data_down$Bemcentinib & data_down$Tozorakimab][[1]] #toz only
down_shared_values <- data_down$`..values..`[data_down$Zilucoplan &  data_down$Bemcentinib & data_down$Tozorakimab][[1]] # shared between all 3
down_shared_bem_toz <- data_down$`..values..`[!data_down$Zilucoplan & data_down$Bemcentinib & data_down$Tozorakimab][[1]] #shared between Bemcentinib and toz
down_shared_zil_toz <- data_down$`..values..`[data_down$Zilucoplan & !data_down$Bemcentinib & data_down$Tozorakimab][[1]] #shared between toz and zil

# Zilucoplan
# Create dataframes for each condition with the condition name as the column name
down_zil_df <- data.frame(Zilucoplan = down_zil_values)
row.names(down_zil_df) <- down_zil_df$Zilucoplan
# Split the string into individual genes
down_zil_values_split <- strsplit(down_zil_values, ",")[[1]]
# Create a data frame with genes as individual columns
down_zil_df <- data.frame((down_zil_values_split))  # Transpose to make genes columns
# Assign proper column names
colnames(down_zil_df) <- "Zilucoplan"
row.names(down_zil_df) <- down_zil_df$Zilucoplan
# Display the resulting data frame
down_zil_df

# Bemcentinib Data
down_bem_values_split <- strsplit(down_bem_values, ",")[[1]]
down_bem_df <- data.frame(down_bem_values_split)  # Create a data frame
colnames(down_bem_df) <- "Bemcentinib"  # Assign column name
row.names(down_bem_df) <- down_bem_df$Bemcentinib  # Set row names

# Tozorakimab Data
down_toz_values_split <- strsplit(down_toz_values, ",")[[1]]
down_toz_df <- data.frame(down_toz_values_split)  # Create a data frame
colnames(down_toz_df) <- "Tozorakimab"  # Assign column name
row.names(down_toz_df) <- down_toz_df$Tozorakimab  # Set row names

# Shared Data
down_shared_values_split <- strsplit(down_shared_values, ",")[[1]]
down_shared_df <- data.frame(down_shared_values_split)  # Create a data frame
colnames(down_shared_df) <- "Shared"  # Assign column name
row.names(down_shared_df) <- down_shared_df$Shared  # Set row names

# Shared Bemcentinib and Tozorakimab Data
down_shared_bem_toz_split <- strsplit(down_shared_bem_toz, ",")[[1]]
down_shared_bem_toz_df <- data.frame(down_shared_bem_toz_split)  # Create a data frame
colnames(down_shared_bem_toz_df) <- "Shared"  # Assign column name
row.names(down_shared_bem_toz_df) <- down_shared_bem_toz_df$Shared  # Set row names

# Shared Zilucoplan and Tozorakimab Data
down_shared_zil_toz_split <- strsplit(down_shared_zil_toz, ",")[[1]]
down_shared_zil_toz_df <- data.frame(down_shared_zil_toz_split)  # Create a data frame
colnames(down_shared_zil_toz_df) <- "Shared"  # Assign column name
row.names(down_shared_zil_toz_df) <- down_shared_zil_toz_df$Shared  # Set row names


# Define the set of  genes
down_Zil_set <- down_zil_df$Zilucoplan
down_toz_set<-down_toz_df$Tozorakimab
down_Bem_set<-down_bem_df$Bemcentinib
down_Shared_set<-down_shared_results$Shared
down_shared_bem_toz<-down_shared_bem_toz_df$Shared
down_shared_toz_zil<-down_shared_zil_toz_df$Shared

# now for up
## enrichment
# need to extract the genes from the intersection
#up
names(up_genes_list) <- c("Bemcentinib","Tozorakimab","Zilucoplan") 
data_up<- get.venn.partitions(up_genes_list, force.unique = TRUE, keep.elements = TRUE, hierarchical = FALSE)
# Extract relevant columns
# Convert the "..values.." column to a character vector
data_up$..values.. <- sapply(data_up$..values.., function(x) paste(x, collapse = ","))
# Convert the list to a data frame
up_partition_df <- as.data.frame(data_up)
# Extract relevant values for Zilucoplan, Bemcentinib, Tozorakimab for the intersections
up_zil_values <- data_up$`..values..`[data_up$Zilucoplan & !data_up$Bemcentinib & !data_up$Tozorakimab][[1]] #zil only
up_bem_values <- data_up$`..values..`[!data_up$Zilucoplan &  data_up$Bemcentinib & !data_up$Tozorakimab][[1]] #Bemcentinib only
up_toz_values <- data_up$`..values..`[!data_up$Zilucoplan &  !data_up$Bemcentinib & data_up$Tozorakimab][[1]] #toz only
up_shared_values <- data_up$`..values..`[data_up$Zilucoplan &  data_up$Bemcentinib & data_up$Tozorakimab][[1]] # shared between all 3
up_shared_bem_toz <- data_up$`..values..`[!data_up$Zilucoplan & data_up$Bemcentinib & data_up$Tozorakimab][[1]] #shared between Bemcentinib and toz
up_shared_zil_toz <- data_up$`..values..`[data_up$Zilucoplan & !data_up$Bemcentinib & data_up$Tozorakimab][[1]] #shared between toz and zil


# Zilucoplan
# Create dataframes for each condition with the condition name as the column name
up_zil_df <- data.frame(Zilucoplan = up_zil_values)
row.names(up_zil_df) <- up_zil_df$Zilucoplan
# Split the string into individual genes
up_zil_values_split <- strsplit(up_zil_values, ",")[[1]]
# Create a data frame with genes as individual columns
up_zil_df <- data.frame((up_zil_values_split))  # Transpose to make genes columns
# Assign proper column names
colnames(up_zil_df) <- "Zilucoplan"
row.names(up_zil_df) <- up_zil_df$Zilucoplan
# Display the resulting data frame
up_zil_df

# Bemcentinib Data
up_bem_values_split <- strsplit(up_bem_values, ",")[[1]]
up_bem_df <- data.frame(up_bem_values_split)  # Create a data frame
colnames(up_bem_df) <- "Bemcentinib"  # Assign column name
row.names(up_bem_df) <- up_bem_df$Bemcentinib  # Set row names

# Tozorakimab Data
up_toz_values_split <- strsplit(up_toz_values, ",")[[1]]
up_toz_df <- data.frame(up_toz_values_split)  # Create a data frame
colnames(up_toz_df) <- "Tozorakimab"  # Assign column name
row.names(up_toz_df) <- up_toz_df$Tozorakimab  # Set row names

# Shared Data
up_shared_values_split <- strsplit(up_shared_values, ",")[[1]]
up_shared_df <- data.frame(up_shared_values_split)  # Create a data frame
colnames(up_shared_df) <- "Shared"  # Assign column name
row.names(up_shared_df) <- up_shared_df$Shared  # Set row names

# Shared Bemcentinib and Tozorakimab Data
up_shared_bem_toz_split <- strsplit(up_shared_bem_toz, ",")[[1]]
up_shared_bem_toz_df <- data.frame(up_shared_bem_toz_split)  # Create a data frame
colnames(up_shared_bem_toz_df) <- "Shared"  # Assign column name
row.names(up_shared_bem_toz_df) <- up_shared_bem_toz_df$Shared  # Set row names

# Shared Zilucoplan and Tozorakimab Data
up_shared_zil_toz_split <- strsplit(up_shared_zil_toz, ",")[[1]]
up_shared_zil_toz_df <- data.frame(up_shared_zil_toz_split)  # Create a data frame
colnames(up_shared_zil_toz_df) <- "Shared"  # Assign column name
row.names(up_shared_zil_toz_df) <- up_shared_zil_toz_df$Shared  # Set row names


# Define the set of  genes
up_Zil_set <- up_zil_df$Zilucoplan
up_toz_set<-up_toz_df$Tozorakimab
up_Bem_set<-up_bem_df$Bemcentinib
up_shared_set<-up_shared_df$Shared
up_shared_bem_toz<-up_shared_bem_toz_df$Shared
up_shared_toz_zil<-up_shared_zil_toz_df$Shared




# Define the background gene set
#here using soc
load("soc_data.Rdata")
background_genes <- row.names(soc_counts)
#down
gene_sets <- list(
  down_zil_set=down_zil_df$Zilucoplan,
  down_toz_set=down_toz_df$Tozorakimab,
  down_bem_set=down_bem_df$Bemcentinib,
  down_shared_set=down_shared_df$Shared,
  down_shared_bem_toz=down_shared_bem_toz_df$Shared,
  down_shared_toz_zil=down_shared_zil_toz_df$Shared,
  up_zil_set=up_zil_df$Zilucoplan,
  up_toz_set=up_toz_df$Tozorakimab,
  up_bem_set=up_bem_df$Bemcentinib,
  up_shared_set=up_shared_df$Shared,
  up_shared_bem_toz=up_shared_bem_toz_df$Shared,
  up_shared_toz_zil=up_shared_zil_toz_df$Shared)

# Set up plot parameters
plot_width <- 8
plot_height <- 10
#define root.name
root.name <- "v2_Drug_DEG_overlap"
# List of GO categories
go_categories <- c("MF", "BP", "CC")

# Loop through each gene set
for (gene_set_name in names(gene_sets)) {
  gene_set <- gene_sets[[gene_set_name]]
  
  for (go in go_categories) {
    # Perform GO enrichment analysis for the current gene set
    ego <- tryCatch({
      enrichGO(gene = gene_set,
               universe = background_genes,
               OrgDb = "org.Hs.eg.db",
               keyType = "ENSEMBL",
               ont = go,
               pAdjustMethod = "BH",
               qvalueCutoff = 0.05,
               readable = TRUE)
    }, error = function(e) NULL)
    
    if (!is.null(ego$geneID) && nrow(ego) > 0) {
      # Barplot for top 20 enriched pathways for the current gene set
      bar_plot <- barplot(ego, showCategory = 20) +
        ggtitle(paste("Top 20 enriched", go, "pathways for", gene_set_name)) +
        theme_minimal()
      
      # Save plot as PDF
      ggsave(filename = paste(root.name, gene_set_name, go, "_barplot.pdf", sep = ""), plot = bar_plot, width = plot_width, height = plot_height)
    } else {
      cat(paste("No enriched terms found for", go, "in the", gene_set_name, ". Skipping..."))
    }
    
    # Save enrichment results as CSV
    if (!is.null(ego$geneID) && nrow(ego) > 0) {
      write.csv(ego, file = paste(root.name, gene_set_name, go, "_enrichment_results.csv", sep = ""), row.names = FALSE)
    } else {
      cat(paste("No enriched terms found for", go, "in the", gene_set_name, ". Skipping..."))
    }
  }
}

```

Now use these enrichment results to see if similar or unique processes are enriched by unique and shared degs across treatments

```{r}

## there are a significant number of shared terms between toz and bem 
# expected due to the number of shared degs
# to be able to plot, keep only the bem/toz shared degs which also appear in the bem_unique section
#these files will be output from the above enrichment analysis as csv, will need to read back in
bem_unique <- read.csv("Drug_DEG_overlapdown_Bem_setBP_enrichment_results.csv") %>%
  mutate(overlap = "bem_unique", regulation = "downregulated")
bem_toz_shared <- read.csv("Drug_DEG_overlapdown_shared_bem_tozBP_enrichment_results.csv") %>%
  mutate(overlap = "bem_toz_shared", regulation = "downregulated")
shared_terms <- intersect(bem_unique$ID, bem_toz_shared$ID)
filtered_bem_data <- bem_unique %>%
  filter(ID %in% shared_terms) %>%
  bind_rows(bem_toz_shared %>% filter(ID %in% shared_terms))

#  read in remaining data and process
other_files <- list(
  bem_toz_zil_shared_down = list(file = "Drug_DEG_overlapdown_Shared_setBP_enrichment_results.csv", overlap = "bem_toz_zil_shared", regulation = "downregulated"),
  toz_zil_shared_down = list(file = "Drug_DEG_overlapdown_shared_toz_zilBP_enrichment_results.csv", overlap = "toz_zil_shared", regulation = "downregulated"),
  toz_zil_shared_up = list(file = "Drug_DEG_overlapup_shared_toz_zilBP_enrichment_results.csv", overlap = "toz_zil_shared", regulation = "upregulated"),
  bem_toz_shared_up = list(file = "Drug_DEG_overlapup_shared_bem_tozBP_enrichment_results.csv", overlap = "bem_toz_shared", regulation = "upregulated"),
  toz_unique = list(file = "Drug_DEG_overlapup_toz_setBP_enrichment_results.csv", overlap = "toz_unique", regulation = "upregulated"),
  zil_unique = list(file = "Drug_DEG_overlapup_Zil_setBP_enrichment_results.csv", overlap = "zil_unique", regulation = "upregulated")
)

other_data <- bind_rows(lapply(other_files, function(info) {
  read.csv(info$file) %>%
    mutate(overlap = info$overlap, regulation = info$regulation)
}))

#combine Filtered and Other Data
final_data <- bind_rows(filtered_bem_data, other_data)

# now add new columns for each treatment indicating whether it is associated with a specific term
final_data_with_treatments <- final_data %>%
  mutate(
    # Add columns for each treatment and assign "yes" or "no"
    bemcentinib = case_when(
      str_detect(overlap, "bem") ~ "yes", 
      TRUE ~ "no"
    ),
    tozorakimab = case_when(
      str_detect(overlap, "toz") ~ "yes", 
      TRUE ~ "no"
    ),
    zilucoplan = case_when(
      str_detect(overlap, "zil") ~ "yes", 
      TRUE ~ "no"
    )
  )

# Now add 
#  Reshape the data to have a "treatment" column
final_data_treatment <- final_data_with_treatments %>%
  # Step 2: Pivot the data longer to create a single 'treatment' column
  pivot_longer(cols = c(bemcentinib, tozorakimab, zilucoplan), 
               names_to = "treatment_type", 
               values_to = "has_treatment") %>%
  # Step 3: Filter rows where "has_treatment" is "yes"
  filter(has_treatment == "yes") %>%
  # Step 4: Create the new 'treatment' column based on the treatment type
  mutate(treatment = case_when(
    treatment_type == "bemcentinib" ~ "Bemcentinib",
    treatment_type == "tozorakimab" ~ "Tozorakimab",
    treatment_type == "zilucoplan" ~ "Zilucoplan",
    TRUE ~ NA_character_
  )) %>%
  # Step 5: Drop the treatment_type and has_treatment columns
  select(-treatment_type, -has_treatment)


# Assuming final_data_treatment is your dataframe
final_data_with_overlap_type <- final_data_treatment %>%
  mutate(
    # Create a new column 'overlap_type' based on whether 'shared' or 'unique' is in the overlap column
    overlap_type = case_when(
      str_detect(overlap, "shared") ~ "shared",
      str_detect(overlap, "unique") ~ "unique",
      TRUE ~ NA_character_  # If neither, assign NA or handle as you wish
    )
  )

#terms may be shared between two or three sections - only want to keep one instance of this in the plot
# Ensure that each combination of treatment and Description is unique
final_data_unique <- final_data_with_overlap_type %>%
  distinct(Description, treatment, overlap_type, .keep_all = TRUE)


final_data_with_overlap_type_up <- final_data_unique %>%
  filter(regulation == "upregulated")
```

Plot bubble plot for up and down, combine, extract shared legend
```{r fig.width=16, fig.height=18}
# Plot Bubble Plot
up<-ggplot(final_data_with_overlap_type_up, aes(
  x = treatment, 
  y = factor(Description, levels = rev(sort(unique(Description)))),  # Reverse the order of ids
  size = Count, 
  color = regulation
)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(1, 5), limits = c(0,100)) + # Adjust bubble size
  scale_color_manual(values = c("downregulated" = "navy", "upregulated" = "darkred")) +
  theme_bw() +
  facet_grid(~ overlap_type, scales = "free_y") + # Facet by overlap
  labs(
    title = "Upregulated",
    x = "Overlap Category",
    y = "GO Term",
    size = "Gene Count",
    color = "Regulation"
  ) +
  theme(
    axis.title =element_blank(),
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)) 

up_legend<-up +guides(color="none")

final_data_with_overlap_type_down <- final_data_unique %>%
  filter(regulation == "downregulated")


# Plot Bubble Plot
down<-ggplot(final_data_with_overlap_type_down, aes(
  x = treatment, 
  y = factor(Description, levels = rev(sort(unique(Description)))),  # Reverse the order of ids
  size = Count, 
  color = regulation
)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(1, 5), limits = c(0,100)) + # Adjust bubble size
  scale_color_manual(values = c("downregulated" = "navy", "upregulated" = "darkred")) +
  theme_bw() +
  facet_grid(~ overlap_type, scales = "free_y") + # Facet by overlap
  labs(
    title = "Downregulated",
    x = "Overlap Category",
    y = "GO Term",
    size = "Gene Count",
    color = "Regulation"
  ) +
  theme(
    axis.title =element_blank(),
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
    
  )
down_legend<-down+guides(color="none")

#shared legend

final_plot <- plot_grid(
  up_legend + theme(legend.position = "none"),  # Remove legend from the first plot
  down_legend + theme(legend.position = "none"),  # Remove legend from the second plot
  ncol = 2, align = "v", rel_heights = c(1, 1)
)

# Extract the legend from one of the plots
combined_legend <- get_legend(up_legend)
# Combine the plots and legend with controlled spacing
final_plot_with_legend <- plot_grid(
  final_plot, combined_legend, 
  ncol = 2,  # Place the legend next to the plots
  rel_widths = c(0.85, 0.05)  # Adjust relative widths
)
# Display the final plot
final_plot_with_legend


```
