---
  title: "05_Coexpression_analysis_visualisation"
#output: html_notebook
---
Co-expression analysis was performed using pyWGCNA in previous Jupyter notebooks:
04_coexpression_pywgcna.ipynb

# About
This code is for further analysis/intepretation and visualisation of gene co-expression modules

1.  Analysis
(a) Module enrichment analysis using GO terms
(b) Module enrichment analysis using BTM gene sets
(c) Cluster enrichment analysis using both GO and BTM
(d) Gene signature 



2. Visualisation
(a) Heatmap and barplot visualisation of each module correlation and p value with traits of interest
(b) Facet barplot of significant modules 
(c) Consensus bubble plot of treatment arms vs standard of care
(d) Hub gene extraction for visualisation in cytoscape


## Load libraries

```{r message=FALSE}
# Packages to load
packages <- c(
  "clusterProfiler", "cowplot","fgsea", "ggnewscale","ggplot2", "ggsankey", "grid", "gridExtra", "patchwork","rstatix", "tidyverse")


############################################################################################################
load_libraries <- function(packages) {
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

# Load all packages
load_libraries(packages)

```

## Load data

Accompanying manually curated files for visualistion:
"top_pathways" = top 5 enriched GO:BP terms following REVIGO redundancy filtering for barplot plotting for figure 
"module_enrichment_summary" =  manually curated file summarising the main enrichment for each module into key biological process terms
also contains meta data for each of the treatment arms (bem_meta, toz_meta, zil_meta, soc_meta) to use for plotting

```{r}

load("./data_files/wgcna_visualisation_data.Rdata") 

```

## 1. Enrichment analysis of WGCNA modules

Enrichment was performed using clusterprofiler, to align with enrichment performed on DEGs.
Enrichment was also perfomed using BTM gene sets to ascertain cellular specific enrichment
of modules.

### 1a - Gene Ontology enrichment of WGCNA modules

This will perform enrichment on each module in each treatment arm for the BP, MF and CC Gene Ontology categories, creating a new output folder for each module and storing the output in that folder.

The input folder for each treatment arm, "module_information_**treatment**.csv", where **treatment** is replaced by the respective treatment, was created and saved during the 04_coexpression_pywgcna analysis

```{r}

# Define the function to perform the GO enrichment analysis
perform_go_enrichment_analysis <- function(cohort, input_file, output_dir) {
  # Load the data
  module_list <- read.csv(input_file)
  
  # Rename columns
  module_list <- module_list[1:2]
  colnames(module_list) <- c("ENSG", "colors")
  
  # Pivot the data
  module_list$index <- ave(seq_along(module_list$colors), module_list$colors, FUN = seq_along)
  df_pivoted <- spread(module_list, key = colors, value = ENSG)
  df_pivoted$index <- NULL
  module_list_pivoted_clean <- df_pivoted
  module_list_pivoted_clean[is.na(module_list_pivoted_clean)] <- ""
  write.csv(module_list_pivoted_clean, paste0(output_dir, "/Module_information_", cohort, "_wide.csv"))
  
  # Set background genes
  background_genes <- module_list$ENSG
  
  # GO enrichment analysis
  go_categories <- c("MF", "BP", "CC")
  plot_width <- 8
  plot_height <- 10
  base_dir <- paste(output_dir, cohort, "Module_GO", sep="/")
  if (!file.exists(base_dir)) {
    dir.create(base_dir, recursive = TRUE)
  }
  
  for (column in names(module_list_pivoted_clean)) {
    column_dir <- file.path(base_dir, column)
    if (!file.exists(column_dir)) {
      dir.create(column_dir, recursive = TRUE)
    }
    for (go in go_categories) {
      ego <- tryCatch({
        enrichGO(gene = module_list_pivoted_clean[, column],
                 universe = background_genes,
                 OrgDb = "org.Hs.eg.db",
                 keyType = "ENSEMBL",
                 ont = go,
                 pAdjustMethod = "BH",
                 qvalueCutoff = 0.05,
                 readable = TRUE)
      }, error = function(e) NULL)
      
      if (!is.null(ego) && nrow(ego) > 0) {
        bar_plot <- barplot(ego, showCategory = 20) +
          ggtitle(paste("Top 20 enriched", go, "pathways for", column)) +
          theme_minimal()
        ggsave(filename = file.path(column_dir, paste(column, go, "barplot.pdf", sep = "_")), plot = bar_plot, width = plot_width, height = plot_height)
        
        dot_plot <- dotplot(ego, showCategory = 20) +
          ggtitle(paste("Top 20 enriched", go, "pathways for", column)) +
          theme_minimal()
        ggsave(filename = file.path(column_dir, paste(column, go, "dotplot.pdf", sep = "_")), plot = dot_plot, width = plot_width, height = plot_height)
        
        enriched_terms <- ego@result[, c("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "qvalue", "geneID", "Count")]
        enriched_filename <- file.path(column_dir, paste(column, "_", go, "_enriched_pathways.csv", sep = ""))
        write.csv(enriched_terms, enriched_filename, row.names = FALSE)
      } else {
        message(paste("No enriched terms found for", go, "in", column, ". Skipping..."))
      }
    }
  }
}

# Treatment cohorts and their input files
cohorts <- list(
  Bemcentinib = "module_information_bemcentinib.csv",
  Tozorakimab = "module_information_tozorakimab.csv",
  Zilucoplan = "module_information_zilucoplan.csv",
  SoC = "module_information_soc.csv"
)

# Run the analysis for each cohort
for (cohort in names(cohorts)) {
  input_file <- cohorts[[cohort]]
  output_dir <- paste("./", sep="")
  perform_go_enrichment_analysis(cohort, input_file, output_dir)
}



```

### 1b - BTM enrichment of WGCNA modules

As above for GO, the analysis was performed using the BTM gene sets, again creating indiviudal folders for each module to store the output. 

```{r}

# Function to perform BTM enrichment analysis
perform_BTM_enrichment_analysis <- function(cohort, input_file, output_dir) {
  # Load module list
  module_list <- read.csv(input_file)
  
  # Rename columns
  module_list <- module_list[, c("gene_name", "dynamicColors")]
  colnames(module_list) <- c("gene", "colors")
  
  # Pivot data
  module_list$index <- ave(seq_along(module_list$colors), module_list$colors, FUN = seq_along)
  df_pivoted <- spread(module_list, key = colors, value = gene)
  df_pivoted$index <- NULL
  module_list_pivoted_clean <- df_pivoted
  module_list_pivoted_clean[is.na(module_list_pivoted_clean)] <- ""
  
  # Set universe/background genes as the input genes for this WGCNA
  background_genes <- module_list$gene
  
  # Load BTM pathways
  pathways <- read.gmt("BTM_for_GSEA_20131008.gmt")
  
  
  # Create output dir
  plot_height <- 10
  plot_width <-10
  base_dir <- file.path(output_dir, paste(cohort, "/BTM_enrichment", sep="/"))
  if (!file.exists(base_dir)) {
    dir.create(base_dir, recursive = TRUE)
  }
  
  # Loop through columns
  for (column in names(module_list_pivoted_clean)) {
    column_genes <- module_list_pivoted_clean[, column]
    # Remove empty strings from column_genes vector
    column_genes <- column_genes[nzchar(column_genes)]
    
    # Perform enrichment for that module
    enricher_BTM <- enricher(gene = column_genes, universe = background_genes, TERM2GENE = pathways, pAdjustMethod = "BH")
    # Check if enrichment results exist
    if (!is.null(enricher_BTM)) {
      # Check if enrichment results have rows
      if (nrow(enricher_BTM@result) > 0) {
        column_dir <- file.path(base_dir, column)
        if (!file.exists(column_dir)) {
          dir.create(column_dir, recursive = TRUE)
        }
        if (!is.null(enricher_BTM$geneID) && nrow(enricher_BTM) > 0) {
          # Barplot
          bar_plot <- barplot(enricher_BTM, showCategory = 20) +
            ggtitle(paste("Top 20 enriched pathways for", column)) +
            theme_minimal()
          ggsave(filename = file.path(column_dir, paste(column, "barplot.pdf", sep = "_")), plot = bar_plot, width = plot_width, height = plot_height)
          
          # Dotplot
          dot_plot <- dotplot(enricher_BTM, showCategory = 20) +
            ggtitle(paste("Top 20 enriched pathways for", column)) +
            theme_minimal()
          ggsave(filename = file.path(column_dir, paste(column, "dotplot.pdf", sep = "_")), plot = dot_plot, width = plot_width, height = plot_height)
        } else {
          cat(paste("No enriched terms found for", column, "in the BTM gene set. Skipping..."))
        }
        # Save enriched terms to a CSV file
        enriched_terms <- enricher_BTM@result[, c("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "qvalue", "geneID", "Count")]
        enriched_filename <- file.path(column_dir, paste(column, "_enriched_pathways.csv", sep = ""))
        write.csv(enriched_terms, enriched_filename, row.names = FALSE)
      } else {
        message(paste("No enriched terms found for", column, ". Skipping..."))
      }
    } else {
      cat("No enriched terms found for", column, "in the BTM gene set. Skipping...\n")
    }
  }
}

# Define the treatment cohorts and their input files
cohorts <- list(
  Bemcentinib = "module_information_bemcentinib.csv",
  Tozorakimab = "module_information_tozorakimab.csv",
  Zilucoplan = "module_information_zilucoplan.csv",
  SoC = "module_information_soc.csv"
)


# Run the analysis for each cohort
for (cohort in names(cohorts)) {
  input_file <- cohorts[[cohort]]
  output_dir <- paste("./", sep="")
  perform_BTM_enrichment_analysis(cohort, input_file, output_dir)
}


```

Now summarise the cell-specific significantly enriched BTM for each treatment arm module
to see if certain modules have cell specific enrichment associated with them

```{r message=FALSE}

# Define cell-related keywords and their final versions
cell_keywords <- c("monocyte", "monocytes", "Monocyte", "B cells", "B cell", "T cells", "T cell", 
                   "thymocyte", "thymocytes", "Thymocyte", "Thymocytes", "NK cells", "NK cell",
                   "Neutrophils", "Neutrophil", "neutrophils", "neutrophil", "plasma cell", 
                   "lymphocyte", "DC", "DCs", "dendritic cell", "dendritic cells", "myeloid", 
                   "platelet", "Plasma cell", "Plasma cells", "plasma cells")

# Finalized cell type names
final_cell_types <- c("Monocytes", "Monocytes", "Monocytes", "B cells", "B cells", "T cells", "T cells",
                      "T cells", "T cells", "T cells", "T cells", "NK cells", "NK cells",
                      "Neutrophils", "Neutrophils", "Neutrophils", "Neutrophils", "B cells",
                      "B cells", "Dendritic cells", "Dendritic cells", "Dendritic cells", "Dendritic cells", 
                      "Myeloid", "Platelets", "B cells", "B cells", "B cells")

# Create a named vector for replacing keywords
replacement_map <- setNames(final_cell_types, cell_keywords)

# Function to process a single CSV file and count cell-related keywords
process_csv <- function(filepath, replacement_map) {
  # Read the CSV file
  df <- read_csv(filepath)
  
  # Filter significant rows (qvalue < 0.05)
  significant_df <- df %>% filter(qvalue < 0.05)
  
  # Initialize a count vector
  count_vector <- setNames(rep(0, length(unique(final_cell_types))), unique(final_cell_types))
  
  if (nrow(significant_df) > 0) {
    # Replace keywords in the Description column
    for (keyword in names(replacement_map)) {
      replacement <- replacement_map[[keyword]]
      significant_df$Description <- gsub(keyword, replacement, significant_df$Description, ignore.case = TRUE)
    }
    
    # Count occurrences of each final cell type in the Description column
    for (cell_type in unique(final_cell_types)) {
      count_vector[cell_type] <- sum(grepl(cell_type, significant_df$Description, ignore.case = TRUE))
    }
  }
  
  # Create a summarized data frame
  summarized_df <- data.frame(
    CellType = names(count_vector),
    TotalCount = count_vector,
    stringsAsFactors = FALSE
  ) %>%
    pivot_wider(names_from = CellType, values_from = TotalCount, values_fill = list(TotalCount = 0))
  
  return(summarized_df)
}

# List of treatment arms and their respective directories
treatment_arms <- list(
  Bemcentinib = "./Bemcentinib/BTM_enrichment/",
  Tozorakimab = "./Tozorakimab/BTM_enrichment/",
  SoC = "./SoC/BTM_enrichment/",
  Zilucoplan = "./Zilucoplan/BTM_enrichment/"
)

# Initialize an empty list to store the summary data frames
summary_list <- list()

# Loop through each treatment arm and its directory
for (treatment_arm in names(treatment_arms)) {
  dir_path <- treatment_arms[[treatment_arm]]
  
  # Get a list of module dirs
  module_dirs <- list.dirs(dir_path, full.names = TRUE, recursive = FALSE)
  
  # Loop through each module directory
  for (module_dir in module_dirs) {
    module_name <- basename(module_dir)
    csv_file <- file.path(module_dir, paste0(module_name, "_enriched_pathways.csv"))
    
    # Check if the csv file exists
    if (file.exists(csv_file)) {
      # Process the csv file and get the summarised counts
      summarized_counts <- process_csv(csv_file, replacement_map)
      
      # Add the treatment arm and module information
      summarized_counts <- summarized_counts %>%
        mutate(TreatmentArm = treatment_arm, Module = module_name) %>%
        select(TreatmentArm, Module, everything())
      
      # Add the summarized counts to the list
      summary_list[[length(summary_list) + 1]] <- summarized_counts
    } else {
      # If the file does not exist, create a dummy row with zero counts
      dummy_counts <- data.frame(
        TreatmentArm = treatment_arm,
        Module = module_name,
        matrix(0, ncol = length(unique(final_cell_types)), nrow = 1)
      )
      colnames(dummy_counts)[3:ncol(dummy_counts)] <- unique(final_cell_types)
      
      # Add the dummy counts to the list
      summary_list[[length(summary_list) + 1]] <- dummy_counts
    }
  }
}

# Combine all summary data frames into one
summary_df <- bind_rows(summary_list)

# Print the final summary data frame
print(summary_df)

```

### 1c - Enrichment of WGCNA consensus clusters

```{r}
#read in module_information csvs which were saved for each drug treatment arm 
#add drug name to start of module column before combining
Bem_modules<-read.csv("module_information_bemcentinib.csv", row.names=1, header=TRUE)
colnames(Bem_modules)[2] <- paste("Bemcentinib", colnames(Bem_modules)[2], sep = "_") #add treatment infront of colour to pinpoint which module colour treatment when merging
Toz_modules<-read.csv("module_information_tozorakimab.csv", row.names=1, header=TRUE)
colnames(Toz_modules)[2] <- paste("Tozorakimab", colnames(Toz_modules)[2], sep = "_")
Zil_modules<-read.csv("module_information_zilucoplan.csv",row.names=1, header=TRUE)
colnames(Zil_modules)[2] <- paste("Zilucoplan", colnames(Zil_modules)[2], sep = "_")

# Create a subsetted new data frame with just the desired column and the row names
Bem_modules_sub <- Bem_modules[, c("Bemcentinib_moduleColors"), drop = FALSE]
Toz_modules_sub <- Toz_modules[, c("Tozorakimab_moduleColors"), drop = FALSE]
Zil_modules_sub <- Zil_modules[, c("Zilucoplan_moduleColors"), drop = FALSE]

#create background list of all 3 
# Convert row names to a column and rename the column to 'Genes'
Bem_modules_genes <- Bem_modules_sub %>% rownames_to_column(var = "Genes")
Toz_modules_genes <- Toz_modules_sub %>% rownames_to_column(var = "Genes")
Zil_modules_genes <- Zil_modules_sub %>% rownames_to_column(var = "Genes")
# Merge the data frames
background_genes <- full_join(Bem_modules_genes, Toz_modules_genes, by = "Genes") %>%
  full_join(Zil_modules_genes, by = "Genes")
# Select only the 'Genes' column
background_genes <- dplyr::select(background_genes, Genes)
# Drop duplicates 
background_genes <- distinct(background_genes)
# Count the number of unique values in the 'Genes' column
num_unique_genes <- background_genes %>%
  summarize(unique_genes = n_distinct(Genes))
# Print the result
print(num_unique_genes)
#unique_genes
#1        30141
#this list is for background for clusterprofiler analysis
background <- background_genes$Genes

####### Cluster 1 #########
#create gene lists, by defining which modules i want included
#cluster 1 = Bem (gainsboro), Toz (salmon), Zil (brown)
# Specify the values to match
cluster1_Bem_desired_colors <- c("gainsboro")
cluster1_Toz_desired_colors <- c("salmon")
cluster1_Zil_desired_colors <- c("brown")
# Create a logical vector indicating whether each row matches the desired colors
cluster1_Bem_matching_rows <- Bem_modules_sub$Bemcentinib_moduleColors %in% cluster1_Bem_desired_colors
cluster1_Toz_matching_rows <- Toz_modules_sub$Toz_moduleColors %in% cluster1_Toz_desired_colors
cluster1_Zil_matching_rows <- Zil_modules_sub$Zilucoplan_moduleColors %in% cluster1_Zil_desired_colors

# Extract the row names where the condition is TRUE
cluster1_Bem_matching_row_names <- rownames(Bem_modules_sub[cluster1_Bem_matching_rows, , drop = FALSE])
cluster1_Toz_matching_row_names <- rownames(Toz_modules_sub[cluster1_Toz_matching_rows, , drop = FALSE])
cluster1_Zil_matching_row_names <- rownames(Zil_modules_sub[cluster1_Zil_matching_rows, , drop = FALSE])
# Create a new data frame with a single column of matching row names
immune_cluster_1_ids <- data.frame(immune_cluster_1_ids = c(cluster1_Bem_matching_row_names,cluster1_Toz_matching_row_names, cluster1_Zil_matching_row_names))
dim(immune_cluster_1_ids)


#cluster 2
#create gene lists, by defining which modules i want included
#cluster 1 = Bem (darksalmon), Toz (chocolate), Zil (black)
# Specify the values to match
cluster2_Bem_desired_colors <- c("linen", "tomato")
cluster2_Toz_desired_colors <- c("tan")
cluster2_Zil_desired_colors <- c("coral")
# Create a logical vector indicating whether each row matches the desired colors
cluster2_Bem_matching_rows <- Bem_modules_sub$Bemcentinib_moduleColors %in% cluster2_Bem_desired_colors
cluster2_Toz_matching_rows <- Toz_modules_sub$Toz_moduleColors %in% cluster2_Toz_desired_colors
cluster2_Zil_matching_rows <- Zil_modules_sub$Zilucoplan_moduleColors %in% cluster2_Zil_desired_colors

# Extract the row names where the condition is TRUE
cluster2_Bem_matching_row_names <- rownames(Bem_modules_sub[cluster2_Bem_matching_rows, , drop = FALSE])
cluster2_Toz_matching_row_names <- rownames(Toz_modules_sub[cluster2_Toz_matching_rows, , drop = FALSE])
cluster2_Zil_matching_row_names <- rownames(Zil_modules_sub[cluster2_Zil_matching_rows, , drop = FALSE])
# Create a new data frame with a single column of matching row names
immune_cluster_2_ids <- data.frame(immune_cluster_2_ids = c(cluster2_Bem_matching_row_names,cluster2_Toz_matching_row_names, cluster2_Zil_matching_row_names))
dim(immune_cluster_2_ids)

#cluster 3
#create gene lists, by defining which modules i want included
#cluster 1 = Bem (darksalmon), Toz (chocolate), Zil (black)
# Specify the values to match
cluster3_Bem_desired_colors <- c("rosybrown")
cluster3_Toz_desired_colors <- c("orangered", "burlywood")
cluster3_Zil_desired_colors <- c("indianred")
# Create a logical vector indicating whether each row matches the desired colors
cluster3_Bem_matching_rows <- Bem_modules_sub$Bemcentinib_moduleColors %in% cluster3_Bem_desired_colors
cluster3_Toz_matching_rows <- Toz_modules_sub$Toz_moduleColors %in% cluster3_Toz_desired_colors
cluster3_Zil_matching_rows <- Zil_modules_sub$Zilucoplan_moduleColors %in% cluster3_Zil_desired_colors

# Extract the row names where the condition is TRUE
cluster3_Bem_matching_row_names <- rownames(Bem_modules_sub[cluster3_Bem_matching_rows, , drop = FALSE])
cluster3_Toz_matching_row_names <- rownames(Toz_modules_sub[cluster3_Toz_matching_rows, , drop = FALSE])
cluster3_Zil_matching_row_names <- rownames(Zil_modules_sub[cluster3_Zil_matching_rows, , drop = FALSE])
# Create a new data frame with a single column of matching row names
immune_cluster_3_ids <- data.frame(immune_cluster_3_ids = c(cluster3_Bem_matching_row_names,cluster3_Toz_matching_row_names, cluster3_Zil_matching_row_names))
dim(immune_cluster_3_ids)

#cluster 4
#create gene lists, by defining which modules i want included
#cluster 1 = Bem (darksalmon), Toz (chocolate), Zil (black)
# Specify the values to match
cluster4_Bem_desired_colors <- c("snow", "white")
cluster4_Toz_desired_colors <- c("sandybrown")
cluster4_Zil_desired_colors <- c("gainsboro")
# Create a logical vector indicating whether each row matches the desired colors
cluster4_Bem_matching_rows <- Bem_modules_sub$Bemcentinib_moduleColors %in% cluster4_Bem_desired_colors
cluster4_Toz_matching_rows <- Toz_modules_sub$Toz_moduleColors %in% cluster4_Toz_desired_colors
cluster4_Zil_matching_rows <- Zil_modules_sub$Zilucoplan_moduleColors %in% cluster4_Zil_desired_colors

# Extract the row names where the condition is TRUE
cluster4_Bem_matching_row_names <- rownames(Bem_modules_sub[cluster4_Bem_matching_rows, , drop = FALSE])
cluster4_Toz_matching_row_names <- rownames(Toz_modules_sub[cluster4_Toz_matching_rows, , drop = FALSE])
cluster4_Zil_matching_row_names <- rownames(Zil_modules_sub[cluster4_Zil_matching_rows, , drop = FALSE])
# Create a new data frame with a single column of matching row names
immune_cluster_4_ids <- data.frame(immune_cluster_4_ids = c(cluster4_Bem_matching_row_names,cluster4_Toz_matching_row_names, cluster4_Zil_matching_row_names))
dim(immune_cluster_4_ids)

#cluster 5
#create gene lists, by defining which modules i want included
#cluster 1 = Bem (darksalmon), Toz (chocolate), Zil (black)
# Specify the values to match
cluster5_Bem_desired_colors <- c("darksalmon")
cluster5_Toz_desired_colors <- c("chocolate")
cluster5_Zil_desired_colors <- c("black")
# Create a logical vector indicating whether each row matches the desired colors
cluster5_Bem_matching_rows <- Bem_modules_sub$Bemcentinib_moduleColors %in% cluster5_Bem_desired_colors
cluster5_Toz_matching_rows <- Toz_modules_sub$Toz_moduleColors %in% cluster5_Toz_desired_colors
cluster5_Zil_matching_rows <- Zil_modules_sub$Zilucoplan_moduleColors %in% cluster5_Zil_desired_colors

# Extract the row names where the condition is TRUE
cluster5_Bem_matching_row_names <- rownames(Bem_modules_sub[cluster5_Bem_matching_rows, , drop = FALSE])
cluster5_Toz_matching_row_names <- rownames(Toz_modules_sub[cluster5_Toz_matching_rows, , drop = FALSE])
cluster5_Zil_matching_row_names <- rownames(Zil_modules_sub[cluster5_Zil_matching_rows, , drop = FALSE])
# Create a new data frame with a single column of matching row names
immune_cluster_5_ids <- data.frame(immune_cluster_5_ids = c(cluster5_Bem_matching_row_names,cluster5_Toz_matching_row_names, cluster5_Zil_matching_row_names))
dim(immune_cluster_5_ids)

#cluster 6
#create gene lists, by defining which modules i want included
#cluster 1 = Bem (darksalmon), Toz (chocolate), Zil (black)
# Specify the values to match
cluster6_Bem_desired_colors <- c("darkgrey", "silver", "lightcoral")
cluster6_Toz_desired_colors <- c("moccasin", "gainsboro")
cluster6_Zil_desired_colors <- c("darkgrey", "lightgrey", "chocolate")
# Create a logical vector indicating whether each row matches the desired colors
cluster6_Bem_matching_rows <- Bem_modules_sub$Bemcentinib_moduleColors %in% cluster6_Bem_desired_colors
cluster6_Toz_matching_rows <- Toz_modules_sub$Toz_moduleColors %in% cluster6_Toz_desired_colors
cluster6_Zil_matching_rows <- Zil_modules_sub$Zilucoplan_moduleColors %in% cluster6_Zil_desired_colors

# Extract the row names where the condition is TRUE
cluster6_Bem_matching_row_names <- rownames(Bem_modules_sub[cluster6_Bem_matching_rows, , drop = FALSE])
cluster6_Toz_matching_row_names <- rownames(Toz_modules_sub[cluster6_Toz_matching_rows, , drop = FALSE])
cluster6_Zil_matching_row_names <- rownames(Zil_modules_sub[cluster6_Zil_matching_rows, , drop = FALSE])
# Create a new data frame with a single column of matching row names
immune_cluster_6_ids <- data.frame(immune_cluster_6_ids = c(cluster6_Bem_matching_row_names,cluster6_Toz_matching_row_names, cluster6_Zil_matching_row_names))
dim(immune_cluster_6_ids)

#Define set of genes
cluster_1_set <- immune_cluster_1_ids$immune_cluster_1_ids
cluster_2_set <- immune_cluster_2_ids$immune_cluster_2_ids
cluster_3_set <- immune_cluster_3_ids$immune_cluster_3_ids
cluster_4_set <- immune_cluster_4_ids$immune_cluster_4_ids
cluster_5_set <- immune_cluster_5_ids$immune_cluster_5_ids
cluster_6_set <- immune_cluster_6_ids$immune_cluster_6_ids

# Update gene_sets with the new gene sets
gene_sets <- list(
  Cluster_1 = cluster_1_set,
  Cluster_2 = cluster_2_set,
  Cluster_3 = cluster_3_set,
  Cluster_4 = cluster_4_set,
Cluster_5 = cluster_5_set,
Cluster_6 = cluster_6_set
)

# Set up plot parameters
plot_width <- 8
plot_height <- 10
#define root.name
root.name <- "WGCNA_consensus_module_enrichment_"
# List of GO categories
go_categories <- c("MF", "BP", "CC")
# Create the directory if it doesn't exist
dir.create("./Consensus/Clusters/GO/", recursive = TRUE)

for (gene_set_name in names(gene_sets)) {
  gene_set <- gene_sets[[gene_set_name]]
  
  for (go in go_categories) {
    # Perform GO enrichment analysis for the current gene set
    ego <- tryCatch({
      enrichGO(gene = gene_set,
               universe = background,
               OrgDb = "org.Hs.eg.db",
               keyType = "ENSEMBL",
               ont = go,
               pAdjustMethod = "BH",
               qvalueCutoff = 0.05,
               readable = TRUE)
    }, error = function(e) NULL)
    
    if (!is.null(ego$geneID) && nrow(ego) > 0) {
      # Barplot for top 20 enriched pathways for the current gene set
      bar_plot <- barplot(ego, showCategory = 20) +
        ggtitle(paste("Top 20 enriched", go, "pathways for", gene_set_name, "_set")) +
        theme_minimal()
      
      # Save plot as PDF
      ggsave(filename = paste("./Consensus/Clusters/GO/",root.name, gene_set_name, "_", go, "_barplot.pdf", sep = ""), plot = bar_plot, width = plot_width, height = plot_height)
    } else {
      cat(paste("No enriched terms found for", go, "in the", gene_set_name, "_set. Skipping..."))
    }
    
    # Save enrichment results as CSV
    if (!is.null(ego$geneID) && nrow(ego) > 0) {
      write.csv(ego, file = paste("./Consensus/Clusters/GO/",root.name, gene_set_name, "_", go, "_enrichment_results.csv", sep = ""), row.names = FALSE)
    } else {
      cat(paste("No enriched terms found for", go, "in the", gene_set_name, "_set. Skipping..."))
    }
  }
}

```

#### (i) REVIGO
Some of these Biological Process outputs are large. 
The online tool REVIGO was used to reduce the resulting list, where the top 5 terms from each were selected for visualisation. 
This output is stored in the wgcna_visualisation_data.Rdata file.
This REVIGO output was then used to plot the top 5 terms for each cluster as barplot. 

```{r, fig.height=15, fig.width=7.5}

                      
# Define colors for each cluster
cluster_colors <- c("Cluster1" = "skyblue", 
                    "Cluster2" = "red", 
                    "Cluster3" = "forestgreen", 
                    "Cluster4" = "darkorange", 
                    "Cluster5" = "gold", 
                    "Cluster6" = "purple")

ggplot(top_pathways, aes(x=Count, y=reorder(Name, -qvalue), fill=cluster)) +
  geom_bar(stat='identity', color="black", alpha=0.3) +
  theme_bw(base_size = 14) +
  scale_fill_manual(values = cluster_colors) +  
  theme(axis.title.y = element_blank()) + 
  facet_wrap(~cluster, scales = "free", ncol = 1) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        legend.position = "none") +
  ggh4x::facetted_pos_scales(x = list(     
    Cluster1 = scale_x_continuous(expand = c(0, 0), limits = c(0,300)),
    Cluster2 = scale_x_continuous(expand = c(0, 0), limits = c(0,50)),
    Cluster3 = scale_x_continuous(expand = c(0, 0), limits = c(0,100)),
    Cluster4 = scale_x_continuous(expand = c(0, 0), limits = c(0,200)),
    Cluster5 = scale_x_continuous(expand = c(0, 0), limits = c(0,100)),
    Cluster6 = scale_x_continuous(expand = c(0, 0), limits = c(0,400))
  ))


```

### 1d - Gene signatures
Visualisation of gene individual gene signatures according to eigengene expression of each module
Module eigenegene expression for each sample was extracted from the pyWGCNA object and saved as csv for processing in R. 

#### (i) Determine percentage of patients with specific gene signature
```{r}
# General function to process a dataset
process_dataset <- function(eigengene_file, meta_file, treatment_name, patient_count) {
  # Read the data
  eigengene <- read.csv(eigengene_file, header = TRUE)
  meta <- meta_file
  
  # Filter metadata
  id <- eigengene$sample_name
  meta_sub <- meta %>%
    select(Sample_id, Unique_ID, Day, ICU_HDU_stay_during_hospital_admission, 
           Death_or_Respiratory_Failure_at_D29, Sustained_Clinical_Response_by_D29) %>%
    filter(Sample_id %in% id) %>%
    mutate(outcome = case_when(
      ICU_HDU_stay_during_hospital_admission == "Yes" | 
        Death_or_Respiratory_Failure_at_D29 == "Yes" | 
        Sustained_Clinical_Response_by_D29 == "No" ~ "poor", 
      ICU_HDU_stay_during_hospital_admission == "No" & 
        Death_or_Respiratory_Failure_at_D29 == "No" & 
        Sustained_Clinical_Response_by_D29 == "Yes" ~ "good",
      TRUE ~ "unknown"
    ))
  
  # Convert eigengene expression to high/low
  eigengene_exp <- eigengene
  eigengene_exp[, -1] <- apply(eigengene_exp[, -1], 2, function(x) {
    ifelse(x > 0, "high", "low")
  })
  
  # Merge metadata with eigengene expression
  eigengene_meta <- merge(meta_sub, eigengene_exp, by.x = "Sample_id", by.y = "sample_name")
  
  # Pivot data for expression count
  eigengene_exp_count <- eigengene_meta %>%
    pivot_longer(cols = ends_with("eigengeneExp"), 
                 names_to = "module", 
                 values_to = "expression") %>%
    group_by(Day, outcome, module, expression) %>%
    summarize(count = n(), .groups = 'drop') %>%
    complete(Day, outcome, module, expression, fill = list(count = 0))
  
  # Calculate proportion and percentage
  eigengene_exp_count_prop <- eigengene_exp_count %>%
    mutate(
      proportion = count / patient_count,
      percentage = (count / patient_count) * 100
    )
  
  # Reshape to wide format for plotting
  eigengene_exp_count_prop_wide <- eigengene_exp_count_prop %>%
    unite("Day_outcome_expression", Day, outcome, expression, sep = "_") %>%
    pivot_wider(names_from = Day_outcome_expression, 
                values_from = c(count, proportion, percentage),
                values_fn = list(count = sum, proportion = sum, percentage = sum)) %>%
    mutate(module = str_replace(module, "eigengeneExp", treatment_name))
  
  # Matched analysis for gene signature shifts
  shift_df <- eigengene_meta %>%
    filter(Day %in% c("D1", "D5")) %>%
    select(Unique_ID, Day, outcome, ends_with("eigengeneExp")) %>%
    pivot_longer(cols = ends_with("eigengeneExp"), names_to = "module", values_to = "expression") %>%
    pivot_wider(names_from = Day, values_from = expression) %>%
    mutate(shift = case_when(
      D1 == "low" & D5 == "high" ~ "Low to High",
      D1 == "high" & D5 == "low" ~ "High to Low",
      D1 == "low" & D5 == "low" ~ "Always Low",
      D1 == "high" & D5 == "high" ~ "Always High",
      TRUE ~ NA_character_
    ))
  
  return(list(
    count_prop_wide = eigengene_exp_count_prop_wide,
    shift = shift_df
  ))
}

# Apply the function to each dataset
# the eigengene file is the one extracted from the pyWGCNA object in 04_coexpression_pywgcna
datasets <- list(
  bemcentinib = list(
    eigengene_file = "bemcentinib_all_modules_eigengene_data.csv",
    meta_file = bem_meta,
    treatment_name = "bemcentinib",
    patient_count = 11
  ),
  tozorakimab = list(
    eigengene_file = "tozorakimab_all_modules_eigengene_data.csv",
    meta_file = toz_meta,
    treatment_name = "tozorakimab",
    patient_count = 29
  ),
  zilucoplan = list(
    eigengene_file = "zilucoplan_all_modules_eigengene_data.csv",
    meta_file = zil_meta,
    treatment_name = "zilucoplan",
    patient_count = 11
  )
)

# Initialize an empty list to store count_prop_wide for combining later
count_prop_wide_list <- list()

# Loop through each dataset to process and save results
for (treatment_name in names(datasets)) {
  # Process the dataset
  result <- process_dataset(
    eigengene_file = datasets[[treatment_name]]$eigengene_file,
    meta_file = datasets[[treatment_name]]$meta_file,
    treatment_name = datasets[[treatment_name]]$treatment_name,
    patient_count = datasets[[treatment_name]]$patient_count
  )
  
  # output individual dfs with treatment-specific prefixes in the global environment
  count_prop_wide_name <- paste0(treatment_name, "_count_prop_wide")
  shift_df_name <- paste0(treatment_name, "_shift_df")
  
  assign(count_prop_wide_name, result$count_prop_wide, envir = .GlobalEnv)
  assign(shift_df_name, result$shift, envir = .GlobalEnv)
  
  # Add to the list for combining
  count_prop_wide_list[[treatment_name]] <- result$count_prop_wide
}

# Combine wide count data for all treatments
combined_count_wide <- bind_rows(count_prop_wide_list)

# Save combined data
#write.csv(combined_count_wide, "combined_eigenegene_counts_wide.csv")


#now need to mege modules with similar ontology enrichment for plotting and merge with module enrichment summary file
# Select only the "percentage_" columns
percentage_df <- combined_count_wide %>%
  select(module, starts_with("percentage_"))

# List of row combinations to merge - ones that are closely linked in the cluster network (e.g. in same cluster)
merge_combinations <- list(
  tomato_linen_bemcentinib = c("tomato_bemcentinib", "linen_bemcentinib"),
  gainsboro_moccasin_tozorakimab = c("gainsboro_tozorakimab", "moccasin_tozorakimab"),
  papyawhip_linen_tozorakimab = c("papyawhip_tozorakimab", "linen_tozorakimab")
)

# Iterate through the combinations and calculate averages
merged_rows <- lapply(names(merge_combinations), function(new_row_name) {
  # Filter and compute the average for the specified rows
  percentage_df %>%
    filter(module %in% merge_combinations[[new_row_name]]) %>%
    summarise(across(starts_with("percentage_"), mean, na.rm = TRUE)) %>%
    mutate(module = new_row_name)
}) %>%
  bind_rows() # Combine all the merged rows into a single DataFrame

# Step 4: Remove the original rows and add the new merged rows
percentage_df_final <- percentage_df %>%
  filter(!module %in% unlist(merge_combinations)) %>%
  bind_rows(merged_rows)

# View the resulting DataFrame
head(percentage_df_final)

#now using the manually curated module enricvhment summary file to merge with gene sig file 
# this is saved  wgcna_visualisation_data.Rdata as "module_enrichment_summary_terms"
# merge
percentage_mod_enrichment_summary<-merge(module_enrichment_summary_terms, percentage_df_final, by.x="module_treatment", by.y="module")
percentage_mod_enrichment_summary
#save for later use if necessary
#write.csv(percentage_mod_enrichment_summary, "percentage_mod_enrichment_summary.csv")


```

#### (ii) Day 5 gene signature bubble plot

Now to plot the proportion of patients with a high or low eigengene expression for a given gene signature (as defined by enrichment analysis for a module) at day 5 following treatment to see how the modulation of gene signatures impacts on outcomes. 

```{r fig.height=5, fig.width=14}
# reshape the module enrichment summary data to long for plotting as a bubble/dot plot
bubble_data <- percentage_mod_enrichment_summary %>%
  pivot_longer(
    cols = starts_with("percentage_"), 
    names_to = c("Timepoint", "Outcome"), 
    names_pattern = "percentage_(D[1|5])_(.*)", 
    values_to = "Score"
  ) %>%
  select(treatment, module, module_treatment, term, Outcome, Timepoint, Score)


# onlyplotting day5, so keep day 5 rows only
D5_expression_data <- filter(bubble_data, Timepoint %in% c("D5"))
#D5_expression_data <- filter(filtered_data, Timepoint %in% c("D5"))
D5_expression_data <- D5_expression_data %>%
  separate(Outcome, into = c("response", "expression"), sep = "_", remove = FALSE)

# Create a custom labeller fr response facet
response_labeller <- c("good" = "Sustained Clinical\nImprovement",
                       "poor" = "Poor Clinical\nOutcome")
D5_expression_data <- D5_expression_data %>%
  mutate(
    term = ifelse(term == "Immunometabolism", "Immuno- metabolism", term),
    treatment = recode(
      treatment,
      "zilucoplan" = "Zilucoplan",
      "tozorakimab" = "Tozorakimab",
      "bemcentinib" = "Bemcentinib"
    )
  )


colours <- c("Bemcentinib" = "forestgreen", # Forest Green for Bemcentinib
             "Tozorakimab" = "gold",       # Gold for Tozorakimab
             "Zilucoplan" = "coral")

D5_expression_data$expression<-factor(D5_expression_data$expression, levels = c("low", "high"))
D5_expression_data$treatment <-factor(D5_expression_data$treatment, levels = c("Zilucoplan", "Tozorakimab","Bemcentinib" ))

plot_D5_expression <- ggplot(D5_expression_data, aes(x = expression, y = treatment, group = Outcome)) +
  geom_jitter(aes(size = Score, fill = treatment),
              alpha = 0.75,
              color = "black",                # Set the border color to black
              shape = 21,                     # Use shape = 21 for points with fill and border
              position = position_jitterdodge(0.01)) +
  scale_fill_manual(values=colours)+
  # scale_fill_distiller(palette = "BrBG")+
  #  scale_fill_gradient2(low = "purple", mid = "white", high = "orange", midpoint = 0.5) +  # Adjust for fill scale
  scale_size_continuous(
    range = c(1, 15),   # Point size range for visualization
    limits = c(0, 100) # Define the scale limits
    # oob = scales::squish # Handle values outside the range by squishing them into the limits
  ) +
  scale_y_discrete(position = "right")+ #move y labels to the right hand side
  labs(x = "Gene Set Expression", y = "Treatment", fill = "Treatment", size = "Percentage") +
  theme_bw(base_size=12) +
  theme(
    #axis.text.x = element_blank(), # Remove labels for this plot
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),  # Remove y-axis label for term
    #  axis.title.x = element_blank(),  # Remove x-axis label for Timepoint
    legend.position = "bottom")+     # Hide legend for this plot+
  facet_grid(response ~ term, 
             scales = "free_x", 
             switch="y",
             labeller = labeller(response = response_labeller, term = label_wrap_gen(10))) +
  ggtitle("Gene Signature Expression at Day 5")+
  guides(
    fill = guide_legend(override.aes = list(size = 15)),  # Increase dot size for color legend
  )
plot_D5_expression

```

#### (iii) Day 1 to Day 5 sankey plot

To visualise individual flow of patient shift in expression between day 1 and day 5, plot as a sankey plot

Here we are interested in those 4 key immune signatures shown to be important for outcomes throughout the stud:
1. B cell signalling
2. Antiviral immunity
3. Immune cell lactivation and inflammation
4. Platelet and coagulation

first need to process dfs ready for plotting, generate aesthetics etc

```{r}
# Function to process shift dataframes
process_shift_data <- function(shift_df, treatment_name) {
  shift_df %>%
    mutate(
      module = gsub("_eigengeneExp", "", module),  # Remove 'eigengeneExp' from module names
      treatment = treatment_name,  # Add treatment column
      module_treatment = paste(module, treatment_name, sep = "_")  # Create module_treatment column, important as some module colours may overlap between treatments, but they are different!
    )
}

# Process each treatment's shift dataframe
bem_shift <- process_shift_data(bemcentinib_shift_df, "bemcentinib")
toz_shift <- process_shift_data(tozorakimab_shift_df, "tozorakimab")
zil_shift <- process_shift_data(zilucoplan_shift_df, "zilucoplan")

# Combine all shift data into a single dataframe
all_shift <- bind_rows(bem_shift, toz_shift, zil_shift)

# Define modules of interest with corresponding terms
modules_of_interest <- c(
  "tomato_bemcentinib" = "Platelet and coagulation",
  "white_bemcentinib" = "Immune cell activation and inflammation",
  "rosybrown_bemcentinib" = "Antiviral immunity",
  "maroon_bemcentinib" = "B-cell activation and signalling",
  "tan_tozorakimab" = "Platelet and coagulation",
  "burlywood_tozorakimab" = "Antiviral immunity",
  "sandybrown_tozorakimab" = "Immune cell activation and inflammation",
  "lightsalmon_zilucoplan" = "B-cell activation and signalling"
)

# Filter and map the 'term' column based on 'module_treatment'
all_shift_filtered <- all_shift %>%
  filter(module_treatment %in% names(modules_of_interest)) %>%
  mutate(term = modules_of_interest[module_treatment])

# Check the filtered data
head(all_shift_filtered)

# Recode and reorder factors in df_endotypes
df_endotypes <- all_shift_filtered %>%
  mutate(
    treatment = recode(treatment, 
                       "bemcentinib" = "Bemcentinib", 
                       "tozorakimab" = "Tozorakimab", 
                       "zilucoplan" = "Zilucoplan"),
    outcome = recode(outcome, 
                     "poor" = "Poor Clinical\nOutcome", 
                     "good" = "Sustained Clinical\nImprovement"),
    term = recode(term,
                  "Immune cell activation and inflammation" = "Immune cell activation\nand inflammation",
                  "B-cell activation and signalling" = "B-cell activation\nand signalling",
                  "Platelet and coagulation" = "Platelet and\ncoagulation")
  )

# Expression and treatment colors
expression_colours <- c(
  "Standard\nof Care" = "skyblue",
  "Bemcentinib" = "forestgreen",
  "Tozorakimab" = "gold",
  "Zilucoplan" = "coral",
  "low" = "navy",
  "high" = "darkred",
  "always high" = "darkred",
  "high to low" = "aquamarine2",
  "low to high" = "orange",
  "always low" = "navy",
  "Poor Clinical\nOutcome" = "#CC3300",
  "Sustained Clinical\nImprovement" = "#660066",
  "Immune cell activation\nand inflammation" = "#CC3333",
  "Antiviral immunity" = "#339966",
  "B-cell activation\nand signalling" = "#A65628",
  "Platelet and\ncoagulation" = "#F781BF"
)

```

```{r, fig.height=12, fig.width=20}


#Extract each term of interest into new df for plotting

# 1. Bcell signalling
bcell_endotypes_bem<- df_endotypes %>%
  filter(term=="B-cell activation\nand signalling", treatment == "Bemcentinib")
bcell_endotypes_zil<- df_endotypes %>%
  filter(term=="B-cell activation\nand signalling", treatment == "Zilucoplan")

#turn into df for sankey
sankey_bcell_bem <- bcell_endotypes_bem %>%
  make_long(D1,D5, outcome)%>%
  mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")))

#Create plot
plot_sankey_bcell_bem <- ggplot(sankey_bcell_bem, aes(x = x, 
                                                      next_x = next_x, 
                                                      node = node, 
                                                      next_node = next_node,
                                                      fill = factor(node),
                                                      label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  # scale_x_discrete(
  #   limits = c("d1", "d5", "outcome"),  # Current labels
  #   labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")) + #remove x labels for plotting one above each other in final plot
  theme_sankey(base_size = 16) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title =  element_blank(),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank())+     # Remove x-axis text

  # Add green box for bem
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "forestgreen", col = "black", lwd = 2)),  # Larger green box
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf) +  # Adjust the size and position
  # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -4, ymax = 4 ) +  # Adjust position to be just around the text
  # Add the text "Bemcentinib"
  annotation_custom(
    grob = textGrob("Bemcentinib", 
                    rot = 90,  
                    gp = gpar(fontsize = 16, col = "black")),  # Text inside the white box
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)  # Position the text

#Now for zilucoplan bcell
#turn into df for sankey
sankey_bcell_zil <- bcell_endotypes_zil %>%
  make_long(D1,D5, outcome)%>%
  
  mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome", "Sustained Clinical\nImprovement")))

plot_sankey_bcell_zil <- ggplot(sankey_bcell_zil, aes(x = x, 
                                                      next_x = next_x, 
                                                      node = node, 
                                                      next_node = next_node,
                                                      fill = factor(node),
                                                      label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  # scale_x_discrete(
  #   limits = c("d1", "d5", "outcome"),  # Current labels
  #   labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")) + #remove x labels for plotting one above each other in final plot
  theme_sankey(base_size = 16) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title =  element_blank(),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank())+     # Remove x-axis text
  # Add red/coral box for zilucoplan
    annotation_custom(
    grob = rectGrob(gp = gpar(fill = "coral", col = "black", lwd = 2)),  # Box for Zilucoplan
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf) +
  # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -3, ymax = 3 ) +    # Adjust position to be just around the text
 # add zilucoplan as text
  annotation_custom(
    grob = textGrob("Zilucoplan", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)

#combined both plots into a single b cell plot
bcell<-grid.arrange(plot_sankey_bcell_bem, 
                    plot_sankey_bcell_zil,
                    ncol=1,
                    top= textGrob("B-cell activation and signalling",gp=gpar(fontsize=20)),
                    heights = c(4, 4))
#2. antiviral immunity 
#first bemcentinib
antiviral_endotypes_bem<- df_endotypes %>%
  filter(term=="Antiviral immunity", treatment == "Bemcentinib")
antiviral_endotypes_toz<- df_endotypes %>%
  filter(term=="Antiviral immunity", treatment == "Tozorakimab")
#turn into df for sankey
sankey_antiviral_bem <- antiviral_endotypes_bem %>%
  make_long(D1,D5, outcome)%>%
    mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")))
                                           
plot_sankey_antiviral_bem <- ggplot(sankey_antiviral_bem, aes(x = x, 
                                                              next_x = next_x, 
                                                              node = node, 
                                                              next_node = next_node,
                                                              fill = factor(node),
                                                              label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  theme_sankey(base_size = 16) +
  # scale_x_discrete(
  #   limits = c("d1", "d5", "outcome"),  # Current labels
  #   labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")) + #remove x labels for plotting one above each other in final plot
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title = element_blank(),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank())+     # Remove x-axis text
 # Add green box for bemcentinib
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "forestgreen", col = "black", lwd = 2)),  # Box around text
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf ) +
   # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -4, ymax = 4  ) +# Adjust position to be just around the text
#Add text for bemcentinib
  annotation_custom(
    grob = textGrob("Bemcentinib", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)
# Now for tozorakimab antiviral
#turn into df for sankey
sankey_antiviral_toz <- antiviral_endotypes_toz %>%
  make_long(D1,D5, outcome)%>%
  
  mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")))

plot_sankey_antiviral_toz <- ggplot(sankey_antiviral_toz, aes(x = x, 
                                                              next_x = next_x, 
                                                              node = node, 
                                                              next_node = next_node,
                                                              fill = factor(node),
                                                              label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  # scale_x_discrete(
  #   limits = c("d1", "d5", "outcome"),  # Current labels
  #   labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")) + #remove x labels for plotting one above each other in final plot
  theme_sankey(base_size = 16) +
   theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title = element_blank(),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank())+    # Remove x-axis text
 # Add yellow/gold box for Tozorakimab
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "gold", col = "black", lwd = 2)),  # Box around text
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf ) +
  # Add a white box around the text "Tozorakimab"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -9, ymax = 9) +  # Adjust position to be just around the text
  # Add text for Tozorakimab
  annotation_custom(
    grob = textGrob("Tozorakimab", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)
 #combined the two antiviral sankey plots
antiviral<-grid.arrange(plot_sankey_antiviral_bem, 
                        plot_sankey_antiviral_toz,
                        ncol=1,
                        top= textGrob("Antiviral immunity",gp=gpar(fontsize=20)),
                        heights = c(4, 4))  # Adjust heights as necessary

#3. immune cell activation and inflammation
inflamm_endotypes_bem<- df_endotypes %>%
  filter(term=="Immune cell activation\nand inflammation", treatment == "Bemcentinib")
inflamm_endotypes_toz<- df_endotypes %>%
  filter(term=="Immune cell activation\nand inflammation", treatment == "Tozorakimab")
#First bemcentinib
#turn into df for sankey
sankey_inflamm_bem <- inflamm_endotypes_bem %>%
  make_long(D1,D5, outcome)%>%
    mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome",
                                             "Sustained Clinical\nImprovement")))

plot_sankey_inflamm_bem <- ggplot(sankey_inflamm_bem, aes(x = x, 
                                                          next_x = next_x, 
                                                          node = node, 
                                                          next_node = next_node,
                                                          fill = factor(node),
                                                          label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  # scale_x_discrete(
  #   limits = c("d1", "d5", "outcome"),  # Current labels
  #   labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")) + #remove x labels for plotting one above each other in final plot
  theme_sankey(base_size = 16) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title = element_blank(),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank())+     # Remove x-axis text
 # Add green box for Bem
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "forestgreen", col = "black", lwd = 2)),  # Box around text
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf) +
   # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -3.5, ymax = 3.5) +  # Adjust position to be just around the text
 # Add text for Bemcentinib
  annotation_custom(
    grob = textGrob("Bemcentinib", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)
# Now for Tozorakimab
#turn into df for sankey
sankey_inflamm_toz <- inflamm_endotypes_toz %>%
  make_long(D1,D5, outcome)%>%
    mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome",
                                   "Sustained Clinical\nImprovement")),
    
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome",
                                             "Sustained Clinical\nImprovement"))
  )

plot_sankey_inflamm_toz <- ggplot(sankey_inflamm_toz, aes(x = x, 
                                                          next_x = next_x, 
                                                          node = node, 
                                                          next_node = next_node,
                                                          fill = factor(node),
                                                          label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  scale_x_discrete(
    limits = c("D1", "D5", "outcome"),  # Current labels
    labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")  # New labels
  ) +
  theme_sankey(base_size = 16) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title = element_blank(),
    axis.title.x = element_blank())+  # Remove x-axis title
    # axis.text.x = element_blank()     # Remove x-axis text
  # Add yellow/gold box for Toz
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "gold", col = "black", lwd = 2)),  # Box around text
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf) +
  # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -9, ymax = 9) +  # Adjust position to be just around the text
    # Add texr for Tozorakimab
  annotation_custom(
    grob = textGrob("Tozorakimab", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)
  
# combine both plots for inflamm
inflamm<-grid.arrange(plot_sankey_inflamm_bem, 
                      plot_sankey_inflamm_toz,
                      ncol=1,
                      top= textGrob("Immune cell activation and inflammation",gp=gpar(fontsize=20)),
                      heights = c(4, 4))  # Adjust heights as necessary

# 4.Platelets and coagulation terms
platelet_endotypes_bem<- df_endotypes %>%
  filter(term=="Platelet and\ncoagulation", treatment == "Bemcentinib")
platelet_endotypes_toz<- df_endotypes %>%
  filter(term=="Platelet and\ncoagulation", treatment == "Tozorakimab")
#First for Bem
#turn into df for sankey
sankey_platelet_bem <- platelet_endotypes_bem %>%
  make_long(D1,D5, outcome)%>%
    mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")))
  
plot_sankey_platelet_bem <- ggplot(sankey_platelet_bem, aes(x = x, 
                                                            next_x = next_x, 
                                                            node = node, 
                                                            next_node = next_node,
                                                            fill = factor(node),
                                                            label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  # scale_x_discrete(
  #   limits = c("d1", "d5", "outcome"),  # Current labels
  #   labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")  # New labels
  # ) +
  theme_sankey(base_size = 16) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title = element_blank(),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank() )+     # Remove x-axis text
 # Add green box for Bem
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "forestgreen", col = "black", lwd = 2)),  # Box around text
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf) +
   # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -3.5, ymax = 3.5) +  # Adjust position to be just around the text
 # Add text for Bemcentinib
    annotation_custom(
    grob = textGrob("Bemcentinib", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)
# Now for Tozorakimab
#turn into df for sankey
sankey_platelet_toz <- platelet_endotypes_toz %>%
  make_long(D1,D5, outcome)%>%
    mutate(
    # Reorder levels within 'node' for 'd1', 'd5', 'outcome' to flip high and low
    node = factor(node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")),
    next_node = factor(next_node, levels = c("low", "high",  "Poor Clinical\nOutcome","Sustained Clinical\nImprovement")))
 
plot_sankey_platelet_toz <- ggplot(sankey_platelet_toz, aes(x = x, 
                                                            next_x = next_x, 
                                                            node = node, 
                                                            next_node = next_node,
                                                            fill = factor(node),
                                                            label = node)) +  
  geom_sankey(flow.alpha = 0.5, node.colour = 1, smooth = 6, width = 0.2) +
  geom_sankey_label(
    size = 3.5, color = 1,  fill = "white")+
  scale_fill_manual(values = expression_colours) +
  scale_x_discrete(
    limits = c("D1", "D5", "outcome"),  # Current labels
    labels = c("Day 1 Expression", "Day 5 Expression", "Clinical Outcome")  # New labels
  ) +
  theme_sankey(base_size = 16) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    legend.position = "none",
    plot.title = element_blank(),
    axis.title.x = element_blank()  # Remove x-axis title
    # axis.text.x = element_blank()     # Remove x-axis text
  )+
  #add yellow/gold box for Toz
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "gold", col = "black", lwd = 2)),  # Box around text
    xmin = 0.7, xmax = 0.9, ymin = -Inf, ymax = Inf) +
   # Add a white box around the text "Bemcentinib"
  annotation_custom(
    grob = rectGrob(gp = gpar(fill = "white", col = "black", lwd = 2)),  # White box around the text
    xmin = 0.75, xmax = 0.85, ymin = -9, ymax = 9) +# Adjust position to be just around the text
  # Add text for Tozorakimab
  annotation_custom(
    grob = textGrob("Tozorakimab", 
                    rot = 90,  
                    gp = gpar(fontsize = 16)),
    xmin = 0.8, xmax = 0.8, ymin = -Inf, ymax = Inf)
# Combine both sankey plots
platelet<-grid.arrange(plot_sankey_platelet_bem, 
                       plot_sankey_platelet_toz,
                       ncol=1,
                       top= textGrob("Platelet and coagulation",gp=gpar(fontsize=20)),
                       heights = c(4, 4))  # Adjust heights as necessary

# Combine the four plots into a 2x2 grid
combined_plots <- grid.arrange(
  bcell, 
  antiviral,
  inflamm, 
  platelet,
  ncol = 2,  # 2 columns
  nrow = 2,  # 2 rows
  #top = textGrob("Endotype Analysis", gp = gpar(fontsize = 20))  # Optional title for the combined plot
  padding = unit(50, "cm")  # Reduce the space between plots
)



```

#### (iv) Gene signature shift for poor outcomes
Now we are just selecting those with poor outcome to see their shift in expression categorised for the 4 key immune signatures

```{r, fig.height=8, fig.width=6}
# now plot the poor outcome individuals to show where the treatment failure may be
# unique Patient_group labels for each treatment group
unique_ids <- df_endotypes %>%
  filter(outcome == "Poor Clinical\nOutcome") %>%  # Filter for poor clinical outcomes
  select(Unique_ID, treatment) %>%  # Keep only relevant columns
  distinct() %>%  # Ensure unique combinations of Unique_ID and treatment
  group_by(treatment) %>%  # Group by treatment
  mutate(Patient_group = paste(treatment, row_number(), sep = " #")) %>%  # Assign sequential IDs
  ungroup()
# Merge the Patient_group back into the original dataframe
df_modified <- df_endotypes %>%
  left_join(unique_ids, by = c("Unique_ID", "treatment")) %>%  # Join by Unique_ID and treatment
  filter(outcome == "Poor Clinical\nOutcome")  # Keep only poor clinical outcomes
#Reverse the order of Patient_group for plotting
df_modified$Patient_group <- factor(df_modified$Patient_group, 
                                    levels = rev(unique(df_modified$Patient_group)))
# Create the heatmap plot
ggplot(df_modified, aes(x = term, y = Patient_group, fill = shift)) +
  geom_tile(color = "white", linewidth = 0.5) +  # Create tiles with a border for clarity
  facet_grid(treatment~., scales = "free_y", space="free") +    # Facet by treatment with labels on the right
  scale_fill_manual(values = c("Low to High" = "coral", 
                               "High to Low" = "skyblue", 
                               "Always Low" = "navy", 
                               "Always High" = "darkred")) +
  labs(x = "", 
       y = "Patient", 
       fill = "Expression Shift") +
  scale_x_discrete(expand = c(0,0))+
  
  scale_y_discrete(expand = c(0,0))+
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
        panel.grid = element_blank())+
  ggtitle("Gene signature shifts in\nindividuals with poor outcomes")

```

# 2. Visualisation

## 2a Heatmap and barplot visualisation of each module correlation and p value with traits of interest

Creating module gene barplot showing the  number of genes assigned to a module and module-trait relationship heatmap for each of the treatment groups and then combine as one single plot

For this, will need the correlation and pvalue matrix saved from 04_coexpression_pywgcna, as well as the "module_information"df for each treatment
#### (i) Tozorakimab 
```{r}
# Create the heatmap for Tozorakimab group
toz_heatmap_corr <- read.csv("tozorakimab_module_trait_cor_matrix.csv", row.names = 1, header = TRUE)
toz_heatmap_pval <- read.csv("tozorakimab_module_trait_pvalue_matrix.csv", row.names = 1, header = TRUE)
toz_combo <- merge(toz_heatmap_corr, toz_heatmap_pval, by = "row.names")
# Save this over
toz_module_order <- row.names(toz_heatmap_corr)
# Change colnames
colnames(toz_combo)[colnames(toz_combo) == "Row.names"] <- "Modules"
# Module number for barplot
toz_modules <- read.csv("Module_information_Tozorakimab.csv")
# First, summarize the data by moduleColors
toz_module_number <- toz_modules %>%
  group_by(moduleColors) %>%
  summarize(total_genes = n())

# Create new column by adding 'ME' in front of 'toz_module_number'
toz_module_number <- as.data.frame(toz_module_number)
toz_cols = toz_module_number$moduleColors 
toz_module_number$Modules <- paste0("ME", toz_module_number$moduleColors)

# Set as row names for ordering the same way as the heatmap 
row.names(toz_module_number) <- toz_module_number$Modules

# Use this as the colours instead
toz_heatmap_col_order <- gsub("ME", "", rownames(toz_heatmap_corr))

# Reorder the levels in Modules
toz_module_number$Modules <- factor(toz_module_number$Modules, levels = toz_module_order)
toz_module_number$moduleColors <- factor(toz_module_number$moduleColors, levels = toz_heatmap_col_order)

# Make longer
toz_df_long_corr <- toz_combo %>%
  pivot_longer(
    cols = ends_with(".x"),
    names_to = "Trait",
    values_to = "Corr"
  ) %>%
  mutate(Trait = sub("\\.x$", "", Trait))

toz_df_long_pval <- toz_combo %>%
  pivot_longer(
    cols = ends_with(".y"),
    names_to = "Trait",
    values_to = "pval"
  ) %>%
  mutate(Trait = sub("\\.y$", "", Trait))

# Merge the two data frames based on "Modules" and "Trait"
toz_result_df <- merge(toz_df_long_corr, toz_df_long_pval, by = c("Modules", "Trait"))
# Reorder columns if needed
toz_result_df <- toz_result_df[, c("Modules", "Trait", "Corr", "pval")]
# Convert to numeric or character if it's not already
toz_result_df$Modules <- as.character(toz_result_df$Modules)
toz_result_df$pval <- as.numeric(toz_result_df$pval)
toz_result_df$Corr <- as.numeric(toz_result_df$Corr)

# Print the result
print(toz_result_df)

# Define your custom palette
toz_my_palette <- rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582", "#FDDBC7",
                        "#F7F7F7", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"))

# Reorder the levels in Trait
toz_result_df$Trait <- factor(toz_result_df$Trait, levels = rev(c("Day", "Remdesivir", "ICU", "D29_Death_RespFailure", "D29_Sustained_Clinical_Response")))

# Reorder the levels in Modules
toz_result_df$Modules <- factor(toz_result_df$Modules, levels = toz_module_order)

# Create a colour bar
# Extract colors from Module labels
toz_module_colors <- sub("^ME", "", toz_module_order)
# Convert module_colors to factor with desired order
toz_module_colors_factor <- factor(toz_module_colors, levels = unique(toz_module_colors))

# Define the new labels for traits
toz_new_trait_labels <- c(
  "Day" = "Day",
  "Remdesivir" = "Remdesivir",
  "ICU" = "ICU",
  "D29_Death_RespFailure" = "Death or Respiratory Failure",
  "D29_Sustained_Clinical_Response" = "Sustained Clinical Response"
)

# Create barplot
toz_barplot <- ggplot(data = toz_module_number, aes(x = Modules, y = total_genes, fill = toz_cols)) +
  geom_col(aes(x = Modules, y = total_genes, fill = toz_cols), colour = "black", fill = toz_cols, position = "dodge") +
  geom_text(aes(label = total_genes), position = position_dodge(width = 0.9), vjust = -0.25, size = 2) +
  ylab("Number of genes") + 
  ggtitle("Tozorakimab modules")+
  theme_classic(base_size = 8) +
  scale_y_continuous(limits = c(0, 12000)) +
  scale_fill_manual(values = toz_cols) +
  coord_cartesian(expand = FALSE) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(size = 6, vjust = -15),
    plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())

# Create the heatmap
toz_heatmap <- ggplot(toz_result_df) +
  geom_tile(aes(x = Modules, y = Trait, fill = Corr, width = 1, height = 1)) +
  scale_fill_gradientn(
    colors = toz_my_palette,  # Use module colors
    limits = c(-1, 1), 
    breaks = c(-1, 0, 1),
    labels = scales::number_format(accuracy = 0.1)) +
  geom_text(aes(x = Modules, y = Trait, label = paste(round(Corr, 2))), 
            vjust = 0, hjust = 0.5, size = 2, color = ifelse(toz_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(toz_result_df$pval < 0.05, "bold", "plain")) +
  geom_text(aes(x = Modules, y = Trait, label = paste("(", format(round(pval, 3), nsmall = 3), ")", sep = "")), # format pvalue to 3 decimal places
            vjust = 1, hjust = 0.5, size = 2, color = ifelse(toz_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(toz_result_df$pval < 0.05, "bold", "plain")) +
  theme_minimal(base_size = 9.5) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank(),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.title = element_text(size = 6),  # Adjust legend title size
    plot.title = element_text(size = 10),  # Adjust title text size
    legend.key.width = unit(0.2, "cm"),  # Adjust legend color scale width
    legend.key.height = unit(0.5, "cm")) +  # Adjust legend color scale height
  labs(
    x = "",
    y = "",
    fill = "Correlation") +
  scale_y_discrete(labels = toz_new_trait_labels) +
  theme(legend.position = "right") +
  coord_fixed(ratio = 1) +
  new_scale_fill() +
  geom_tile(data = tibble(Modules = paste0(toz_module_order),
                          d = toz_module_colors_factor),
            aes(x = Modules, y = 0.25, fill = d, height = 0.5)) +
  scale_fill_identity()

# Combine the two plots using patchwork
toz_plot <- toz_barplot / plot_spacer() / toz_heatmap + plot_layout(heights = c(0.25, -0.3, 1))  
# plot_annotation(  title = 'Tozorakimab Modules') & theme(plot.title = element_text(hjust = 0.5))

# Display the combined plot
print(toz_plot)

```
#### (ii) Bemcentinib
```{r}
# Read in data
### for heatmap
bem_heatmap_corr <- read.csv("bemcentinib_module_trait_cor_matrix.csv", row.names = 1, header = TRUE)
bem_heatmap_pval <- read.csv("bemcentinib_module_trait_pvalue_matrix.csv", row.names = 1, header = TRUE)
bem_combo <- merge(bem_heatmap_corr, bem_heatmap_pval, by = "row.names")

# Rename specific modules as silver will not show as a colour, use slategrey (close enough) instead
row.names(bem_heatmap_corr)[row.names(bem_heatmap_corr) == "MEsilver"] <- "MEslategrey"
row.names(bem_heatmap_pval)[row.names(bem_heatmap_pval) == "MEsilver"] <- "MEslategrey"
bem_combo <- merge(bem_heatmap_corr, bem_heatmap_pval, by = "row.names")

# Module order
bem_module_order <- row.names(bem_heatmap_corr)
colnames(bem_combo)[colnames(bem_combo) == "Row.names"] <- "Modules"

# Module number for barplot
bem_modules <- read.csv("Module_information_Bemcentinib.csv")

# First, summarise the data by moduleColors
bem_module_number <- bem_modules %>%
  group_by(moduleColors) %>%
  summarize(total_genes = n())

# Change silver to slategrey
bem_module_number <- as.data.frame(bem_module_number)
bem_module_number$moduleColors[bem_module_number$moduleColors == "silver"] <- "slategrey"
cols = bem_module_number$moduleColors

# Create new column by adding 'ME' in front of 'bem_module_number'
bem_module_number$Modules <- paste0("ME", bem_module_number$moduleColors)

# Set as row names for ordering the same way as the heatmap 
row.names(bem_module_number) <- bem_module_number$Modules

# Remove the string "ME" from the 'module_column'
bem_heatmap_col_order <- gsub("ME", "", row.names(bem_heatmap_corr)) # Use this as the colors instead
bem_heatmap_col_order[bem_heatmap_col_order == "silver"] <- "slategrey"

# Save this order
bem_module_order <- row.names(bem_heatmap_corr)

# Reorder the levels in Modules
bem_module_number$Modules <- factor(bem_module_number$Modules, levels = bem_module_order)
bem_module_number$moduleColors <- factor(bem_module_number$moduleColors, levels = bem_heatmap_col_order)

# Convert combo dataframe to long format for correlations and p-values
bem_df_long_corr <- bem_combo %>%
  pivot_longer(
    cols = ends_with(".x"),
    names_to = "Trait",
    values_to = "Corr"
  ) %>%
  mutate(Trait = sub("\\.x$", "", Trait))

bem_df_long_pval <- bem_combo %>%
  pivot_longer(
    cols = ends_with(".y"),
    names_to = "Trait",
    values_to = "pval"
  ) %>%
  mutate(Trait = sub("\\.y$", "", Trait))

# Merge the two data frames based on "Modules" and "Trait"
bem_result_df <- merge(bem_df_long_corr, bem_df_long_pval, by = c("Modules", "Trait"))

# Reorder columns if needed
bem_result_df <- bem_result_df[, c("Modules", "Trait", "Corr", "pval")]

# Convert pval to numeric if it's not already
bem_result_df$Modules <- as.character(bem_result_df$Modules)
bem_result_df$pval <- as.numeric(bem_result_df$pval)
bem_result_df$Corr <- as.numeric(bem_result_df$Corr)

# Print the result
print(bem_result_df)

# Define your custom palette
bem_my_palette <- rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582", "#FDDBC7",
                        "#F7F7F7", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"))

# Reorder the levels in Trait
bem_result_df$Trait <- factor(bem_result_df$Trait, levels = rev(c("Day", "Remdesivir", "ICU", "D29_Death_RespFailure", "D29_Sustained_Clinical_Response")))
# Reorder the levels in Modules
bem_result_df$Modules <- factor(bem_result_df$Modules, levels = bem_module_order)

# Create a colour bar
# Extract colors from Module labels
bem_module_colors <- sub("^ME", "", bem_module_order)

# Convert module_colors to factor with desired order
bem_module_colors_factor <- factor(bem_module_colors, levels = unique(bem_module_colors))

# Define the new labels for traits
bem_new_trait_labels <- c(
  "Day" = "Day",
  "Remdesivir" = "Remdesivir",
  "ICU" = "ICU",
  "D29_Death_RespFailure" = "Death or Respiratory Failure",
  "D29_Sustained_Clinical_Response" = "Sustained Clinical Response"
)

# Create barplot
bem_barplot <- ggplot(data = bem_module_number, aes(x = Modules, y = total_genes, fill = cols)) +
  geom_col(aes(x = Modules, y = total_genes, fill = cols), colour = "black", position = "dodge") +
  geom_text(aes(label = total_genes), position = position_dodge(width = 0.9), vjust = -0.25, size = 2) +
  ylab("Number of genes") + 
  ggtitle("Bemcentinib modules")+
  theme_classic(base_size = 8) +
  scale_y_continuous(limits = c(0, 8000)) +
  scale_fill_manual(values = cols) +
  coord_cartesian(expand = FALSE) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(size = 6, vjust = -15),
    plot.title = element_text(hjust = 0.5)
  ) +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank())

# Create the heatmap
bem_heatmap <- ggplot(bem_result_df) +
  geom_tile(aes(x = Modules, y = Trait, fill = Corr, width = 1, height = 1)) +
  scale_fill_gradientn(
    colors = bem_my_palette,  # Use module colors
    limits = c(-1, 1), 
    breaks = c(-1, 0, 1),
    labels = scales::number_format(accuracy = 0.1)) +
  geom_text(aes(x = Modules, y = Trait, label = paste(round(Corr, 2))), 
            vjust = 0, hjust = 0.5, size = 2, color = ifelse(bem_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(bem_result_df$pval < 0.05, "bold", "plain")) +
  geom_text(aes(x = Modules, y = Trait, label = paste("(", format(round(pval, 3), nsmall = 3), ")", sep = "")), # format pvalue to 3 decimal places
            vjust = 1, hjust = 0.5, size = 2, color = ifelse(bem_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(bem_result_df$pval < 0.05, "bold", "plain")) +
  theme_minimal(base_size = 9.5) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank(),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.title = element_text(size = 6),  # Adjust legend title size
    plot.title = element_text(size = 10),  # Adjust title text size
    legend.key.width = unit(0.2, "cm"),  # Adjust legend color scale width
    legend.key.height = unit(0.5, "cm")) +  # Adjust legend color scale height
  labs(
    x = "",
    y = "",
    fill = "Correlation") +
  scale_y_discrete(labels = bem_new_trait_labels) +
  theme(legend.position = "right") +
  coord_fixed(ratio = 1) +
  new_scale_fill() +
  geom_tile(data = tibble(Modules = paste0(bem_module_order),
                          d = bem_module_colors_factor),
            aes(x = Modules, y = 0.25, fill = d, height = 0.5)) +
  scale_x_discrete(labels = c("MEslategrey" = "MEsilver", levels(bem_result_df$Modules))) +
  scale_fill_identity()

# Combine the two plots using patchwork
bem_plot <- bem_barplot / plot_spacer() / bem_heatmap + plot_layout(heights = c(0.25, -0.29, 1))  
#  plot_annotation(title = 'Bemcentinib Modules') & theme(plot.title = element_text(hjust = 0.5))

# Display the combined plot
print(bem_plot)

```

#### (iii) Zilucoplan
```{r}
# Read in data
### for heatmap
zil_heatmap_corr <- read.csv("zilucoplan_module_trait_cor_matrix.csv", row.names = 1, header = TRUE)
zil_heatmap_pval <- read.csv("zilucoplan_module_trait_pvalue_matrix.csv", row.names = 1, header = TRUE)
zil_combo <- merge(zil_heatmap_corr, zil_heatmap_pval, by = "row.names")

# Rename specific modules
row.names(zil_heatmap_corr)[row.names(zil_heatmap_corr) == "MEsilver"] <- "MEslategrey"
row.names(zil_heatmap_pval)[row.names(zil_heatmap_pval) == "MEsilver"] <- "MEslategrey"
zil_combo <- merge(zil_heatmap_corr, zil_heatmap_pval, by = "row.names")

# Module order
zil_module_order <- row.names(zil_heatmap_corr)
colnames(zil_combo)[colnames(zil_combo) == "Row.names"] <- "Modules"

# Module number for barplot
zil_modules <- read.csv("Module_information_Zilucoplan.csv")

# First, summarize the data by moduleColors
zil_module_number <- zil_modules %>%
  group_by(moduleColors) %>%
  summarize(total_genes = n())

# Change silver to slategrey
zil_module_number <- as.data.frame(zil_module_number)
zil_module_number$moduleColors[zil_module_number$moduleColors == "silver"] <- "slategrey"
zil_module_number <- as.data.frame(zil_module_number)
zil_cols = zil_module_number$moduleColors 

# Create new column by adding 'ME' in front of 'zil_module_number'
zil_module_number$Modules <- paste0("ME", zil_module_number$moduleColors)

# Set as row names for ordering the same way as the heatmap 
row.names(zil_module_number) <- zil_module_number$Modules

# Remove the string "ME" from the 'module_column'
zil_heatmap_col_order <- gsub("ME", "", row.names(zil_heatmap_corr)) # Use this as the colors instead
zil_heatmap_col_order[zil_heatmap_col_order == "silver"] <- "slategrey"

# Save this order
zil_module_order <- row.names(zil_heatmap_corr)

# Reorder the levels in Modules
zil_module_number$Modules <- factor(zil_module_number$Modules, levels = zil_module_order)
zil_module_number$moduleColors <- factor(zil_module_number$moduleColors, levels = zil_heatmap_col_order)

# Convert combo dataframe to long format for correlations and p-values
zil_df_long_corr <- zil_combo %>%
  pivot_longer(
    cols = ends_with(".x"),
    names_to = "Trait",
    values_to = "Corr"
  ) %>%
  mutate(Trait = sub("\\.x$", "", Trait))

zil_df_long_pval <- zil_combo %>%
  pivot_longer(
    cols = ends_with(".y"),
    names_to = "Trait",
    values_to = "pval"
  ) %>%
  mutate(Trait = sub("\\.y$", "", Trait))

# Merge the two data frames based on "Modules" and "Trait"
zil_result_df <- merge(zil_df_long_corr, zil_df_long_pval, by = c("Modules", "Trait"))

# Reorder columns if needed
zil_result_df <- zil_result_df[, c("Modules", "Trait", "Corr", "pval")]

# Convert pval to numeric if it's not already
zil_result_df$Modules <- as.character(zil_result_df$Modules)
zil_result_df$pval <- as.numeric(zil_result_df$pval)
zil_result_df$Corr <- as.numeric(zil_result_df$Corr)

# Print the result
print(zil_result_df)

# Define your custom palette
zil_my_palette <- rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582", "#FDDBC7",
                        "#F7F7F7", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"))

# Reorder the levels in Trait
zil_result_df$Trait <- factor(zil_result_df$Trait, levels = rev(c("Day", "Remdesivir", "ICU", "D29_Death_RespFailure", "D29_Sustained_Clinical_Response")))

# Reorder the levels in Modules
zil_result_df$Modules <- factor(zil_result_df$Modules, levels = zil_module_order)

# Create a colour bar
# Extract colors from Module labels
zil_module_colors <- sub("^ME", "", zil_module_order)

# Convert module_colors to factor with desired order
zil_module_colors_factor <- factor(zil_module_colors, levels = unique(zil_module_colors))

# Define the new labels for traits
zil_new_trait_labels <- c(
  "Day" = "Day",
  "Remdesivir" = "Remdesivir",
  "ICU" = "ICU",
  "D29_Death_RespFailure" = "Death or Respiratory Failure",
  "D29_Sustained_Clinical_Response" = "Sustained Clinical Response"
)

# Create barplot
zil_barplot <- ggplot(data = zil_module_number, aes(x = Modules, y = total_genes, fill = zil_cols)) +
  geom_col(aes(x = Modules, y = total_genes, fill = zil_cols), colour = "black", position = "dodge") +
  geom_text(aes(label = total_genes), position = position_dodge(width = 0.9), vjust = -0.25, size = 2) +
  ylab("Number of genes") + 
  ggtitle("Zilucoplan modules")+
  theme_classic(base_size = 8) +
  scale_y_continuous(limits = c(0, 10000)) +
  scale_fill_manual(values = zil_cols) +
  coord_cartesian(expand = FALSE) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(size = 6, vjust = -15),
    plot.title = element_text(hjust = 0.5)
  ) +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank())

# Create the heatmap
zil_heatmap <- ggplot(zil_result_df) +
  geom_tile(aes(x = Modules, y = Trait, fill = Corr, width = 1, height = 1)) +
  scale_fill_gradientn(
    colors = zil_my_palette,  # Use module colors
    limits = c(-1, 1), 
    breaks = c(-1, 0, 1),
    labels = scales::number_format(accuracy = 0.1)) +
  geom_text(aes(x = Modules, y = Trait, label = paste(round(Corr, 2))), 
            vjust = 0, hjust = 0.5, size = 2, color = ifelse(zil_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(zil_result_df$pval < 0.05, "bold", "plain")) +
  geom_text(aes(x = Modules, y = Trait, label = paste("(", format(round(pval, 3), nsmall = 3), ")", sep = "")), # format pvalue to 3 decimal places
            vjust = 1, hjust = 0.5, size = 2, color = ifelse(zil_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(zil_result_df$pval < 0.05, "bold", "plain")) +
  theme_minimal(base_size = 9.5) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank(),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.title = element_text(size = 6),  # Adjust legend title size
    plot.title = element_text(size = 10),  # Adjust title text size
    legend.key.width = unit(0.2, "cm"),  # Adjust legend color scale width
    legend.key.height = unit(0.5, "cm")) +  # Adjust legend color scale height
  labs(
    x = "",
    y = "",
    fill = "Correlation") +
  scale_y_discrete(labels = zil_new_trait_labels) +
  theme(legend.position = "right") +
  coord_fixed(ratio = 1) +
  new_scale_fill() +
  geom_tile(data = tibble(Modules = paste0(zil_module_order),
                          d = zil_module_colors_factor),
            aes(x = Modules, y = 0.25, fill = d, height = 0.5)) +
  scale_x_discrete(labels = c("MEslategrey" = "MEsilver", levels(zil_result_df$Modules))) +
  scale_fill_identity()

# Combine the two plots using patchwork
zil_plot <- zil_barplot / plot_spacer() / zil_heatmap + plot_layout(heights = c(0.25, -0.2, 1)) 
# plot_annotation(title = 'Zilucoplan Modules') & theme(plot.title = element_text(hjust = 0.5))

# Display the combined plot
print(zil_plot)

```

#### (iv) Standard of Care

```{r}
# Read in data
### for heatmap
soc_heatmap_corr <- read.csv("soc_module_trait_cor_matrix.csv", row.names = 1, header = TRUE)
soc_heatmap_pval <- read.csv("soc_module_trait_pvalue_matrix.csv", row.names = 1, header = TRUE)
soc_combo <- merge(soc_heatmap_corr, soc_heatmap_pval, by = "row.names")

# Rename specific modules - again silver won't show as a colour so using slategrey here as a close sub
row.names(soc_heatmap_corr)[row.names(soc_heatmap_corr) == "MEsilver"] <- "MEslategrey"
row.names(soc_heatmap_pval)[row.names(soc_heatmap_pval) == "MEsilver"] <- "MEslategrey"
soc_combo <- merge(soc_heatmap_corr, soc_heatmap_pval, by = "row.names")

# Module order
soc_module_order <- row.names(soc_heatmap_corr)
colnames(soc_combo)[colnames(soc_combo) == "Row.names"] <- "Modules"

# Module number for barplot
soc_modules <- read.csv("Module_information_soc.csv")

# First, summarize the data by moduleColors
soc_module_number <- soc_modules %>%
  group_by(moduleColors) %>%
  summarize(total_genes = n())

# Change silver to slategrey
soc_module_number <- as.data.frame(soc_module_number)
soc_module_number$moduleColors[soc_module_number$moduleColors == "silver"] <- "slategrey"
soc_module_number <- as.data.frame(soc_module_number)
soc_cols = soc_module_number$moduleColors 

# Create new column by adding 'ME' in front of 'soc_module_number'
soc_module_number$Modules <- paste0("ME", soc_module_number$moduleColors)

# Set as row names for ordering the same way as the heatmap 
row.names(soc_module_number) <- soc_module_number$Modules

# Remove the string "ME" from the 'module_column'
soc_heatmap_col_order <- gsub("ME", "", row.names(soc_heatmap_corr)) # Use this as the colors instead
soc_heatmap_col_order[soc_heatmap_col_order == "silver"] <- "slategrey"

# Save this order
soc_module_order <- row.names(soc_heatmap_corr)

# Reorder the levels in Modules
soc_module_number$Modules <- factor(soc_module_number$Modules, levels = soc_module_order)
soc_module_number$moduleColors <- factor(soc_module_number$moduleColors, levels = soc_heatmap_col_order)

# Convert combo dataframe to long format for correlations and p-values
soc_df_long_corr <- soc_combo %>%
  pivot_longer(
    cols = ends_with(".x"),
    names_to = "Trait",
    values_to = "Corr"
  ) %>%
  mutate(Trait = sub("\\.x$", "", Trait))

soc_df_long_pval <- soc_combo %>%
  pivot_longer(
    cols = ends_with(".y"),
    names_to = "Trait",
    values_to = "pval"
  ) %>%
  mutate(Trait = sub("\\.y$", "", Trait))

# Merge the two data frames based on "Modules" and "Trait"
soc_result_df <- merge(soc_df_long_corr, soc_df_long_pval, by = c("Modules", "Trait"))

# Reorder columns if needed
soc_result_df <- soc_result_df[, c("Modules", "Trait", "Corr", "pval")]

# Convert pval to numeric if it's not already
soc_result_df$Modules <- as.character(soc_result_df$Modules)
soc_result_df$pval <- as.numeric(soc_result_df$pval)
soc_result_df$Corr <- as.numeric(soc_result_df$Corr)

# Print the result
print(soc_result_df)

# Define your custom palette
soc_my_palette <- rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582", "#FDDBC7",
                        "#F7F7F7", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"))

# Reorder the levels in Trait
soc_result_df$Trait <- factor(soc_result_df$Trait, levels = rev(c("Day", "Remdesivir", "ICU", "D29_Death_RespFailure", "D29_Sustained_Clinical_Response")))

# Reorder the levels in Modules
soc_result_df$Modules <- factor(soc_result_df$Modules, levels = soc_module_order)

# Create a colour bar
# Extract colors from Module labels
soc_module_colors <- sub("^ME", "", soc_module_order)

# Convert module_colors to factor with desired order
soc_module_colors_factor <- factor(soc_module_colors, levels = unique(soc_module_colors))

# Define the new labels for traits
soc_new_trait_labels <- c(
  "Day" = "Day",
  "Remdesivir" = "Remdesivir",
  "ICU" = "ICU",
  "D29_Death_RespFailure" = "Death or Respiratory Failure",
  "D29_Sustained_Clinical_Response" = "Sustained Clinical Response"
)

# Create barplot
soc_barplot <- ggplot(data = soc_module_number, aes(x = Modules, y = total_genes, fill = soc_cols)) +
  geom_col(aes(x = Modules, y = total_genes, fill = soc_cols), colour = "black", position = "dodge") +
  geom_text(aes(label = total_genes), position = position_dodge(width = 0.9), vjust = -0.25, size = 2) +
  ylab("Number of genes") + 
  ggtitle("Standard of Care modules")+
  theme_classic(base_size = 8) +
  scale_y_continuous(limits = c(0, 10000)) +
  scale_fill_manual(values = soc_cols) +
  coord_cartesian(expand = FALSE) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(size = 6, vjust = -15),
    plot.title = element_text(hjust = 0.5)
  ) +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank())

# Create the heatmap
soc_heatmap <- ggplot(soc_result_df) +
  geom_tile(aes(x = Modules, y = Trait, fill = Corr, width = 1, height = 1)) +
  scale_fill_gradientn(
    colors = soc_my_palette,  # Use module colors
    limits = c(-1, 1), 
    breaks = c(-1, 0, 1),
    labels = scales::number_format(accuracy = 0.1)) +
  geom_text(aes(x = Modules, y = Trait, label = paste(round(Corr, 2))), 
            vjust = 0, hjust = 0.5, size = 2, color = ifelse(soc_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(soc_result_df$pval < 0.05, "bold", "plain")) +
  geom_text(aes(x = Modules, y = Trait, label = paste("(", format(round(pval, 3), nsmall = 3), ")", sep = "")), # format pvalue to 3 decimal places
            vjust = 1, hjust = 0.5, size = 2, color = ifelse(soc_result_df$pval < 0.05, "black", "gray13"), 
            fontface = ifelse(soc_result_df$pval < 0.05, "bold", "plain")) +
  theme_minimal(base_size = 9.5) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank(),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.title = element_text(size = 6),  # Adjust legend title size
    plot.title = element_text(size = 10),  # Adjust title text size
    legend.key.width = unit(0.2, "cm"),  # Adjust legend color scale width
    legend.key.height = unit(0.5, "cm")) +  # Adjust legend color scale height
  labs(
    x = "",
    y = "",
    fill = "Correlation") +
  scale_y_discrete(labels = soc_new_trait_labels) +
  theme(legend.position = "right") +
  coord_fixed(ratio = 1) +
  new_scale_fill() +
  geom_tile(data = tibble(Modules = paste0(soc_module_order),
                          d = soc_module_colors_factor),
            aes(x = Modules, y = 0.25, fill = d, height = 0.5)) +
  scale_x_discrete(labels = c("MEslategrey" = "MEsilver", levels(soc_result_df$Modules))) +
  scale_fill_identity()

# Combine the two plots using patchwork
soc_plot <- soc_barplot / plot_spacer() / soc_heatmap + plot_layout(heights = c(0.25, -0.2, 1)) 
# plot_annotation(title = 'Standard of Care Modules') & theme(plot.title = element_text(hjust = 0.5))

# Display the combined plot
print(soc_plot)

```

#### (v) Combined plot for manuscript figure
```{r fig.width= 20, fig.height=10}
#combined one
# Create the top row of plots (SOC and ZIL)
top_row <- (soc_barplot /plot_spacer() / soc_heatmap + plot_layout(heights = c(0.25, -0.1, 1))) |  # Combine without spacer
  zil_barplot / plot_spacer() /zil_heatmap + plot_layout(heights = c(0.25, -0.1, 1))

# Create the bottom row of plots (BEM and TOZ)
bottom_row <- (bem_barplot /plot_spacer() / bem_heatmap + plot_layout(heights = c(0.25, -0.225,1)) |  # Combine without spacer
                 toz_barplot / plot_spacer() /toz_heatmap + plot_layout(heights = c(0.25, -0.25, 1)))

# Combine both rows, adjusting the space between left and right
combined_plot <- top_row /
  bottom_row + 
  plot_layout(
    widths = c(5, 0.01, 5),   # Reduce space between left and right columns with small value for middle
    heights = c(4, 4),        # Adjust height for top and bottom rows
    guides = "collect", # Collect legends if necessary
    
  )
# Add figure labels manually
combined_plot <- combined_plot + 
  plot_annotation(
    tag_levels = 'A',          # Labels as "A", "B", "C", etc.
    tag_prefix = "(",           # No prefix for simplicity
    tag_suffix = ")",           # No suffix
    tag_sep = "."              # Separator after the label
  ) &
  theme(
    plot.tag.position = c(0.1, 1),   # Adjust position of the figure label (top-left corner)
    plot.tag = element_text(size = 8, face = "bold")  # Customize appearance of labels
  )

# Display the final combined plot with labels
print(combined_plot)
```

## 2b Facet barplot of significant modules 

This is again using the correlation matrix and pval matrix output for each treatment arm saved in 04_coexpression_pywgcna

```{r fig.height = 8, fig.width=5}
#set cols:
# Define a color palette
cols <- rev(scales::hue_pal()(5))
cols <- setNames(cols, levels(combined_sig_df$Trait))


# Function to clean up and process the data for each treatment
process_data <- function(corr_file, pval_file, treatment_name) {
  # Read corr and pval data
  corr_data <- read.csv(corr_file, row.names = 1, header = TRUE)
  pval_data <- read.csv(pval_file, row.names = 1, header = TRUE)
  
  # Merge the correlation and p-value data
  combo <- merge(corr_data, pval_data, by = "row.names")
  colnames(combo)[colnames(combo) == "Row.names"] <- "Modules"
  
  # Reshape data
  long_corr <- combo %>%
    pivot_longer(cols = ends_with(".x"), names_to = "Trait", values_to = "Corr") %>%
    mutate(Trait = sub("\\.x$", "", Trait))
  
  long_pval <- combo %>%
    pivot_longer(cols = ends_with(".y"), names_to = "Trait", values_to = "pval") %>%
    mutate(Trait = sub("\\.y$", "", Trait))
  
  # Combine correlation and p-value data
  df <- merge(long_corr, long_pval, by = c("Modules", "Trait"))
  
  # Filter by p-value < 0.05
  df <- df %>%
    mutate(Treatment = treatment_name) %>%
    filter(pval < 0.05) %>%
    mutate(Modules = as.character(Modules),
           Trait = gsub("D29_Death_RespFailure", "Death or Respiratory Failure", Trait),
           Trait = gsub("D29_Sustained_Clinical_Response", "Sustained Clinical Response", Trait))
  
  return(df)
}

# File paths
file_paths <- list(
  Bem = list(corr = "bemcentinib_module_trait_cor_matrix.csv", 
             pval = "bemcentinib_module_trait_pvalue_matrix.csv"),
  Zil = list(corr = "zilucoplan_module_trait_cor_matrix.csv", 
             pval = "zilucoplan_module_trait_pvalue_matrix.csv"),
  Toz = list(corr = "tozorakimab_module_trait_cor_matrix.csv", 
              pval = "tozorakimab_module_trait_pvalue_matrix.csv"),
  SoC = list(corr = "soc_module_trait_cor_matrix.csv", 
             pval = "soc_module_trait_pvalue_matrix.csv")
)

# Process each dataset and combine into one data frame
combined_sig_df <- bind_rows(
  process_data(file_paths$Bem$corr, file_paths$Bem$pval, "Bemcentinib"),
  process_data(file_paths$Zil$corr, file_paths$Zil$pval, "Zilucoplan"),
  process_data(file_paths$Toz$corr, file_paths$Toz$pval, "Tozorakimab"),
  process_data(file_paths$SoC$corr, file_paths$SoC$pval, "SoC")
)

# Add color based on Treatment
treatment_colors <- c(Bemcentinib = "forestgreen", Tozorakimab = "yellow", Zilucoplan = "red", SoC = "skyblue")
combined_sig_df <- combined_sig_df %>%
  mutate(colour = treatment_colors[Treatment])

swr = function(string, nwrap=10) {
  paste(strwrap(string, width=nwrap), collapse="\n")
}
swr = Vectorize(swr)
# Create line breaks in Year
combined_sig_df$Trait = swr(combined_sig_df$Trait)
# Create levels for colouring facets
treatment_levels <- unique(combined_sig_df$Treatment)
trait_levels <- unique(combined_sig_df$Trait)

# Create a data frame with all possible combinations of Treatment and Trait
colours <- expand.grid(Treatment = treatment_levels, Trait = trait_levels)

# Add Corr and Modules columns to colours dataframe
colours$Corr <- 1
colours$Modules <- 1

# Add specific colour column based on Treatment
colours <- colours %>%
  mutate(colour = treatment_colors[Treatment])

# Set the desired order of levels for Treatment
desired_levels <- c("Bemcentinib", "Tozorakimab", "Zilucoplan", "SoC")

# Reorder the levels of the Treatment factor variable
combined_sig_df$Treatment <- factor(combined_sig_df$Treatment, levels = desired_levels)


# Create a bar plot
ggplot(combined_sig_df, aes(x = Corr, y = Modules, fill = Corr)) +
  theme_bw() +
  geom_rect(data = colours, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf), fill = colours$colour, alpha = 0.15)+ 
  geom_col(position = position_dodge2(preserve = "single"), color = "black") +
  geom_vline(xintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  scale_fill_gradient2(name = "Correlation Score", low = "darkblue", high = "darkred", limits = c(-1, 1)) +
  scale_x_continuous(breaks = c(-1, 0, 1), limits = c(-1, 1)) +
  facet_grid(Treatment ~ Trait, scales = "free", space = "free", switch = "y") +
  scale_y_discrete(position = "right") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "bottom")
```


## 2c Consensus bubble plot of treatment arms vs standard of care

Here plotting the jaccard matrix between each treatment arm and the standard of care to visualise the modules with significant overlap

First prepare the data
```{r}
# Define variables
drug_list <- list(
  list(
    drug = "Tozorakimab",
    study = "Tozorakimab_vs_SoC"
  ),
  list(
    drug = "Bemcentinib",
    study = "Bemcentinib_vs_SoC"
  ),
  list(
    drug = "Zilucoplan",
    study = "Zilucoplan_vs_SoC"
  )
)

# Function to read and preprocess data - these are the jaccard and pvals for each treatment v soc consensus analysis
read_and_preprocess <- function(drug) {
  # Dynamically create file paths based on the drug name
  jaccard_file_name <- paste0("jaccard_matrix_soc_", drug, ".csv")
  p_value_file_name <- paste0("p_value_matrix_soc_", drug, ".csv")
  
  # Read the data
  jaccard <- read.csv(jaccard_file_name, row.names = 1, header = TRUE, check.names = FALSE)
  pval <- read.csv(p_value_file_name, row.names = 1, header = TRUE, check.names = FALSE)
  return(list(jaccard = jaccard, pval = pval))
}

# Function to transform to long format and combine
combine_data <- function(jaccard, pval) {
  jaccard_long <- jaccard %>%
    as.data.frame() %>%
    rownames_to_column("row_names") %>%
    gather(key = "column_names", value = "value.Jaccard", -row_names)
  
  pval_long <- pval %>%
    as.data.frame() %>%
    rownames_to_column("row_names") %>%
    gather(key = "column_names", value = "value.pvalue", -row_names)
  
  combined <- left_join(jaccard_long, pval_long, by = c("row_names", "column_names")) %>%
    mutate(Combination = paste(row_names, column_names, sep = "-"))
  
  return(combined)
}

# Function to filter and process combined data - we only need to keep one set of comparisons or the plot will be duplicated
filter_data <- function(data, drug) {
  filtered <- data %>%
    filter(row_names != column_names) %>%
    filter(
      !(grepl("SoC_Core", row_names) & grepl("SoC_Core", column_names)) &
        !(grepl(drug, row_names) & grepl(drug, column_names))
    ) %>%
    mutate(
      pair = paste(pmin(row_names, column_names), pmax(row_names, column_names), sep = "-")
    ) %>%
    distinct(pair, .keep_all = TRUE) %>%
    select(-pair) %>%
    mutate(log_pvalue = log10(value.pvalue + 1e-320))
  
  return(filtered)
}

# Run functions to process data
# Store filtered data frames
filtered_data_list <- list()

for (drug_info in drug_list) {
  data <- read_and_preprocess(drug_info$drug)
  combined <- combine_data(data$jaccard, data$pval)
  filtered <- filter_data(combined, drug_info$drug)
  
  # Store the filtered data frame in the list
  filtered_data_list[[drug_info$drug]] <- filtered
}

# Assign filtered data frames to the global environment
toz_filtered <- filtered_data_list[["Tozorakimab"]]
bem_filtered <- filtered_data_list[["Bemcentinib"]]
zil_filtered <- filtered_data_list[["Zilucoplan"]]
# Now, `toz_filtered`, `bem_filtered`, and `zil_filtered` are accessible as data frames in the environment.
```

Now plot the combined consensus bubble plot
```{r fig.width=12, fig.height=5}

# Create the plot with the filtered data
toz_filtered$title <- "Tozorakimab"
toz_plot<-ggplot(toz_filtered, aes(x = row_names, y = column_names, size = as.numeric(value.Jaccard), color = as.numeric(log_pvalue))) +
  geom_point(alpha = 0.7) +
  geom_point(color=ifelse(toz_filtered$log_pvalue > -1.30103, 'gray', NA)) +
  theme_bw() +
  theme(
    
    plot.margin = margin(0, 1, 0, 0),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()) +  # Remove minor gridlines
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        axis.title = element_blank(),
        strip.text.x = element_text(size = 20)) +
  scale_size_continuous(range = c(2, 10), name = "Jaccard score") +
  scale_color_viridis(trans = "reverse", name = "log10(p-value)") +
  facet_grid(. ~ title)+
  theme(
    legend.position = "none",
    strip.background = element_rect(
      colour = "black",
      fill = alpha("yellow", 0.5)  # Adjust the alpha value (0.0 to 1.0)
    )
  )


bem_filtered$title <- "Bemcentinib"
bem_plot<-ggplot(bem_filtered, aes(x = row_names, y = column_names, size = as.numeric(value.Jaccard), color = as.numeric(log_pvalue))) +
  geom_point(alpha = 0.7) +
  geom_point(color=ifelse(bem_filtered$log_pvalue > -1.30103, 'gray', NA)) +
  theme_bw() +
  theme(
    plot.margin = margin(0, 1, 0, 1),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()) +  # Remove minor gridlines
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(size = 20)) +
  scale_size_continuous(range = c(2, 10), name = "Jaccard score") +
  scale_color_viridis(trans = "reverse", name = "log10(p-value)") +
  facet_grid(. ~ title)+
  theme(
    legend.position = "none",
    strip.background = element_rect(
      colour = "black",
      fill = alpha("forestgreen", 0.5)  # Adjust the alpha value (0.0 to 1.0)
    )
  )


zil_filtered$title <- "Zilucoplan"
zil_plot<-ggplot(zil_filtered, aes(x = row_names, y = column_names, size = as.numeric(value.Jaccard), color = as.numeric(log_pvalue))) +
  geom_point(alpha = 0.7) +
  geom_point(color=ifelse(zil_filtered$log_pvalue > -1.30103, 'gray', NA)) +
  theme_bw() +
  theme(
    plot.margin = margin(0, 0, 0, 0.1),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()) +  # Remove minor gridlines
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(size = 20)) +
  scale_size_continuous(range = c(2, 10), name = "Jaccard score") +
  scale_color_viridis(trans = "reverse", name = "log10(p-value)") +
  facet_grid(. ~ title)+
  theme(
    # legend.position = "none",
    strip.background = element_rect(
      colour = "black",
      fill = alpha("red", 0.5)  # Adjust the alpha value (0.0 to 1.0)
    )
  )

# Combine the two plots using patchwork
plot <- toz_plot | bem_plot | zil_plot
plot+ plot_layout(       # Adjust height for top and bottom rows
  guides = "collect"# Collect legends if necessary
  
)


```

## 2d Hub gene extraction for visualisation in cytoscape

For each treatment module within a cluster, the top 10 hub genes, based on connectivity score are extracted for plotting in cytoscape. The input file is the modulecolour_top_hub_genes.csv for each module saved from the 04_coexpression_pywgcna using the top_n_hub_genes function. This code pulls in these files which are saved in subdirectories of treatment/modulecolour depending on the treatment and modulecolour

```{r}
# Drug names and associated modules
drug_modules <- list(
  Bemcentinib = c("gainsboro", "tomato", "linen", "darksalmon", "snow", "white", "rosybrown", "darkgrey", "silver", "lightcoral"),
  Tozorakimab = c("salmon", "tan", "chocolate", "sandybrown", "orangered", "burlywood", "moccasin", "gainsboro"),
  Zilucoplan = c("brown", "coral", "black", "gainsboro", "indianred", "darkgrey", "lightgrey", "chocolate")
)

# Cluster definitions
clusters <- list(
  Cluster1 = list(Bemcentinib = c("gainsboro"), Tozorakimab = c("salmon"), Zilucoplan = c("brown")),
  Cluster2 = list(Bemcentinib = c("tomato", "linen"), Tozorakimab = c("tan"), Zilucoplan = c("coral")),
  Cluster3 = list(Bemcentinib = c("rosybrown"), Tozorakimab = c("orangered", "burlywood"), Zilucoplan = c("indianred")),
  Cluster4 = list(Bemcentinib = c("snow", "white"), Tozorakimab = c("sandybrown"), Zilucoplan = c("gainsboro")),
 Cluster5 = list(Bemcentinib = c("darksalmon"), Tozorakimab = c("chocolate"), Zilucoplan = c("black")),
   Cluster6 = list(Bemcentinib = c("darkgrey", "silver", "lightcoral"), Tozorakimab = c("moccasin", "gainsboro"), Zilucoplan = c("darkgrey", "lightgrey", "chocolate"))
)


# Helper function to create variable names and file paths
create_variable_and_read <- function(drug, color) {
  var_name <- paste0(tolower(drug), "_", color)
  file_path <- paste0("./", drug, "/hub_genes/", color, "_top_hub_genes.csv")
  df <- read.csv(file_path)
  
  # Add new columns
  df$Treatment_Arm <- drug
  df$Module <- color
  df$Treatment_Arm_Module <- paste0(drug, "_", color)
  
  # Replace empty strings in gene_name with corresponding X value
  df$gene_name[df$gene_name == ""] <- as.character(df$X[df$gene_name == ""]) 
  # Order by connectivity and extract top rows
  df <- df[order(-df$connectivity),]
  top10 <- head(df, 10)
  
  # Assign to global environment
  assign(paste0(var_name, "_top10"), top10, envir = .GlobalEnv)
 }

# Loop through each drug and their associated colors
for (drug in names(drug_modules)) {
  for (color in drug_modules[[drug]]) {
    create_variable_and_read(drug, color)
  }
}

# Function to collect cluster data
collect_cluster_data <- function(cluster_name, cluster_drugs, top_n) {
  cluster_data <- data.frame()
  
  for (drug in names(cluster_drugs)) {
    for (color in cluster_drugs[[drug]]) {
      var_name <- paste0(tolower(drug), "_", color, "_top", top_n)
      if (exists(var_name, .GlobalEnv)) {
        df <- get(var_name)
        cluster_data <- rbind(cluster_data, df)
      }
    }
  }
  
  return(cluster_data)
}

# Loop through each cluster and create top_n data frames and node files
for (n in 10) {
  for (i in 1:6) {
    # Collect cluster data
    cluster_data <- collect_cluster_data(paste0("Cluster", i), clusters[[paste0("Cluster", i)]], n)
    cluster_var_name <- paste0("cluster", i, "_top", n)
    assign(cluster_var_name, cluster_data)
    
    # Save the cluster data to CSV
    write.csv(cluster_data, paste0("Cluster", i, "_top", n, "_all_treatments.csv"), quote = FALSE)
    
    # Create node file for top 10 only
    if (n == 10) {
      # Start with the existing cluster data
      node_file <- cluster_data
      
      # Remove the first five columns
      node_file <- node_file[, -c(1:5)]
      
      # Add three new rows for Bemcentinib, Tozorakimab, and Zilucoplan
      new_rows <- data.frame(
        gene_name = c("Bemcentinib", "Tozorakimab", "Zilucoplan"),
        gene_biotype = NA,
        Treatment_Arm = c("Bemcentinib", "Tozorakimab", "Zilucoplan"),
        Module = NA,
        Treatment_Arm_Module = c("Bemcentinib", "Tozorakimab", "Zilucoplan")
      )
      node_file <- rbind(node_file, new_rows)
      
      # Count occurrences of each gene_name
      gene_counts <- table(node_file$gene_name)
      
      # Add the overlap_colour column
      node_file$overlap_colour <- ifelse(
        node_file$gene_name %in% c("Bemcentinib", "Tozorakimab", "Zilucoplan"),
        node_file$gene_name,  # For treatments, use the treatment name
        ifelse(duplicated(node_file$gene_name) | duplicated(node_file$gene_name, fromLast = TRUE),
               "yes",  # If duplicated, set as "yes"
               node_file$Treatment_Arm  # If not duplicated, use the Treatment_Arm
        )
      )
      
      # Add the overlap_size column
      node_file$overlap_size <- ifelse(
        node_file$gene_name %in% c("Bemcentinib", "Tozorakimab", "Zilucoplan"),
        node_file$gene_name,  # For treatments, use the treatment name
        as.integer(gene_counts[node_file$gene_name])  # For others, use count
      )
      
      # Add the type column
      node_file$type <- ifelse(
        node_file$gene_name %in% c("Bemcentinib", "Tozorakimab", "Zilucoplan"),
        "treatment",
        "gene"
      )
      
      # Save the node file to CSV
      write.csv(node_file, paste0("Cluster", i, "_top10_all_treatments_node.csv"), quote = FALSE)
    }
  }
}


```

The output file will have columns to be used for aesthetic mapping (colour, size etc) in cytoscape.


