---
title: "06_Blood_Transcription_Module_Analysis"
#output: html_notebook
---
# Load libraries

```{r message=FALSE}
# Packages to load
packages <- c(
  "clusterProfiler", "ComplexHeatmap", "fgsea", "gpplot2", "pheatmap",
  "rstatix", "tidyverse", "VennDiagram")

########################################################################################################
load_libraries <- function(packages) {
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

# Load all packages
load_libraries(packages)

```



# Blood Transcription Module (BTM) Analysis

## Generate Activity Scores

using the btm_tool programme downloaded from the supplementary information of the
following manuscript: - Li, S., Rouphael, N., Duraisingham, S., Romero-Steiner, S.,
Presnell, S., Davis, C., ..., Pulendran, B. (2014). Molecular signatures of antibody
responses derived from a systems biological study of 5 human vaccines. Nature immunology,
15(2), 195. <https://www.nature.com/articles/ni.2789>

Activity scores were created for each arm of the trial using normalised gene expression
matrix of each arm for Day 1 and Day 5, which can be extracted from the corresponding 03_differential_expression notebook:

```{python}
# Need to evoke Python in the same directory where the downloaded files from Li et al (2014) were unpacked, which should include btm_tool.py and others.

from btm_tool import * #This makes BTM data and a few functions available for use
genetable_to_activityscores('soc_D1D5_CPM.txt','soc_BTM_activity.txt')
genetable_to_activityscores('bem_D1D5_CPM.txt','bem_BTM_activity.txt')
genetable_to_activityscores('toz_D1D5_CPM.txt','toz_BTM_activity.txt')
genetable_to_activityscores('zil_D1D5_CPM.txt','zil_BTM_activity.txt')
```

activity score (txt) outputs were then saved as a Rdata object for subsequent analysis (btm_activity_scores.Rdata)

## Compare activity scores between Day 1 and Day 5

Differentially expressed BTM 

These generated BTM activity scores were then subsequently used to compare differences in module activity scores between Day 1 and Day 5 for each treatment group individually using a paired Wilcoxon test.


```{r}
#load your saved btm activity scores
load("btm_activity_scores.Rdata")
#read in metadata
metadata<-read.csv("accord_metadata.csv")
# List of treatment arms and corresponding BTM activity objects
treatment_arms <- c("Tozorakimab", "Bemcentinib", "Zilucoplan", "Standard of Care")

btm_activity_objects <- list(
  "Tozorakimab" = toz_BTM_activity,
  "Bemcentinib" = bem_BTM_activity,
  "Zilucoplan" = zil_BTM_activity,
  "Standard of Care" = soc_BTM_activity
)

# List to store the results for all treatment arms
all_results <- list()

# Loop through each arm
for (treatment in treatment_arms) {
  
  # Filter metadata for the current treatment arm
  metadata_filtered <- metadata %>%
    filter(Treatment_Arm == treatment) %>%
    select("Sample_id", "Unique_ID", "Day")
  
  # Set row names for metadata
  row.names(metadata_filtered) <- metadata_filtered$Sample_id
  
  # Get the corresponding BTM activity object for the current treatment arm
  activity_data <- btm_activity_objects[[treatment]]
  
  # Set row names for activity data and transpose
  row.names(activity_data) <- activity_data$external_gene_name
  activity_data <- activity_data[-1]
  activity_score <- t(activity_data)
  
  # Combine metadata with activity scores
  meta_score <- cbind(metadata_filtered, activity_score)
  
  # Clean col names for plotting
  names(meta_score)[4:ncol(meta_score)] <- gsub("[[:punct:]]", "", names(meta_score)[4:ncol(meta_score)])
  names(meta_score)[4:ncol(meta_score)] <- gsub(" ", "_", names(meta_score)[4:ncol(meta_score)])
  
  # Convert Day to a factor to set levels
  meta_score$Day <- factor(meta_score$Day, levels = c("D1", "D5"))
  
  # Empty list to store results for this treatment arm
  results_list <- list()
  
  # Loop through each column for Wilcoxon test
  columns_to_test <- names(meta_score)[4:ncol(meta_score)]
  for (col in columns_to_test) {
    data_subset <- meta_score[[col]]
    if (length(unique(data_subset[!is.na(data_subset)])) > 1) {
      test <- meta_score %>%
        wilcox_test(as.formula(paste(col, "~ Day")), paired = TRUE) %>%
        add_xy_position(x = "Day")
      results_list[[col]] <- test
    } else {
      cat("Skipping column", col, "as it has less than two unique non-NA values.\n")
    }
  }
  
  # Combine results into a single dataframe
  results_df <- do.call(rbind, results_list)
  results_df$groups <- NULL
  
  # Add to the overall results list with the treatment arm as the name
  all_results[[treatment]] <- results_df
}

# Access individual results
toz_results <- all_results[["Tozorakimab"]]
bem_results <- all_results[["Bemcentinib"]]
zil_results <- all_results[["Zilucoplan"]]
soc_results <- all_results[["Standard of Care"]]

# Save all results as an R object
save(soc_results,bem_results,toz_results,zil_results, 
     file = "btm_treatment_results.Rdata")


```

## Plot number of BTM between Day 1 and Day 5
Determine how many BTMs are significantly different between Day 1 and Day 5 for each treatment arm and plot

```{r}

# List to store significant results for each treatment arm
significant_results_list <- list()

# Loop through each treatment arm to get only significant results
for (treatment in treatment_arms) {
  # Get the results dataframe for the treatment arm
  treatment_results <- all_results[[treatment]]
  
  # Filter for significant modules (p < 0.05)
  significant_results <- subset(treatment_results, p < 0.05)
  
  # Check if significant_results is not empty
  if (nrow(significant_results) > 0) {
    # Order by p-value
    significant_results <- significant_results[order(significant_results$p), ]
    
    # Add a factor for module names and treatment arm for later grouping
    significant_results$.y.<- factor(significant_results$.y., levels = significant_results$.y.)
    significant_results$Treatment_Arm <- treatment
    
    # Reverse the levels of the module factor for consistent plotting
    significant_results$.y. <- factor(significant_results$.y., levels = rev(levels(significant_results$.y.)))
    
    # Add to the list
    significant_results_list[[treatment]] <- significant_results
  } else {
    # If no significant results, create an empty dataframe with the same structure
    significant_results_list[[treatment]] <- data.frame(module = character(),
                                                        p = numeric(),
                                                        Treatment_Arm = character())
    cat("No significant results for", treatment, "\n")
  }
}

# Create a summary dataframe for the number of significant modules
total_variable_number <- data.frame(
  Treatment_Arm = treatment_arms,
  Number_of_Variables = sapply(significant_results_list, function(x) nrow(x))
)

# Reorder the levels of the Treatment_Arm factor for the plot
desired_order <- c("Standard of Care", "Bemcentinib", "Tozorakimab", "Zilucoplan")
total_variable_number$Treatment_Arm <- factor(total_variable_number$Treatment_Arm, levels = desired_order)

# Plot the summary
ggplot(total_variable_number, aes(x = Treatment_Arm, y = Number_of_Variables, fill = Treatment_Arm)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = Number_of_Variables), vjust = -0.5, color = "black", size = 3) +  # Add text labels
  labs(title = "Significant Blood Transcription Modules",
       x = "",
       y = "Number of Modules") +
  theme_bw() +
  scale_fill_manual(values = alpha(colors, 0.6)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 150)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(), 
        legend.position = "right")

```


## BTM categories

Plotting stacked barchart of each defined category to determine which immune/celll categories are most significant between groups. 

Category file included in Rdata file was manually curated, where the list of modules were assigned to specific immune cell or immune/biological categories based on the module name.

```{r}

toz_significant_results <- significant_results_list[["Tozorakimab"]]
bem_significant_results <- significant_results_list[["Bemcentinib"]]
zil_significant_results <- significant_results_list[["Zilucoplan"]]
soc_significant_results <- significant_results_list[["Standard of Care"]]
all_results <- bind_rows(bem_significant_results, toz_significant_results, zil_significant_results, soc_significant_results)

# Keep significant ones - should already be sig only but good to double check here
significant_results <- all_results %>%
  filter(p <0.05)

# Subset the dataframe to remove columns with all NA values
btm_module_categories <- btm_module_categories[, colSums(!is.na(btm_module_categories)) > 0, drop = FALSE]

# Reshape the category dataframe to long format
category_long <- btm_module_categories %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Variable") %>%
  filter(!is.na(Variable) & Variable != "")  # Filter out empty or NA values in the Variable column

# Join significant_results with category_long based on matching variables
significant_results_long<- significant_results %>%
  left_join(category_long, by = c(".y." = "Variable"))
# Count the occurrences of each Category for each Treatment_Arm
category_counts <- significant_results_long %>%
  group_by(Treatment_Arm, Category) %>%
  summarise(Count = n())

#Calculate sums for each category
category_sums <- aggregate(Count ~ Category, data = category_counts, FUN = sum)
# Merge sums with original data
category_counts <- merge(category_counts, category_sums, by = "Category", suffixes = c("", "_sum"))
#this oesn't show the UNIQUE number of categories (as there are some overlaps)

# Reorder the Category variable based on the Number_of_Variables values in reverse order
category_counts$Count_sum[is.na(category_counts$Count_sum)] <- 0
category_counts$Category <- fct_reorder(category_counts$Category, category_counts$Count_sum, .desc = FALSE)
# Plot the bar plot
#Define the desired order of Treatment_Arm levels and corresponding colors
desired_order <- c("Standard of Care", "Bemcentinib", "Tozorakimab", "Zilucoplan")
colors <- c("skyblue", "forestgreen", "yellow", "red")

# Convert Treatment_Arm to a factor with desired order
category_counts$Treatment_Arm <- factor(category_counts$Treatment_Arm, levels = desired_order)
category_counts$Category <- str_replace(category_counts$Category, "(\\w)(cell)", "\\1 cell")  # Add space before "Cell" if preceded by a word
category_counts$Category <- gsub("[[:punct:]]", " ", category_counts$Category)  # Remove punctuation
ggplot(category_counts, aes(x = Count, y = Category, fill = Treatment_Arm)) +
  geom_bar(stat = "identity", position = "stack", color="black") +
  labs(y = "", x = "Number of modules", fill = "Treatment Arm") +
  ggtitle("Number of Categories for Each Treatment Arm")+
  theme_bw() +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 35)) +
  theme(plot.margin = margin(0.5,0.5,0.5,0.5, "cm"))+
  scale_fill_manual(values = alpha(colors, 0.6)) + 
  scale_y_discrete(limits=rev)+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        #axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "right",
        plot.title.position = "plot")
```


## BTM overlap between treatment arms

look at overlaps of BTM between treatment groups by creating a venn diagram. Can then extract each venn section to look at in more detail if necessary.


```{r}
# Create Venn diagram
venn_list <- list(
  Bem = bem_significant_results$.y.,
  Toz = toz_significant_results$.y.,
  Zil = zil_significant_results$.y.,
  SoC = soc_significant_results$.y.
)

# Plot the Venn diagram
venn.plot <- venn.diagram(
  x = venn_list,
  filename = NULL,
  disable.logging = FALSE, 
  margin = 0.09,
  category.names = c("Bemcentinib", "Tozorakimab",  "Zilucoplan","Standard of Care"),
  #circles
  fill = c(alpha('forestgreen', 0.5),  alpha('#fde725ff', 0.5),alpha('red', 0.6), alpha('skyblue', 0.9)),
  #numbers
  print.mode = c("raw", "percent"),
  cex = 0.8,
  #fontface = "italic",
  fontfamily = "sans",
  # ext.text=FALSE,
  ext.length = 0.5,
  ext.pos = 90,
  ext.percent = 0.005,
  # Set names
  cat.cex = 1,
  cat.fontface = "bold",
  cat.fontfamily = "sans",
  cat.default.pos = "outer")

# Display the Venn diagram
grid.newpage()
grid.draw(venn.plot)
```





# Gene Set enrchment analysis of gene expression data using BTM as gene sets
For this, will use the differential expression data from 03_differential_analysis files for each of the treatment arms.
We will also use the BTM gene set file included in Li et al 2014 (BTM_for_GSEA_20131008.gmt) https://github.com/shuzhao-li/BTM

```{r}
#load data DEG data 
# Read Bem, MEDI, and Zil datasets
bem_degs <- read.table("Bemcentinib_Day5_vs_Day1_results.txt", sep = "\t", header = TRUE, row.names = 1)
toz_degs <- read.table("Tozorakimab_Day5_vs_Day1_results.txt", sep = "\t", header = TRUE, row.names = 1)
zil_degs <- read.table("Zilucoplan_Day5_vs_Day1_results.txt", sep = "\t", header = TRUE, row.names = 1)
soc_degs<-read.table("Standard_of_Care_Day5_vs_Day1_results.txt", sep = "\t", header = TRUE, row.names = 1)

# Create metric for ranking
bem_degs$metric <- -log10(bem_degs$PValue) / sign(bem_degs$logFC)
toz_degs$metric <- -log10(toz_degs$PValue) / sign(toz_degs$logFC)
zil_degs$metric <- -log10(zil_degs$PValue) / sign(zil_degs$logFC)
soc_degs$metric <- -log10(soc_degs$PValue) / sign(soc_degs$logFC)

# Create ranked list for each treatment arm using new metric
#here using pvalue and not adjusted/FDR p as this will results in fewer ties 
bem_ranks <- bem_degs$metric 
names(bem_ranks) <- bem_degs$gene_name 
bem_ranks <- sort(bem_ranks, decreasing = TRUE)

toz_ranks <- toz_degs$metric 
names(toz_ranks) <- toz_degs$gene_name 
toz_ranks <- sort(toz_ranks, decreasing = TRUE)

zil_ranks <- zil_degs$metric 
names(zil_ranks) <- zil_degs$gene_name 
zil_ranks <- sort(zil_ranks, decreasing = TRUE)

soc_ranks <- soc_degs$metric 
names(soc_ranks) <- soc_degs$gene_symbol 
soc_ranks <- sort(soc_ranks, decreasing = TRUE)



# Load pathways
#for this, using the BTM pathway file
pathways <- gmtPathways("BTM_for_GSEA_20131008.gmt")



#pathways table needs converting into df for GSEA
# Convert list to tibble
pathways_tbl <- tibble(pathway = names(pathways), genes = pathways)
pathways_unnested<-as.data.frame(pathways_tbl)
set.seed(1234)
```


```{r}
# Perform GSEA using clusterProfiler
bem_GSEA <- GSEA(bem_ranks, TERM2GENE = pathways_unnested, seed = TRUE)
toz_GSEA <- GSEA(toz_ranks, TERM2GENE = pathways_unnested, seed = TRUE)
zil_GSEA <- GSEA(zil_ranks, TERM2GENE = pathways_unnested, seed = TRUE)
soc_GSEA <- GSEA(soc_ranks, TERM2GENE = pathways_unnested, seed = TRUE)
# save GSEA results to read back in if necessary
#save(bem_GSEA, toz_GSEA,zil_GSEA, soc_GSEA, file = "btm_GSEA.Rdata")

# Plot dotplot for Bem, MEDI, and Zil
dotplot(bem_GSEA, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
dotplot(toz_GSEA, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
dotplot(Zil_GSEA, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
dotplot(soc_GSEA, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
```


## Visualise enriched BTM across treatment arms

Here will be visualizing the enriched BTM that are specifically assoicated with a cell type to detemrine enrichment of cellular activation in the samples

```{r}
#extract terms that match cells of interest
# monocyte
# B cells
# T cells
# NK cells
# Neutrophils
# plasma cell
# lymphocyte
# DC
# dendritic cell
#  myeloid
#platelets
#thymocyte

# Function to extract rows cntaining specified keywords and add a new column for cell type
extract_keywords <- function(data, keywords) {
  # Initialize an empty dataframe to store the extracted rows
  extracted_data <- data.frame(matrix(ncol = ncol(data) + 1, nrow = 0))
  colnames(extracted_data) <- c(colnames(data), "cell_type")
  
  # Loop through each keyword and extract matching rows
  for (keyword in keywords) {
    pattern <- paste0("\\b", keyword, "\\b")  # Word boundary for exact match
    matching_rows <- data[grep(pattern, data$ID), ]
    if (nrow(matching_rows) > 0) {
      matching_rows$cell_type <- keyword
      extracted_data <- rbind(extracted_data, matching_rows)
    }
  }
  
  # Remove duplicates
  extracted_data <- unique(extracted_data)
  
  return(extracted_data)
}

# Define keywords - spelling differs so need to capture all iterations of each 
keywords <- c("monocyte", "monocytes", "Monocyte", "B cells", "B cell", "T cells", "T cell", "thymocyte", "thymocytes", "Thymocyte",
              "Thymocytes", "NK cells", "NK cell",
              "Neutrophils", "Neutrophil","neutrophils", "neutrophil", "plasma cell", "lymphocyte", "DC", "DCs", 
              "dendritic cell", "dendritic cells","myeloid", "platelet", "Plasma cell", "plasma cell", "Plasma cells", "plasma cells")



# Read and extract keywords for each dataframe
bem_GSEA_cell <- extract_keywords(bem_GSEA, keywords)
bem_GSEA_cell$Treatment <- "Bemcentinib"
toz_GSEA_cell <- extract_keywords(toz_GSEA, keywords)
toz_GSEA_cell$Treatment <- "Tozorakimab"
zil_GSEA_cell <- extract_keywords(zil_GSEA, keywords)
zil_GSEA_cell$Treatment<- "Zilucoplan"
soc_GSEA_cell <- extract_keywords(soc_GSEA, keywords)
soc_GSEA_cell$Treatment<- "SoC"

######## plot all 4 arms ######################
combined_cell<-rbind(bem_GSEA_cell, toz_GSEA_cell, zil_GSEA_cell, soc_GSEA_cell)
#subset to keep 
combined_cell <- combined_cell[, c("ID", "NES",  "qvalue", "Treatment","cell_type")]


#want to delete certain cells where "myeloid" overlaps with othercell types
table(combined_cell$ID, combined_cell$cell_type)
# Define the IDs to match
ids_to_remove <- c("myeloid, dendritic cell activation via NFkB (II) (M43.1)",
                   "myeloid, dendritic cell activation via NFkB (I) (M43.0)",
                   "proinflammatory dendritic cell, myeloid cell response (M86.1)",
                   "enriched in myeloid cells and monocytes (M81)", 
                   "enriched in activated dendritic cells/monocytes (M64)", 
                   "plasma cells & B cells, immunoglobulins (M156.0)")

# need to remove duplicated ids - where there are more than one cell type
combined_cell <- combined_cell[!(combined_cell$cell_type == "myeloid" & combined_cell$ID %in% ids_to_remove), ]
combined_cell <- combined_cell[!(combined_cell$cell_type == "monocytes" & combined_cell$ID %in% ids_to_remove), ]
combined_cell <- combined_cell[!(combined_cell$cell_type == "B cells" & combined_cell$ID %in% ids_to_remove), ]


# also neeed to redefine the following term as myeloid - because it has multiple cell types
#Identify the row index with the specified ID
row_index <- which(combined_cell$ID == "enriched in activated dendritic cells/monocytes (M64)")
# Change
combined_cell$cell_type[row_index] <- "Myeloid"


# Replace terms in the cell_type column of combined_cell
combined_cell$cell_type <- gsub("\\bmonocyte\\b", "Monocytes", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bmonocytes\\b", "Monocytes", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bMonocyte\\b", "Monocytes", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bB cell\\b", "B cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bT cell\\b", "T cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bdendritic cells\\b", "Dendritic cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bdendritic cell\\b", "Dendritic cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bDC\\b", "Dendritic cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bNK cell\\b", "NK cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bNK cells\\b", "NK cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bmyeloid\\b", "Myeloid", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bneutrophils\\b", "Neutrophils", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bneutrophil\\b", "Neutrophils", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bplatelet\\b", "Platelets", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bthymocytes\\b", "T cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bPlasma cell\\b", "B cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\bplasma cells\\b", "B cells", combined_cell$cell_type)
combined_cell$cell_type <- gsub("\\blymphocyte\\b", "B cells", combined_cell$cell_type)



combined_cell$cell_type
#write.csv(combined_cell,"C:/Users/ja2y21/OneDrive - University of Southampton/PostDoc/Projects/EME COVID/Data/Analysis/BTM/Results/GSEA/treatment_combined_cell_type_terms.csv")
```

Plot enriched cell types
```{r}
#plot all treatment arms
#with two facets - one as cell type and one as treatment arm
# y axis is terms, x axis is NES score, fill is adj pvalue or qvalue 
#background colour is drug arm 
combined_cell$cell_type<-as.character(combined_cell$cell_type)
#add colour column 
treatment_colors <- c(Bemcentinib = "forestgreen",
                      Tozorakimab = "yellow",
                      Zilucoplan = "red", 
                      SoC = "skyblue")


# Add a new column 'colour' based on the values in the 'Treatment' column
combined_cell <- combined_cell %>%
  mutate(colour = treatment_colors[Treatment])


#This code is if we need to wrap the facet label titles
# swr = function(string, nwrap=10) {
#   paste(strwrap(string, width=nwrap), collapse="\n")
# }
# swr = Vectorize(swr)
# # Create line breaks 
# combined_cell$ID = swr(combined_cell$ID)

#create levels for colouring facets
treatment_levels <- unique(combined_cell$Treatment)
cell_levels <- unique(combined_cell$cell_type)

# Create a data frame with all possible combinations of Treatment and Trait
colours <- expand.grid(Treatment = treatment_levels, cell_type = cell_levels)

# Add Corr and Modules columns
colours$NES <- 1
colours$ID <- 1

# Add a new column 'colour' based on the values in the 'Treatment' column
colours <- colours %>%
  mutate(colour = treatment_colors[Treatment])


#set levels in treatment
# Define the desired order of levels
desired_levels <- c("SoC","Bemcentinib", "Tozorakimab", "Zilucoplan")

# Reorder the levels of the Treatment factor variable
combined_cell$Treatment <- factor(combined_cell$Treatment, levels = desired_levels)

colnames(combined_cell)
combined_cell$cell_type
combined_cell$cell_type <- factor(combined_cell$cell_type, levels = c( "B cells","T cells", "NK cells", "Platelets", "Myeloid","Dendritic cells", "Monocytes",  "Neutrophils"))
combined_cell$cell_type
# Check for NAs in the combined_cell$cell_type column
any_na <- any(is.na(combined_cell$cell_type))

# Print the result
if (any_na) {
  print("There are NAs in the combined_cell$cell_type column.")
} else {
  print("There are no NAs in the combined_cell$cell_type column.")
}

combined_cell$NES<-as.numeric(combined_cell$NES) #make sure NES is numeric for plottinh
```


Plot bubble plot - this code is for all four treatment arms. To plot drug arms only, only
combine the drug dfs and edit subsequent code pertaining to fourth soc arm. To plot only
SoC, do the opposite and remove any code pertaining to the drugs.

```{r, fig.width= 10, fig.height=10}
#plot dotplot
width_scale<-6
ggplot(combined_cell, aes(x = Treatment, y = ID, size = -log10(qvalue), fill = NES)) +
  geom_rect(data = colours, inherit.aes = FALSE, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf), fill = colours$colour, alpha = 0.15) + 
  geom_point(aes(fill = NES), shape = 21) +
  theme_bw() +
  scale_fill_distiller(palette = "RdBu", limits = c(-3, 3))+
  scale_size_continuous(name = "-log10(qvalue)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_blank(), axis.text = element_text(size = 12),
        axis.text.x.bottom = element_blank(), axis.ticks.x = element_blank()) +
  facet_grid(cell_type ~ Treatment, scales = "free", space = "free", switch = "y") +
  scale_y_discrete(position = "right") +
  theme(legend.text=ggplot2::element_text(size=width_scale),
        legend.box.margin = margin(6, 6, 6, 6),
        legend.title=ggplot2::element_text(size=1.5*width_scale,face="bold"),
        legend.position="bottom",
        legend.key.size = grid::unit(width_scale/50, "inch"),
        legend.key.width = grid::unit(width_scale/50, "inch"))

```


## Plotting BTM activity scores as heatmap

As per manuscript figure, the SoC arm BTM activity scores were visualised as a heatmap


```{r}
#need activity scores for soc arm - read back in saved file if needed
SoC_activity_score<-read.table("SoC_BTM_activity.txt",header = TRUE, sep = "\t", fill = TRUE, row.names=1)
# Remove punctuation from row names to match the results df which already has had the punctiation removed
rownames(SoC_activity_score) <- gsub("[[:punct:]]", "", rownames(SoC_activity_score))
# Relace blank spaces with underscores in row names
rownames(SoC_activity_score) <- gsub(" ", "_", rownames(SoC_activity_score))
#filter to keep the sig ones only
results_df<-read.csv("SoC_BTM_activty_Day_stat_test_result.csv")

##extract only significant ones
significant_results <- subset(results_df, p <0.05)
# Filter significant results
# Extract the unique values from the ".y." column in significant_results
unique_values <- unique(significant_results$.y.)
# Filter row names of SoC_activity_score based on unique values
filtered_row_names <- rownames(SoC_activity_score)[rownames(SoC_activity_score) %in% unique_values]
# Filter SoC_activity_score based on filtered row names
significant_SoC_activity_score <- SoC_activity_score[filtered_row_names, ]


#using vst data for this
heat_data<-(significant_SoC_activity_score)
heat_data<-as.data.frame(heat_data)
# need to scale vst counts - so can get a range for heatmap
#scale data across rows to get z-scores
scaled_data<-t(scale(t(heat_data)))
#transform data into matrix for heatmap to read
mat<-data.matrix(scaled_data)



# now to make pretty
#create colour pallete for all
col = circlize::colorRamp2(c(-3, 0, 3), hcl_palette ="RdYlBu", reverse = TRUE)

#create column annotation
#need soc meta
SoC_metadata<- metadata %>%
    filter(Treatment_Arm == "Standard of Care") %>%
    select("Sample_id", "Unique_ID", "Day", "Sustained_Clinical_Response_by_D29","Death_by_D29", "Death_or_Respiratory_Failure_at_D29")
row.names(SoC_metadata)<-SoC_metadata$Sample_id

# Check if column names of heat_data are in the same order as row names of metadata_filtered
all.equal(colnames(heat_data), rownames(SoC_metadata))
#NO
# Order SoC_metadata_filtered by row names alphabetically
SoC_metadata_sorted <- SoC_metadata[order(rownames(SoC_metadata)), ]
# Check  order
all.equal(colnames(heat_data), rownames(SoC_metadata_sorted))
#[1] TRUE

#################################################################
# Extract row names from the original data frame
row_names <- rownames(SoC_metadata_sorted)
annotation_col = data.frame(
  Day = SoC_metadata_sorted$Day, 
  Sustained_Clinical_Response_by_D29 = SoC_metadata_sorted$Sustained_Clinical_Response_by_D29,
  Death_or_Respiratory_Failure_at_D29=SoC_metadata_sorted$Death_or_Respiratory_Failure_at_D29,
  Death_by_D29 = SoC_metadata_sorted$Death_by_D29,
  row.names = row_names)


ann_colours<-list("Day"=c("D1"="darkgreen", 
                          "D5"="mediumpurple"), 
                  "Sustained_Clinical_Response_by_D29"=c("Yes"="skyblue", 
                                                         "No"="orange"), 
                  "Death_or_Respiratory_Failure_at_D29"=c("Yes"="coral", 
                                                         "No"="lightyellow"), 
                  "Death_by_D29"=c("Yes"="deeppink3", 
                                  "No"="darkslategrey"))
#now change col names so can read better in the plot
# Define new column names
new_col_names <- c("Day", "Sustained\nClinical Response", "Death or\nRespiratory Failure", "Death")
# Assign new column names to annotation_col
colnames(annotation_col) <- new_col_names
# Define new names for the list elements based on annotation_col column names
new_names <- colnames(annotation_col)
# Rename the list elements in ann_colours
names(ann_colours) <- new_names


#also modify mat names for plotting
# Custom function to modify row names
modify_row_names <- function(row_names) {
  # Split the row names by '_'
  split_names <- strsplit(row_names, "_")
  modified_names <- character(length(split_names))
  
  for (i in seq_along(split_names)) {
    last_string <- split_names[[i]][length(split_names[[i]])]  # Extract the last string
    # Insert '.' after the 4th character if the last string is longer than 4 characters
    if (nchar(last_string) > 4) {
      last_string <- paste(substr(last_string, 1, 4), ".", substr(last_string, 5, nchar(last_string)), sep = "")
    }
    # Enclose the last string in parentheses
    last_string <- paste0("(", last_string, ")")
    # Remove '_' and replace with blank spaces
    modified_last_string <- gsub("_", " ", last_string)
    
    # Combine modified last string with the rest of the strings
    modified_names[i] <- paste(split_names[[i]][-length(split_names[[i]])], collapse = " ")
    modified_names[i] <- paste(modified_names[i], modified_last_string)
  }
  
  return(modified_names)
}

# Apply custom function to row names
new_row_names <- modify_row_names(rownames(mat))

# Assign modified row names to the matrix
rownames(mat) <- new_row_names


```

```{r fig.width=15, fig.height=15}
set.seed(1234)
ht<-ComplexHeatmap::pheatmap(mat, name="z_score",
                             # show_row_names = FALSE, # shows gene names
                             #show_row_dend=TRUE,
                            column_split = annotation_col$`Sustained\nClinical Response`, #splits col by group of interest
                             #cluster_cols=FALSE,
                             column_title="SoC significant BTM",
                             #column_title_side="top",
                             column_title_gp=gpar(fontsize=15, fontface="bold"),
                             column_title_rot=0,
                             #treeheight_col = 25,
                             #column_km=2,
                             row_km=3,
                             row_gap = unit(c(3, 3), "mm"),
                             #row_names_rot = 0, #rotate row names
                             #row_names_side = "left",
                             heatmap_legend_param = list(
                                  legend_direction = "horizontal", 
                                  legend_width = unit(5, "cm"),
                               title= "Z-score"),
                             border = TRUE,
                             border_color = "darkgrey",
                             cellheight = 15,
                             cellwidth = 15,
                             fontsize = 14,
                             show_colnames = FALSE,
                             col=col,
                             annotation_col = annotation_col,
                             annotation_colors = ann_colours,
                             annotation_names_col = F)

draw(ht,heatmap_legend_side ="bottom", annotation_legend_side = "left")






```

