---
title: "07_Cellular_deconvolution"
#output: html_notebook
---
# Asessing cellular composition of samples
Two ways - 
1. in silico deconvolution using CIBERSORTx (https://cibersortx.stanford.edu/)
2. using gold standard laboratory cell countts from ACCORD trial hematology data


# Load libraries

```{r message=FALSE}
# Packages to load
packages <- c(
 "flextable", "ggplot2", "ggpubr", "gtsummary", "RColorBrewer", "tidyverse")


############################################################################################################
load_libraries <- function(packages) {
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

# Load all packages
load_libraries(packages)

```
# Plotting aesthetics

```{r}
#set colour scheme for stacked plot

n <- 60
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
pie(rep(1,n), col=sample(col_vector, n))
col_graph<-as.data.frame(col_vector)
```

# Load data

```{r}
load("./data_files/cibersort_and_cell_counts.Rdata")
```


## CIBERSORTx

Deconvolution of the bulk RNASeq data was performed using CIBERSORTx online. Normalised
counts for each of the treatment arm extracted from the DE analysis were used as input to CIBERSORTx (https://cibersortx.stanford.edu/)
Job results were downloaded for processing and analysis in R. 
csort_combo contains these results for treatment arms combined into a single df, with additional metadata of interest

First look at Standard of Care arm between day 1 and day 5 and between responders

```{r message=FALSE}
soc_tbl <- csort_combo %>%
  filter(Treatment_Arm == "Standard of Care")%>% 
  select(Treatment_Arm, Day, Unique_ID, 2:23) %>% 
  tbl_strata(strata = Treatment_Arm, 
             ~.x %>%
               tbl_summary(by = Day, include = -Unique_ID, type = list(where(is.numeric) ~ "continuous")) %>% 
               modify_caption("Table 1. Comparison between Day 1 and Day 5 - Standard of Care") %>% 
               add_p(~ "paired.wilcox.test", group=Unique_ID) %>%
               bold_p(t = 0.05) %>%
               italicize_levels()) 
print(soc_tbl)

soc_tbl_response <- csort_combo %>% 
  filter(Treatment_Arm == "Standard of Care")%>% 
  select(Treatment_Arm, Sustained_Clinical_Response_by_D29, 2:23) %>% 
  tbl_strata(strata = Treatment_Arm, 
             ~.x %>%
               tbl_summary(by = Sustained_Clinical_Response_by_D29, type = list(where(is.numeric) ~ "continuous")) %>% 
               modify_caption("Table 2. Comparison between Responders - Standard of Care") %>% 
               add_p(~ "wilcox.test") %>%
               bold_p(t = 0.05) %>%
               italicize_levels())
print(soc_tbl_response)
```


Now Treatment Arms 

```{r message=FALSE}

# bemcentinib
bem_tbl <- csort_combo %>% 
  filter(Treatment_Arm == "Bemcentinib")%>% 
  select(Treatment_Arm, Day, Unique_ID, 2:23) %>% 
  tbl_strata(strata = Treatment_Arm, 
             ~.x %>%
               tbl_summary(by = Day, include = -Unique_ID, type = list(where(is.numeric) ~ "continuous")) %>% 
               modify_caption("Table 3. Comparison between Day 1 and Day 5 - Bemcentinib") %>% 
               add_p(~ "paired.wilcox.test", group=Unique_ID) %>%
               bold_p(t = 0.05) %>%
               italicize_levels()) 
print(bem_tbl)

# Tozorakimab
toz_tbl <- csort_combo %>% 
  filter(Treatment_Arm == "Tozorakimab")%>% 
  select(Treatment_Arm, Day, Unique_ID, 2:23) %>% 
  tbl_strata(strata = Treatment_Arm, 
             ~.x %>%
               tbl_summary(by = Day, include = -Unique_ID, type = list(where(is.numeric) ~ "continuous")) %>% 
               modify_caption("Table 4. Comparison between Day 1 and Day 5 - Tozorakimab") %>% 
               add_p(~ "paired.wilcox.test", group=Unique_ID) %>%
               bold_p(t = 0.05) %>%
               italicize_levels()) 
print(toz_tbl)

# Zilucoplan
zil_tbl <- csort_combo %>% 
    filter(Treatment_Arm == "Zilucoplan")%>% 
  select(Treatment_Arm, Day, Unique_ID, 2:23) %>% 
  tbl_strata(strata = Treatment_Arm, 
             ~.x %>%
               tbl_summary(by = Day, include = -Unique_ID, type = list(where(is.numeric) ~ "continuous")) %>% 
               modify_caption("Table 5. Comparison between Day 1 and Day 5 - Zilucoplan") %>% 
               add_p(~ "paired.wilcox.test", group=Unique_ID) %>%
               bold_p(t = 0.05) %>%
               italicize_levels()) 
print(zil_tbl)


```










Now plotting 
Using the cibersortx proportions to plot as stacked barcharts (props turned into percentages)

```{r fig.height=15, fig.width = 12}

# Split the data into separate dataframes for each Treatment_Arm and turn proportions into percentage for plotting
csort_longer <- csort_combo %>%
  pivot_longer(
    cols = 2:23,                    # Columns to pivot
    names_to = "Cell_Type",         # Name for the new key column
    values_to = "Prop"              # Name for the new value column
  ) %>%
  mutate(Percentages = Prop * 100)


standard_of_care_data <- csort_longer %>% filter(Treatment_Arm == "Standard of Care")
bemcentinib_data <- csort_longer %>% filter(Treatment_Arm == "Bemcentinib")
tozorakimab_data <- csort_longer %>% filter(Treatment_Arm == "Tozorakimab")
zilucoplan_data <- csort_longer %>% filter(Treatment_Arm == "Zilucoplan")


# Aggregate percentage for each group and normalise within the groups

# For Standard of Care
csort_median_soc <- standard_of_care_data %>%
  group_by(Treatment_Arm, Day, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_soc <- csort_median_soc %>%
  group_by(Treatment_Arm, Day) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# For Bemcentinib
csort_median_bemcentinib <- bemcentinib_data %>%
  group_by(Treatment_Arm, Day, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_bemcentinib <- csort_median_bemcentinib %>%
  group_by(Treatment_Arm, Day) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# For Tozorakimab
csort_median_tozorakimab <- tozorakimab_data %>%
  group_by(Treatment_Arm, Day, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_tozorakimab <- csort_median_tozorakimab %>%
  group_by(Treatment_Arm, Day) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# For Zilucoplan
csort_median_zilucoplan <- zilucoplan_data %>%
  group_by(Treatment_Arm, Day, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_zilucoplan <- csort_median_zilucoplan %>%
  group_by(Treatment_Arm, Day) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# Recombine the data into one dataset
csort_combined <- bind_rows(csort_normalized_soc, 
                            csort_normalized_bemcentinib,
                            csort_normalized_tozorakimab,
                            csort_normalized_zilucoplan)

# Now use this for plotting
# Standard of Care Day
standard_of_care_data_day <- csort_combined %>%
  filter(Treatment_Arm == "Standard of Care")


ggplot(standard_of_care_data_day, aes(x = Normalized_Percentage, y = Day, fill = Cell_Type)) +
  geom_bar(stat = "identity", position="stack") +
  labs(x = "Percentage", y = "Day", fill = "Cell Type") +
  theme_bw(base_size = 16) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x=element_text(size=16)) +
  theme(axis.ticks = element_blank()) +
  scale_fill_manual(values=col_vector[c(1:3, 9:60)]) +
  scale_x_continuous(expand = c(0,0), limits = c(0, 101)) +
  guides(fill = guide_legend(ncol = 1))+ 
  theme(legend.position = "none")

# All 3 drugs for Day
drug_data_day <- csort_combined %>%
  filter(Treatment_Arm != "Standard of Care")

ggplot(drug_data_day, aes(x = Day, y = Normalized_Percentage, fill = Cell_Type)) +
  geom_bar(stat = "identity", position="stack") +
  labs(x = "Day", y = "Percentage", fill = "Cell Type") +
  theme_bw(base_size = 20) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x=element_text(size=20)) +
  theme(axis.ticks = element_blank()) +
  scale_fill_manual(values=col_vector[c(1:3, 9:60)]) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 101)) +
  guides(fill = guide_legend(ncol = 1))+
  facet_wrap(~Treatment_Arm)


############

# Now do the same but for the Sustained Clinical Response variable

# Aggregate each group by Sustained_Clinical_Response_by_D29

# For Standard of Care
csort_median_soc_response <- standard_of_care_data %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_soc_response <- csort_median_soc_response %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# For Bemcentinib
csort_median_bemcentinib_response <- bemcentinib_data %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_bemcentinib_response <- csort_median_bemcentinib_response %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# For Tozorakimab
csort_median_tozorakimab_response <- tozorakimab_data %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_tozorakimab_response <- csort_median_tozorakimab_response %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# For Zilucoplan
csort_median_zilucoplan_response <- zilucoplan_data %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29, Cell_Type) %>%
  summarise(Median_Percentage = median(Percentages, na.rm = TRUE), .groups = "drop")

csort_normalized_zilucoplan_response <- csort_median_zilucoplan_response %>%
  group_by(Treatment_Arm, Sustained_Clinical_Response_by_D29) %>%
  mutate(Normalized_Percentage = Median_Percentage / sum(Median_Percentage) * 100)

# Recombine the  data into one dataset for Sustained Clinical Response
csort_combined_response <- bind_rows(csort_normalized_soc_response, 
                                     csort_normalized_bemcentinib_response,
                                     csort_normalized_tozorakimab_response,
                                     csort_normalized_zilucoplan_response)
# Now use this for plotting
# Standard of Care 
standard_of_care_data_response <- csort_combined_response %>%
  filter(Treatment_Arm == "Standard of Care")

ggplot(standard_of_care_data_response, aes(x = Normalized_Percentage, y = Sustained_Clinical_Response_by_D29, fill = Cell_Type)) +
  geom_bar(stat = "identity", position="stack") +
  labs(x = "Percentage", y = "Sustained Response", fill = "Cell Type") +
  theme_bw(base_size = 16) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x=element_text(size=16)) +
  theme(axis.ticks = element_blank()) +
  scale_fill_manual(values=col_vector[c(1:3, 9:60)]) +
  scale_x_continuous(expand = c(0,0), limits = c(0, 101)) +
  guides(fill = guide_legend(ncol = 1))+ 
  theme(legend.position = "none")

# All 3 drugs
drug_data_response<- csort_combined_response %>%
  filter(Treatment_Arm != "Standard of Care")

ggplot(drug_data_response, aes(x = Sustained_Clinical_Response_by_D29, y = Normalized_Percentage, fill = Cell_Type)) +
  geom_bar(stat = "identity", position="stack") +
  labs(x = "Sustained Response", y = "Percentage", fill = "Cell Type") +
  theme_bw(base_size = 20) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x=element_text(size=20)) +
  theme(axis.ticks = element_blank()) +
  scale_fill_manual(values=col_vector[c(1:3, 9:60)]) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 101)) +
  guides(fill = guide_legend(ncol = 1))+
  facet_wrap(~Treatment_Arm)


```



## Cell counts

Using hematology data generated from laboratory cell counting of ACCORD samples
Assessing cell counts for differences  according to responder status
These cell counts are included within the cibersort_and_cell_counts.Rdata file

```{r message=FALSE, fig.width=15, fig.height=6}

# Keep only distinct Unique_IDs for SoC only- removes duplicate rows for the same patient as there will be two d1 and d5 sample ids per unique_id

soc_counts <- cell_counts_meta %>%
  filter(Treatment_Arm == "Standard of Care") %>%
  distinct(Unique_ID, .keep_all = TRUE)  # Keeps the first occurrence of each Unique_ID

# pivot longer for plotting
soc_counts_percentage <- soc_counts %>%
  select(
    Unique_ID, 
    Day, 
    Treatment_Arm, 
    Sustained_Clinical_Response_by_D29, 
    starts_with(c("Screening", "Day_5_")) & ends_with("percentage")
  )%>%
  pivot_longer(
    cols = 5:ncol(.),
    names_to = "Cell_Type_Percentage",
    values_to = "Percentages"
  )

# Extract time and cell type
# Extract time
soc_counts_percentage$Time <- sub("^Screening.*", "Screening", soc_counts_percentage$Cell_Type_Percentage)
soc_counts_percentage$Time <- sub("^Day_(\\d+).*", "Day \\1", soc_counts_percentage$Time)
# Extract cell type
soc_counts_percentage$Cell_Type <- sub(".*_(\\w+)_percentage$", "\\1", soc_counts_percentage$Cell_Type_Percentage)
# Set Unique_ID as character
soc_counts_percentage$Unique_ID <- as.character(soc_counts_percentage$Unique_ID)
# Set levels for Time
soc_counts_percentage$Time <- factor(soc_counts_percentage$Time, levels = c("Screening", "Day 5"))


# Generate plots
 #set dodge for points
 pd <- position_dodge(0.2)
 
# Cell counts between Day 1 and Day 5 
 ggplot(soc_counts_percentage, aes(x = Time, y = Percentages, fill = Time)) +
   geom_boxplot(outlier.shape = NA, alpha=0.6) +
   # geom_line(aes(group = Unique_ID, colour="grey"), position = pd) +
   geom_point(aes(group=Unique_ID), size=3, shape=21, position = pd) +
   labs(x = "Timepoint", y = "Percentage of total cell count", fill = "Time point") +
   theme_bw(base_size = 14) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
   facet_wrap(~ Cell_Type, scales = "free", ncol = 5) +
   scale_fill_manual(values = alpha(c("darkblue", "skyblue"), 0.5)) +
   stat_compare_means(comparisons = list(c("Screening", "Day 5")),paired=TRUE, method="wilcox.test",vjust=0, tip.length = 0.05)+
   scale_y_continuous(expand=expansion(mult = .2))
 
#Cell counts between Responders (Sustained Clinical Response yes/no)
 ggplot(soc_counts_percentage, aes(x = Sustained_Clinical_Response_by_D29, y = Percentages, fill = Sustained_Clinical_Response_by_D29)) +
   geom_boxplot(outlier.shape = NA, alpha=0.6) +
   # geom_line(aes(group = Unique_ID, colour="grey"), position = pd) +
   geom_point(aes(group=Unique_ID), size=3, shape=21, position = pd) +
   labs(x = "Timepoint", y = "Percentage of total cell count", fill = "Time point") +
   theme_bw(base_size = 14) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
   facet_wrap(~ Cell_Type, scales = "free", ncol = 5) +
   scale_fill_manual(values = alpha(c("darkblue", "skyblue"), 0.5)) +
   stat_compare_means(comparisons = list(c("No", "Yes")),method="wilcox.test",vjust=0, tip.length = 0.05)+
   scale_y_continuous(expand=expansion(mult = .2))
 

```


